---
title: "State tax revenue volatility"
author: "Don Boyd"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_notebook: 
    df_print: paged
    fig_height: 6
    fig_width: 8
    toc: yes
    toc_float: yes
    number_sections: yes
  html_document:
    toc: yes
    toc_depth: '3'
    df_print: paged
editor_options:
  chunk_output_type: console
---

<!-- <style> -->
<!-- body .main-container { -->
<!--   max-width: 1500px; -->
<!-- } -->
<!-- </style> -->

<!-- Google drive -->
<!-- https://drive.google.com/drive/folders/1fM7HIP7qrKbIdndF_wQ30WxVb1j_Cq06 -->


```{r runall, eval=FALSE, echo=FALSE} 
# When we want a final report, run the following code selectively "by hand" (interactively)
# -- NEVER using Knit with eval=TRUE
# note <br> breaks line for html output but \n should break for pdf; also could try |

rmdfn <- "./Pew_Tax_Revenue_Volatility.rmd" # this file
outfn <- paste0("tax_volatility_", format(Sys.time(), "%Y-%m-%d"), ".html")
rmarkdown::render(rmdfn, output_format="html_document", output_file=outfn)

# library("RCurl")
# ftpUpload(outfn, "ftp://kffrig:boyd4812@files.000webhost.com/public_html/KFF_RIG_MedicaidCuts_new.html")

# Note may be safest to fully exit RStudio and restart it before running whole thing. Otherwise knitr may get confused
# and include repetitive information in the output html file.

```


# Initialization

This section:

+   sets options
+   loads libraries
+   defines functions
+   defines constants
+   defines themes
+   retrieves previously created data

It also contains some workflow notes as markdown comments.

It doesn't produce any output other than some informational messages.


<!--
Workflow notes:

# code folding:
#   alt-L, alt-shift-L  one section
#   alt-O, alt-shift-O  all sections
#   ctrl-D run current code section (remapped from Ctrl-alt-T)

# yaml notes:
always_allow_html: yes

  html_document:
    toc: yes
    toc_depth: '2'
    df_print: paged
    number_sections: yes

# renv workflow:
# to get it all started:
# renv::init()

# to remove it all:
# renv::deactivate()
# then delete the renv folder

# periodically execute:
# renv::snapshot()

# to see what directory is being used by renv for local package sources, execute:
# renv:::renv_paths_local()


# git workflow:
# when starting work from a different computer
# - pull project from remote
# git checkout -b issue-5  # create branch and checkout branch
# Stage local changes, commit, check status:
# git add foo.txt
# git commit --message "A commit message"
# git status

# now merge
# git checkout master
# git merge issue-5

# then push
# git push

-->

## Preliminaries

```{r setup, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(echo=FALSE, include=FALSE, eval=FALSE, message = FALSE, warning = FALSE)
```


```{r css, echo=FALSE}
# blockquote {
#     padding: 20px;
#     margin: 0 0 20px;
#     font-size: 14px;
#     border-left: 5px solid #eee;
# }

```



```{r get_libs_and_functions, eval=TRUE}
# clear objects from the workspace
rm(list=ls(all=TRUE))
source(here::here("r", "libs_base.r"))
source(here::here("r", "libs_ts.r"))
source(here::here("r", "functions.r"))
devtools::session_info()
# devtools::package_info()

```


```{r functions, eval=TRUE}

corr_tidy <- function(vars){
  # return a tidy data frame with correlations
  mat <- as.matrix((vars))
  Hmisc::rcorr(mat) %>%
    tidy
}


dl <- function(vec) {
  # change in log
  # ASSUME data are sorted by year and there are no gaps and no zeros
  lvec = log(vec)
  laglvec = c(NA, head(lvec, -1))
  lvec - laglvec
}


hpdf <- function(data, colname="value") {
  mod <- hpfilter(data[colname], freq=6.25, type=c("lambda"), drift=FALSE)
  tibble(trend=as.numeric(mod$trend),
         cycle=as.numeric(mod$cycle))
}


hptrend <- function(vec, smooth=NULL){
  vts_hp <- hpfilter(ts(vec), freq=smooth, type=c("lambda"), drift=FALSE)
  as.numeric(vts_hp$trend)
}


sre <- function(data) {
  # short-run elasticity
  mod = lm(dlvalue ~ dlgdp, data=data)
  list(mod=mod, coeff=unname(mod$coefficients["dlgdp"]))
}


sre_gsp <- function(data) {
  # short-run elasticity
  mod = lm(dlvalue ~ dlgdp + dlgsp, data=data)
  list(mod=mod, coeff=unname(mod$coefficients["dlgdp"]))
}

# slog <- function(vec){
#   # signed logarithm
#   sign(vec)*log(abs(vec))
# }


```


```{r constants, eval=TRUE}
# Constants, themes, etc.

theme_report <- theme_bw() + 
  theme(
    plot.caption = element_text(hjust = 0)
    )


nolegend <- theme(legend.position = "none")
nolegend_title <- theme(legend.title = element_blank())

bluered <- c("blue", "red")  # #a50f15 darkred
color2 <- c("blue", "red")


```


## Get previously created data
```{r get_data, include=FALSE, eval=TRUE}
load(file = here::here("data", "general.RData"), verbose=TRUE)
load(file = here::here("data", "econ_national.RData"), verbose=TRUE)
load(file = here::here("data", "econ_state.RData"), verbose=TRUE)
load(file = here::here("data", "taxdata.RData"), verbose=TRUE)

stcodes2 <- readRDS(here::here("data", "stcodes2.rds"))  # build this into bdata

# slim data files down if desired
pewtax  <- pewtax %>%
  select(stabbr, year, taxtype, vartype, pch, level)

```


# Materials for report, organized by topic


## Scene setting - exploration of patterns

### Year over year change in tax revenue and...

#### Nominal GDP

```{r eval=TRUE, include=TRUE}
df <- censustax %>%
  filter(stabbr=="US", name=="tottax") %>%
  select(year, tax=value) %>%
  inner_join(gdpfy, by="year") %>%
  pivot_longer(-year) %>%
  group_by(name) %>%
  mutate(pch=pchya(value, year)) %>%
  ungroup


recs <- recdata(1953)

p <- df %>%
  # filter(year >= 1970) %>%
  ggplot(aes(year, pch, colour=name)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0) +
  scale_x_continuous(name=NULL, breaks=seq(1950, 2030, 5)) +
  scale_y_continuous(name="% change",
                     breaks=seq(-20, 20, 2)) +
  scale_color_manual(values=bluered) +
  gband(recs$peak_decimal, recs$trough_decimal) +
  ggtitle("State Government Tax Revenue and Gross Domestic Product, United States") +
  theme_report +
  nolegend_title
p 
ggsave(filename = here::here("results", "gdp_tax_us.png"), plot = p, width=8, height=6)


```



### Inflation
```{r include=TRUE, eval=TRUE}

glimpse(censustax)
df <- censustax %>%
  filter(stabbr=="US", name=="tottax") %>%
  select(year, tax=value) %>%
  inner_join(gdppi %>% select(year, gdppi), by="year") %>%
  pivot_longer(-year) %>%
  group_by(name) %>%
  mutate(pch=pchya(value, year)) %>%
  ungroup
 
p <- df %>%
  filter(year >= 1955) %>%
  mutate(name=factor(name, levels=c("tax", "gdppi"), labels=c("taxes", "inflation"))) %>%
  ggplot(aes(year, pch, colour=name)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0) +
  scale_x_continuous(name=NULL, breaks=seq(1955, 2025, 5)) +
  scale_y_continuous(name="% change",
                     breaks=seq(-20, 20, 2)) +
  ggtitle("% change in tax revenue and general price level, United States") +
  scale_colour_manual(values=rev(bluered)) +
  theme_report +
  nolegend_title
p 
ggsave(filename = here::here("results", "gdppi_tax_us.png"), plot = p, width=8, height=6)
 

```




# Alternative volatility measures


```{r eval=TRUE}
# generate series with two kinds of volatility
nyears <- 30
base <- 100
grate <- .03
set.seed(12)
grvec <- (1 + grate)^(0:(nyears - 1))  # growth vector
noise <- rnorm(nyears, mean=0, sd=1)  # noise vector
shock_pattern <- c(0, 0, 0, 0, 0, .1)

pattern_smooth <- seq(.10, -.10, length.out=10)
# shock_smooth <- rep_len(pattern_smooth, nyears)
shock_smooth <- rep_len(c(pattern_smooth, -pattern_smooth), nyears) # smooth shock goes down then up

shock_smooth

# force our recession shock and random shocks to start at zero - drop last element of each vector
znoise = c(0, head(noise, -1))
zsmooth = c(0, head(shock_smooth, -1))

sum(znoise)
sum(zsmooth)
sum(abs(znoise))
sum(abs(zsmooth))

# force similar noise
ratio <- sum(abs(znoise)) / sum(abs(zsmooth))


df <- tibble(time=1:nyears, # + 2024,
             trend=base*grvec) %>%
  mutate(tnoise=trend * (1 + znoise / ratio),
         tsmooth=trend * (1 - zsmooth)) %>%
  mutate(tnoise = tnoise * sum(trend) / sum(tnoise),
         tsmooth = tsmooth * sum(trend) / sum(tsmooth),
         tnoise=ifelse(time==1, trend, tnoise),
         tsmooth=ifelse(time==1, trend, tsmooth))

# df <- tibble(time=1:nyears, # + 2024,
#              trend=base*grvec) %>%
#   mutate(tnoise=trend * (1 + znoise/25),
#          tsmooth=trend * (1 - zsmooth)* (1 + znoise/50))


dfl <- df %>%
  pivot_longer(-time) %>%
  group_by(name) %>%
  arrange(time) %>%
  mutate(pch=pchya(value, time),
         pch=ifelse(time==1, 0, pch)) %>%
  group_by(time) %>%
  mutate(dpch=(pch - pch[name=="trend"]) /100,
         dtrend=value - value[name=="trend"],
         pdtrend=dtrend / value[name=="trend"]) %>%
  ungroup
glimpse(dfl)

dfl %>%
  group_by(name) %>%
  summarise(value=sum(value))

dfl %>% filter(name=="tnoise")
dfl %>% filter(name=="tsmooth")
# our first panel is trend, random shock, and cyclical shock


```


## Kinds of volatility

States generally face two kinds of volatility: cyclical volatility, often following the business cycle, and random volatility.

In this section we examine three revenue series:

+   Simple trend that grows at 3% per year
+   Revenue source that has pure cyclical volatility
+   Revenue source with random volatility. 


All three revenue series raise the same amount of money over 30 years.

For each of the following graphs, ask:

+   Which form of volatility would require a **larger rainy day fund** if a state wanted to protect against long downturns?
+   Which form of volatility would make revenue **harder to forecast** and could be a concern if the state is focused on avoiding unanticipated budget shortfalls?


```{r eval=TRUE}

fdtrend <- function(type){
  dfl %>%
    filter(name==type) %>%
    mutate(posneg=ifelse(pdtrend < 0, "neg", "pos")) %>%
    ggplot(aes(time, pdtrend)) +
    # geom_line(colour="black") +
    geom_hline(yintercept = 0) +
    geom_bar(stat="identity", aes(fill = posneg)) + 
    scale_fill_manual(values=rev(bluered)) +
    scale_x_continuous(name="year", breaks=seq(0, 100, 5)) +
    scale_y_continuous(name="% difference from trend",
                       breaks=seq(-.2, .2, .05), 
                       labels=scales::percent_format(accuracy=1),
                       limits=c(-.1, .1)) +
    theme_report +
    nolegend
}


fpdiff <- function(type){
  dfl %>%
    filter(name==type) %>%
    mutate(posneg=ifelse(dpch < 0, "neg", "pos")) %>%
    ggplot(aes(time, dpch)) +
    # geom_line(colour="black") +
    geom_hline(yintercept = 0) +
    geom_bar(stat = "identity", aes(fill = posneg)) + 
    scale_fill_manual(values=rev(bluered)) +
    scale_x_continuous(name="year", breaks=seq(0, 100, 5)) +
    scale_y_continuous(name="shock % growth minus trend % growth",
                       breaks=seq(-.2, .2, .05), 
                       labels=scales::percent_format(accuracy=1),
                       limits=c(-.1, .1)) +
    theme_report +
    nolegend
}


# ppanels <- grid.arrange(p_cycle_vstrend, p_cycle_vsgrowth, 
#              p_random_vstrend, p_random_vsgrowth, 
#              nrow = 2, top="Two kinds of volatility and two kinds of volatility measures")
# ppanels
# grid::grid.draw(ppanels)


p_cycle_vstrend <- fdtrend("tsmooth") + ggtitle("Shock: cyclical\nMeasure: % deviation from trend revenue")
p_random_vstrend <- fdtrend("tnoise") + ggtitle("Shock: random\nMeasure: % deviation from trend revenue")

p_cycle_vsgrowth <- fpdiff("tsmooth") + ggtitle("Shock: cyclical\nMeasure: shock % growth minus trend % growth")
p_random_vsgrowth <- fpdiff("tnoise") + ggtitle("Shock: random\nMeasure: shock % growth minus trend % growth")


```


## Comparing **level** (amount) of actual revenue to trend revenue

We begin by comparing the level (amount) of actual revenue to trend revenue. 


```{r eval=TRUE, include=TRUE}
# plot series and percent change
colors = c("black", "blue", "darkgreen")
p1 <- dfl %>%
  mutate(name=factor(name, 
                     levels=c("trend", "tsmooth", "tnoise"),
                     labels=c("trend", "cyclical", "random"))) %>%
  ggplot(aes(time, value, colour=name)) +
  geom_line() +
  geom_hline(yintercept = 100) +
  scale_y_continuous(name="revenue", limits=c(90, 260), labels=scales::dollar) +
  scale_x_continuous(name="year", breaks=seq(0, 40, 5), limits=c(0, 32)) +
  scale_colour_manual(values=colors) +
  ggtitle("Two kinds of revenue volatility: pure cyclical and pure random shocks",
          subtitle="All 3 revenue series raise the same amount over 30 years") +
  theme_report +
  nolegend_title
p1
ggsave(filename = here::here("results", "trend_cycle_random_level.png"), plot = p1, width=8, height=6)

# dev.cur()
# dev.off()

```


The two-panel graph below shows the *level* of actual revenue minus trend revenue for each source -- that is, it measure deviation from trend. 

+   Which of the two kinds of volatility would subject the state to long periods of shortfalls and might suggest the need for a larger rainy day fund?
+   Which form of volatility has greater year-to-year variation and might be more difficult to forecast? How easily can we tell that if we compare to the trend?

In the graph, cyclical volatility subjects the state to long consecutive periods of underperformance and overperformance, suggesting the need for a larger rainy day fund. The deviation-from-trend measure makes this clear.

It's not clear from the graph which form of volatility subjects the state to greater year-to-year uncertainty: the deviation-from-trend measure does not tell us much about this.


```{r include=TRUE, eval=TRUE, fig.width=6, fig.height=9}

p_rdf <- grid.arrange(p_cycle_vstrend,
                      p_random_vstrend, 
             nrow = 2, top="Two kinds of volatility, % difference from trend revenue")
grid::grid.draw(p_rdf)

ggsave(filename = here::here("results", "dev_trend.png"), plot = p_rdf, width=8, height=12)


```


## Comparing revenue **growth rates** relative to trend growth rates 

The graph below shows the year-to-year growth rates of these revenue sources.

The next two-panel graph compares *growth rates* of actual revenue to growth rates for trend revenue, instead of comparing the *level* of actual revenue to trend revenue. It measures the deviation in the growth rate from the trend growth rate.


```{r eval=TRUE, include=TRUE}
# percent change
colors = c("black", "blue", "darkgreen")
p2 <- dfl %>%
  mutate(name=factor(name, 
                     levels=c("trend", "tsmooth", "tnoise"),
                     labels=c("trend", "cyclical", "random"))) %>%
  ggplot(aes(time, pch / 100, colour=name)) +
  geom_line() +
  geom_hline(yintercept = 0) +
  scale_y_continuous(name="% change", labels=scales::percent) +
  scale_x_continuous(name="year", breaks=seq(0, 40, 5), limits=c(0, 32)) +
  scale_colour_manual(values=colors) +
  ggtitle("Two kinds of revenue volatility: pure cyclical and pure random shocks",
          subtitle="All 3 revenue series raise the same amount over 30 years") +
  theme_report +
  nolegend_title
p2
ggsave(filename = here::here("results", "trend_cycle_random_pch.png"), plot = p2, width=8, height=6)

```

The two-panel graph below shows actual revenue growth minus trend revenue growth for each source. 

+   Which revenue source has greater year-to-year variation and might be more difficult to forecast?
+   Which revenue source would subject the state to long periods of shortfalls and might suggest a larger rainy day fund? How easily can we tell that when we are comparing actual growth rates to trend growth rates?

In the graph, cyclical volatility subjects the state to long consecutive periods of underperformance and overperformance, suggesting the need for a larger rainy day fund. The deviation-from-trend measure makes this clear.

It's not clear from the graph which form of cyclicality subjects the state to greater year-to-year uncertainty: the deviation-from-trend measure does not tell us much about this.


```{r include=TRUE, eval=TRUE, fig.width=6, fig.height=9}

p_forecast <- grid.arrange(p_cycle_vsgrowth,
                      p_random_vsgrowth, 
             nrow = 2, top="Two kinds of volatility, difference from growth rate in trend revenue")
grid::grid.draw(p_forecast)

ggsave(filename = here::here("results", "dev_pch.png"), plot = p_forecast, width=8, height=12)


```

## Formal measures of volatility


```{r vformal, eval=TRUE}
# prepare the data
# https://cran.r-project.org/web/packages/broom/vignettes/broom_and_dplyr.html
# http://r4stats.com/2017/04/18/group-by-modeling-in-r-made-easy/

censustax %>% 
  filter((stabbr == "OH" & name == "cit")) %>%
  mutate(lvalue=log(value), slvalue=slog(value),
         dl=dl(value), dsl=dsl(value))

vbase <- censustax %>%
  filter(!(stabbr == "DE" & name == "gst")) %>% # only 3 obs, 1951-53
  filter(!(stabbr == "OH" & name == "cit")) %>% # negative in 2014
  rename(taxtype=name, tax=value) %>%
  left_join(gdpfy, by="year") %>%
  left_join(sgdpfy %>%
              filter(name=="gdp") %>%
              rename(sgdp=value) %>%
              select(year, stabbr, sgdp),
            by=c("stabbr", "year")) %>%
  arrange(stabbr, taxtype, year) %>% 
  group_by(stabbr, taxtype) %>%
  mutate(taxpch=tax / tax[match(year - 1, year)] * 100 - 100,
         dltax=dl(tax),
         dlgdp=dl(gdp),
         dlsgdp=dl(sgdp),
         trend=hptrend(tax, smooth=6.25))%>%
  ungroup

summary(vbase)
vbase %>% filter(!is.na(dlgdp)) %>% summary() # 1952-2019
vbase %>% filter(!is.na(dlsgdp)) %>% summary() # 1965-2019


vol <- vbase %>%
  filter(year >= 1990) %>%
  group_by(stabbr, taxtype) %>%
  nest() %>%
  mutate(pchsd = map_dbl(data, function(df) sd(df$taxpch, na.rm=TRUE)),
         hpsd = map_dbl(data, function(df) sd(df$tax / df$trend * 100 - 100, na.rm=TRUE)),
         sremod = map(data, safely(function(df) lm(dltax ~ dlgdp, data=df))),
         sregspmod = map(data, safely(function(df) lm(dltax ~ dlsgdp, data=df)))) %>%
  ungroup

# unpack the summary measures from sremod
sredf <- vol %>%
  select(stabbr, taxtype, sremod) %>%
  unnest_wider(col=sremod) %>%  # we need the column "result"
  mutate(tidied=map(result, tidy)) %>%
  unnest(cols=tidied) %>%
  filter(term=="dlgdp") %>%
  select(stabbr, taxtype, sre=estimate, srese=std.error)
  
sregspdf <- vol %>%
  select(stabbr, taxtype, sregspmod) %>%
  unnest_wider(col=c(sregspmod)) %>%  # we need the column "result"
  mutate(tidied=map(result, tidy)) %>%
  unnest(cols=tidied) %>%
  filter(term=="dlsgdp") %>%
  select(stabbr, taxtype, sregsp=estimate, sregspse=std.error)
  
tmp <- vol %>%
  select(stabbr, taxtype, sregspmod) %>%
  filter(row_number()==1)

str(tmp$sregspmod)

measures <- vol %>%
  select(stabbr, taxtype, pchsd, hpsd) %>%
  left_join(sredf, by = c("stabbr", "taxtype")) %>%
  left_join(sregspdf, by = c("stabbr", "taxtype"))

# get all relevant correlations
corr_types <- measures %>%
  select(-stabbr) %>%
  group_by(taxtype) %>%
  nest() %>%
  summarise(map_df(data, corr_tidy), .groups="drop")  %>%
  unite("pair", column1, column2, remove=FALSE)
corr_types

corr_types %>%
  select(taxtype, pair, estimate) %>%
  pivot_wider(names_from = pair, values_from = estimate)

corr_states <- measures %>%
  select(-taxtype) %>%
  group_by(stabbr) %>%
  nest() %>%
  # only include columns we want in the correlation
  summarise(map_df(data, corr_tidy), .groups="drop") %>%
  unite("pair", column1, column2, remove=FALSE)
# corr_states


```

## How correlated are the different measures, by tax?

## Correlation of the main measures
```{r eval=TRUE, include=TRUE}

# hpsd_pchsd	sre_pchsd	sre_hpsd	srese_pchsd	srese_hpsd	srese_sre
vorder <- c("tottax", "iit", "gst", "cit", 
            "selsalestax", "sevtax", "othertax")
vlabs = c("Total taxes", 
          "Individual income tax", "General sales tax", "Corporate income tax", 
          "Selected sales taxes", "Severance taxes", "Other taxes")
morder <- c("hpsd_pchsd", "sre_pchsd", "sre_hpsd",
            "srese_sre", "srese_hpsd", "srese_pchsd")

tabdata <- corr_types %>%
  filter(pair %in% morder) %>% 
  select(taxtype, pair, estimate) %>%
  mutate(taxtype=factor(taxtype, levels=vorder, labels=vlabs),
         pair=factor(pair, levels=morder)) %>%
  arrange(taxtype, pair) %>%
  pivot_wider(names_from = pair, values_from = estimate)
shade_rows <- seq(2, nrow(tabdata), 2)
names(tabdata)

tabdata %>%
  gt() %>%
    tab_header(
      title = "Correlation of selected volatility measures by tax type",
      subtitle = "Estimated over 1990-2018"
      ) %>%
      cols_label(taxtype="Type of tax",
                 hpsd_pchsd=html("Percent change<br>-- with --<br>Deviation from trend"),
                 sre_pchsd=html("Percent change<br>-- with --<br>Short-run elasticity (SRE)"),
                 sre_hpsd=html("Deviation from trend<br>-- with --<br>Short-run elasticity (SRE)"),
                 srese_sre=html("Short-run elasticity (SRE)<br>-- with --<br>Standard error of SRE"),
                 srese_pchsd=html("Percent change<br>-- with --<br>Standard error of SRE"),
                 srese_hpsd=html("Deviation from trend<br>-- with --<br>Standard error of SRE")) %>%
      cols_align(
        align = c("left"),
        columns = taxtype) %>%
      fmt_number(
        columns = contains("_"),
        decimals = 3) %>%
    tab_style(
      style = list(
        cell_fill(color = "lightgrey")
      ),
      locations = cells_body(
        rows=shade_rows
      )
    )
	
# data2 %>%
#     select(-stabbr) %>%
#     group_by(stname) %>%
#     gt() %>%
#     tab_header(
#       title = ttitle,
#       subtitle = "New York and the nation"
#       ) %>%
#     cols_label(stname="",
#                varname="",
#                pch="% change") %>%
#     fmt_currency(
#       columns = c("2017", "2018", "change"),
#       decimals = 0,
#       accounting = TRUE,
#       suffixing = FALSE
#     ) %>%
#       fmt_percent(
#       columns = c("pch"),
#       decimals = 1
#     ) %>%
#     fmt_missing(
#       columns = c("2017", "2018", "change", "pch"),
#       missing_text = "n.m."
#     ) %>%
#     tab_style(
#       style = list(
#         cell_fill(color = "#f7f7f7")
#       ),
#       locations = cells_body(
#         rows=shade_rows
#       )
#     ) %>%
#     cols_align(align = c("left"), columns = c("stname", "varname")) %>%
#     tab_options(row_group.background.color = "lightgreen") %>%
#     tab_footnote(
#     footnote = "For NY minus US rows, % change is NY % change minus US % change",
#     locations = cells_column_labels( 
#       columns = vars(pch)
#       )
#     ) %>%
#     tab_source_note(
#       source_note = note
#       )
```

## Correlation of volatility measures that control for national or state GDP
```{r}
count(corr_types, pair)
# hpsd_pchsd	sre_pchsd	sre_hpsd	srese_pchsd	srese_hpsd	srese_sre
vorder <- c("tottax", "iit", "gst", "cit", 
            "selsalestax", "sevtax", "othertax")
vlabs = c("Total taxes", 
          "Individual income tax", "General sales tax", "Corporate income tax", 
          "Selected sales taxes", "Severance taxes", "Other taxes")
morder <- c("sregsp_sre", "srese_sre", "sregspse_sregsp", "sregspse_srese")


tabdata <- corr_types %>%
  unite("pair", column1, column2) %>%
  filter(pair %in% morder) %>% 
  select(taxtype, pair, estimate) %>%
  mutate(taxtype=factor(taxtype, levels=vorder, labels=vlabs),
         pair=factor(pair, levels=morder)) %>%
  arrange(taxtype, pair) %>%
  pivot_wider(names_from = pair, values_from = estimate)
shade_rows <- seq(2, nrow(tabdata), 2)
names(tabdata)

# sregsp_sre", "srese_sre", "sregspse_sregsp", "sregspse_sre
tabdata %>%
  gt() %>%
    tab_header(
      title = "Correlation of selected volatility measures by tax type",
      subtitle = "Estimated over 1990-2018"
      ) %>%
      cols_label(taxtype="Type of tax",
                 sregsp_sre=html("Short-run elasticity (SRE)<br>-- with --<br>State-specific GDP SRE"),
                 srese_sre=html("Short-run elasticity (SRE)<br>-- with --<br>Its standard error"),
                 sregspse_sregsp=html("State-specific GDP SRE<br>-- with --<br>Its standard error"),
                 sregspse_srese=html("Short-run elasticity (SRE)<br>-- with --<br>Standard error of state-specific GDP SRE")) %>%
      cols_align(
        align = c("left"),
        columns = taxtype) %>%
      fmt_number(
        columns = contains("_"),
        decimals = 3) %>%
    tab_style(
      style = list(
        cell_fill(color = "lightgrey")
      ),
      locations = cells_body(
        rows=shade_rows
      )
    )

```


## Scatterplots of measures
```{r}
# pchsd hpsd sre sregsp
# get index by type of tax, for each state
count(measures, stabbr)
scatdata <- measures %>%
  filter(stabbr != "US") %>%
  select(-c(srese, sregspse)) %>%
  group_by(taxtype) %>%
  mutate(across(-c(stabbr), ~ .x / median(.x, n.rm=TRUE) * 100))
scatdata  

scatdata %>%
  filter(taxtype=="tottax", stabbr != "AK") %>%
  ggplot(aes(x=pchsd, y=hpsd, label=stabbr)) +
   geom_point(colour="blue", size=0.5) +
   geom_text(colour="blue", size=2) +
   geom_abline(slope=1)

scatdata %>%
  filter(taxtype=="tottax", stabbr != "AK") %>%
  ggplot(aes(x=sre, y=hpsd, label=stabbr)) +
   geom_point(colour="blue", size=0.5) +
   geom_text(colour="blue", size=2) +
   geom_abline(slope=1)

scatdata %>%
  filter(taxtype=="tottax", stabbr != "AK") %>%
  ggplot(aes(x=sre, y=sregsp, label=stabbr)) +
   geom_point(colour="blue", size=0.5) +
   geom_text(colour="blue", size=2) +
   geom_abline(slope=1)

scatdata %>%
  filter(taxtype=="tottax", stabbr != "AK") %>%
  ggplot(aes(x=sregsp, y=hpsd, label=stabbr)) +
   geom_point(colour="blue", size=0.5) +
   geom_text(colour="blue", size=2) +
   geom_abline(slope=1)
  

 # vol %>%
 #   select(stabbr, taxtype, vartype, data) %>%
 #   unnest(data) %>%
 #   group_by(stabbr, taxtype, vartype) %>%
 #   summarise(pchsd=sd(pch, na.rm=TRUE), .groups="drop") %>%
 #   pivot_wider(names_from = vartype, values_from = pchsd) %>%
 #   filter(taxtype %in% txs) %>%
 #   ggplot(aes(x=raw, y=adjusted, label=stabbr)) +
 #   geom_point(colour="blue", size=0.5) +
 #   geom_text(colour="blue", size=2) +
 #   geom_abline(slope=1) +
 #        nolegend_title +
 #   facet_wrap(~taxtype, ncol=2, scales = "free") +
 #   theme_report


```


## State maps of volatility
```{r eval=TRUE, include=TRUE}
us_states <- map_data("state")

count(us_states, region)  # 49, including DC!

head(us_states)
p <- ggplot(data = us_states,
            mapping = aes(x = long, y = lat,
                          group = group, fill = region))

p + geom_polygon(color = "gray90", size = 0.1) +
    coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
    guides(fill = FALSE)



```





## volatility by tax
```{r eval=TRUE, include=TRUE}
vranks <- vol %>%
  filter(!(stabbr=="AK" & taxtype=="iit")) %>%
  group_by(taxtype) %>%
  select(stabbr, taxtype, pchsd, hpsd) %>%
  mutate(rpchsd = rank(desc(pchsd)),
         rhpsd = rank(desc(hpsd)))

vranks %>%
  filter(taxtype=="tottax") %>%
  arrange(rhpsd)

# ttorder <- c("tottax", "iit", "gst", "cit", "selsalestax", "sevtax", "othertax")
# vranks %>%
#   select(stabbr, taxtype, rhpsd) %>%
#   mutate(taxtype=factor(taxtype, levels=ttorder)) %>%
#   arrange(taxtype) %>%
#   pivot_wider(names_from = taxtype, values_from = rhpsd) %>%
#   arrange(sevtax)

```

# Volatility over time
```{r include=TRUE, eval=TRUE}
# now we want rolling versions of the volatility measures
vperiod <- 15

# vol_roll <- vbase  %>%
#   filter(stabbr=="NY") %>%
#   filter(year >= 1990) %>%
#   group_by(stabbr, taxtype) %>%
#   # nest() %>%
#   rowwise %>%
#   mutate(pchsd = sd(taxpch, na.rm=TRUE))
# vol_roll

vbase %>%
  filter(stabbr=="MO", taxtype=="sevtax", year >= 1970)

vbase %>%
  filter(stabbr=="AK", taxtype=="cit", year >= 1970)

vol_roll <- vbase  %>%
  # filter(stabbr=="NY") %>%
  filter(!(stabbr=="MO" & taxtype=="sevtax")) %>%
  filter(year >= 1970) %>%
  group_by(stabbr, taxtype) %>%
  mutate(hpp=tax / trend * 100 - 100,
         pchsd = rollsd_p(taxpch, vperiod),
         hpsd = rollsd_p(hpp, vperiod),
         sre=rollsre(dltax, dlgdp, vperiod))
vol_roll
tail(vol_roll)

vol_roll %>%
  filter(stabbr=="NY", taxtype=="tottax") %>%
  ggplot(aes(year, sre)) +
  geom_line()


vol_roll %>%
  filter(stabbr=="CT", taxtype=="tottax") %>%
  ggplot(aes(year, sre)) +
  geom_line()

st <- "MA"
tax <- "tottax"
vol_roll %>%
  select(stabbr, taxtype, year, hpsd, sre) %>%
  filter(year >= 1985) %>%
  filter(stabbr==st, taxtype==tax) %>%
  pivot_longer(cols=c(hpsd, sre)) %>%
  group_by(stabbr, taxtype, name) %>%
  mutate(ivalue=value / value[year==2000] * 100) %>%
  ggplot(aes(year, ivalue, colour=name)) +
  geom_line() +
  geom_point()




rollsre <- function(y, x, nobs) {
  # print("in rollsre")
  sremod <- function(df){
    # print("in sremod")
    mod <- lm(y ~ x, na.action=na.exclude, data=df)
    sre <- coef(summary(mod))[2, "Estimate"]
    sre
  }
  safe_sremod <- safely(sremod)
  # print("before df")
  df <- data.frame(y=y, x=x)
  # print("after df")
  zoo::rollapply(df, nobs, function(df) sremod(as.data.frame(df)),
                 by.column = FALSE,
                 fill=NA, align="right")
}

rr2 <- rollapply(seat2, width = 36,
                FUN = function(z) coef(lm(y ~ y1 + y12, data = as.data.frame(z))),
                by.column = FALSE, align = "right")

rollsd_p <- function(x, nobs) {
  # population standard deviation
  sdp <- function(x){
    n <- sum(!is.na(x))
    sd(x, na.rm=TRUE) * sqrt((n - 1) / n)
  }
  zoo::rollapply(x, nobs, function(x) sdp(x),
                 fill=NA, align="right")
}

vol <- vbase %>%
  filter(year >= 1990) %>%
  group_by(stabbr, taxtype) %>%
  nest() %>%
  mutate(pchsd = map_dbl(data, function(df) sd(df$taxpch, na.rm=TRUE)),
         hpsd = map_dbl(data, function(df) sd(df$tax / df$trend * 100 - 100, na.rm=TRUE)),
         sremod = map(data, safely(function(df) lm(dltax ~ dlgdp, data=df))),
         sregspmod = map(data, safely(function(df) lm(dltax ~ dlsgdp, data=df)))) %>%
  ungroup


```




# Impact of policy changes

It is better, conceptually, to adjust for policy changes. But how much difference does it make?

```{r}
glimpse(pewtax)
count(pewtax, year) # 1995-2014
count(pewtax, taxtype) # 14, including cit, gst, iit, mft, tottax
count(pewtax, vartype) # adjusted, policy, raw
# value is percent change, nominal
summary(pewtax)




```


## Exploratory look at the impact of adjusting for policy changes

### Percent change in adjusted and unadjusted revenue, selected states
```{r eval=TRUE, include=TRUE, fig.width=8, fig.height=8}
# data(package="bdata")
#    beargn  beargn.name          n
#    <chr>   <chr>            <int>
#  1 ""      ""                   2
#  2 "fwr"   "Far West"           6
#  3 "glr"   "Great Lakes"        5
#  4 "mdatl" "Mid Atlantic"       6
#  5 "neng"  "New England"        6
#  6 "plr"   "Plains"             7
#  7 "rmr"   "Rocky Mountain"     5
#  8 "ser"   "Southeast"         12
#  9 "swr"   "Southwest"          4
# 10 "usr"   "United States"      1
# find the state with the largest tax revenue in each region, by type of tax

states_ranked <- censustax %>%
  filter(year==2019, stabbr %in% state.abb) %>%
  left_join(stcodes %>% select(stabbr, beargn, beargn.name)) %>%
  group_by(name, beargn) %>%
  arrange(desc(value)) %>%
  mutate(rank=row_number()) %>%
  ungroup

f <- function(tax, taxlabel){
  p <- pewtax %>%
        filter(taxtype==tax) %>%
        select(stabbr, vartype, year, pch) %>%
        left_join(states_ranked %>%
        filter(name==tax) %>%
        select(stabbr, beargn.name, rank), by="stabbr") %>%
        group_by(beargn.name) %>%
        filter(rank==min(rank)) %>%
        ungroup %>%
        mutate(vartype=factor(vartype, 
                              levels=c("raw", "adjusted", "policy"),
                              labels=c("actual", "adjusted", "policy change")),
               stname=stname(stabbr),
               stregion=paste0(beargn.name, ": ", stname)) %>%
        arrange(beargn.name, vartype) %>%
        ggplot(aes(year, pch, colour=vartype)) +
        geom_point() +
        geom_line() +
        scale_colour_manual(values=colors) +
        scale_y_continuous(name="% change", breaks=seq(-1, 1, .05),
                           labels=scales::percent_format(accuracy = 1), limits=c(NA, NA)) +
        scale_x_continuous(name=NULL, breaks=seq(1950, 2030, 5)) +
        geom_hline(aes(yintercept=0)) +
        facet_wrap(~stregion, ncol=2, scales="free_y") +
        ggtitle(paste0(taxlabel, " revenue: Percent change in actual versus policy-adjusted revenue"),
                subtitle="Largest state in each region with data for this revenue source") +
        labs(caption="Source: Data provided by Pew Charitable Trusts") +
        theme_report +
        nolegend_title
    
        ggsave(filename = here::here("results", paste0("policy_", tax, "_selstates.png")), 
               plot = p, 
               width=8, height=8)
        p
}

colors <- c("blue", "red", "grey")


f("tottax", "Total tax")
f("iit", "Individual income tax")
f("gst", "General sales tax")
f("cit", "Corporate income tax")
# f("mft", "Motor fuel tax")
  

```


## Specific volatility measures, with and without policy changes
```{r}
# construct levels from the pewtax data
# pewtax

pewtax %>% filter(stabbr=="NY", taxtype=="tottax", vartype=="raw")


vbase <- pewtax %>%
  filter(vartype != "policy") %>%
  left_join(sgdpfy %>%
              filter(name=="gdp") %>%
              rename(sgdp=value) %>%
              select(year, stabbr, sgdp),
            by=c("stabbr", "year")) %>%
  arrange(stabbr, taxtype, vartype, year) %>% 
  group_by(stabbr, taxtype, vartype) %>%
  mutate(pch=pch * 100,
         dlvalue=dl(level),
         dlsgdp=dl(sgdp),
         trend=hptrend(level, smooth=6.25)) %>%
  ungroup


# construct volatility measures for each state, raw and adjusted
vol <- pewtax %>%
  filter(vartype != "policy") %>%
  left_join(sgdpfy %>%
              filter(name=="gdp") %>%
              rename(gdp=value) %>%
              select(year, stabbr, gdp),
            by=c("stabbr", "year")) %>%
  arrange(stabbr, taxtype, vartype, year) %>% 
  group_by(stabbr, taxtype, vartype) %>%
  mutate(pch=pch * 100,
         dlvalue=dl(level),
         dlgdp=dl(gdp)) %>%
  ungroup %>%
  group_nest(stabbr, taxtype, vartype) %>%
  mutate(hpf=map2(data, "level", safely(hpdf)),
         sre=map(data, safely(sre)))
glimpse(vol)


# analyze several measures
vol2 <- vol %>%
  hoist(data,
        year="year",
        level="level",
        pch="pch") %>%
  select(-data) %>%
  hoist(hpf, trend=c("result", "trend")) %>%
  select(-hpf) %>%
  unnest(cols=c(year, level, pch, trend))


txs <- c("tottax", "iit", "gst", "mft", "cit", "proptax")
txs <- c("tottax", "iit", "gst", "cit")

 vol %>%
   select(stabbr, taxtype, vartype, data) %>%
   unnest(data) %>%
   group_by(stabbr, taxtype, vartype) %>%
   summarise(pchsd=sd(pch, na.rm=TRUE), .groups="drop") %>%
   pivot_wider(names_from = vartype, values_from = pchsd) %>%
   filter(taxtype %in% txs) %>%
   ggplot(aes(x=raw, y=adjusted, label=stabbr)) +
   geom_point(colour="blue", size=0.5) +
   geom_text(colour="blue", size=2) +
   geom_abline(slope=1) +
        nolegend_title +
   facet_wrap(~taxtype, ncol=2, scales = "free") +
   theme_report
 

glimpse(vol2) 
vmeasures <-  vol2 %>%
  mutate(pdtrend=level / trend * 100 - 100) %>%
  group_by(stabbr, taxtype, vartype) %>%
  summarise(pchsd=sd(pch, na.rm=TRUE),
            pdtrendsd=sd(pdtrend, na.rm=TRUE),
            sre=first(sre),
            .groups="drop") %>%
  hoist(sre, sre_coeff=c("result", "coeff")) %>%
  select(-sre)%>%
  pivot_longer(-c(stabbr, taxtype, vartype), names_to = "measure") %>%
  group_by(taxtype, vartype, measure) %>%
  arrange(value) %>%
  mutate(rank=row_number())


txs <- c("tottax", "iit", "gst", "cit")
vranks <- vmeasures %>%
  filter(taxtype %in% txs) %>%
  select(-value) %>%
  pivot_wider(names_from = vartype,
              values_from = rank)
count(vranks, measure)
  

vranks %>%
  group_by(taxtype, measure) %>%
  summarise(cval=cor(raw, adjusted,
                     use="pairwise.complete.obs", method="spearman")) %>%
  pivot_wider(names_from = measure, values_from = cval)



vranks %>%
  filter(taxtype=="tottax", measure=="pdtrendsd") %>%
  ungroup %>%
  select(raw, adjusted) %>%
  correlate()



Orange %>% 
  group_by(Tree) %>%
  summarize(correlation = cor(age, circumference))
 
# library(corrr)
# d %>% 
#   correlate() %>% 
#   focus(mpg:drat, mirror = TRUE) %>% 
#   network_plot()
 
#  
# df3 <-  df2 %>%
#   # select(-sre) %>%
#   unnest_wider(hpf) %>%
#   unnest_wider(result) %>%
#   unnest(c(data, trend)) %>%
#   select(stabbr, taxtype, vartype, year, value, pch, trend, sre) %>%
#   mutate(pdtrend=value / trend * 100 - 100) 

df4 <- df3 %>%
  pivot_wider(names_from = vartype, values_from = pdtrend) %>%
  mutate(change = adjusted - raw)

changes <- df3 %>%
  group_by(stabbr, taxtype, vartype) %>%
  summarise(pchsd=sd(pch, na.rm=TRUE),
            pdtrendsd=sd(pdtrend, na.rm=TRUE), .groups="drop")

changes %>%
  select(-pchsd) %>%
  pivot_wider(names_from = vartype, values_from = pdtrendsd) %>%
  mutate(change=adjusted - raw) %>%
  arrange(-abs(change))

 
df3 %>%
   group_by(taxtype) %>%
   summarise(n=n(),
             raw=median(raw, na.rm=TRUE),
             adjusted=median(adjusted, na.rm=TRUE),
             change=median(change, na.rm=TRUE)) %>%
   filter(n > 9) %>%
   arrange(desc(change))

df2 %>%
  filter(stabbr=="NY", taxtype=="tottax") %>%
  arrange(year)

df3 %>%
  filter(stabbr=="NY", taxtype=="tottax") %>%
  arrange(year)
  
 
 
 changes <- df2 %>%
   select(stabbr, taxtype, vartype, data) %>%
   unnest(data) %>%
   group_by(stabbr, taxtype, vartype) %>%
   summarise(pchsd=sd(pch, na.rm=TRUE), .groups="drop") %>%
   pivot_wider(names_from = vartype, values_from = pchsd) %>%
   mutate(change=adjusted - raw)
 
 changes %>%
   group_by(taxtype) %>%
   summarise(n=n(),
             raw=median(raw, na.rm=TRUE),
             adjusted=median(adjusted, na.rm=TRUE),
             change=median(change, na.rm=TRUE)) %>%
   filter(n > 9) %>%
   arrange(desc(change))
 
 
 #  hoist(data, coeff=c("result", "coeff"))

sredf <- df2 %>%
  hoist(sre, coeff=c("result", "coeff")) %>% # reach within result to get coeff
  select(stabbr, taxtype, vartype, coeff) %>%
  group_by(taxtype, vartype) %>%
  arrange(desc(coeff)) %>%
  mutate(srank=row_number()) %>%
  ungroup

count(sredf, taxtype)

txs <- c("tottax", "iit", "gst", "mft", "cit", "proptax")
txs <- c("tottax", "iit", "gst", "cit")

sredf %>%
  select(-srank) %>%
  pivot_wider(names_from = vartype, values_from = coeff) %>%
  # filter(taxtype=="tottax") %>%
  filter(taxtype %in% txs) %>%
  ggplot(aes(x=raw, y=adjusted, label=stabbr)) +
  geom_point(colour="blue", size=0.5) +
  geom_text(colour="blue", size=2) +
  geom_abline(slope=1) +
        nolegend_title +
  facet_wrap(~taxtype, ncol=2, scales = "free") +
  theme_report

txs <- c("tottax", "iit", "gst", "mft", "cit", "proptax")
txs <- c("tottax", "iit", "gst", "cit")
sredf %>%
  select(-coeff) %>%
  pivot_wider(names_from = vartype, values_from = srank) %>%
  filter(taxtype %in% txs) %>%
  ggplot(aes(x=raw, y=adjusted, label=stabbr)) +
  geom_point(colour="blue", size=0.5) +
  geom_text(colour="blue", size=2) +
  geom_abline(slope=1) +
        nolegend_title +
  scale_y_reverse() +
  scale_x_reverse() +
  facet_wrap(~taxtype, ncol=2, scales = "free") +
  theme_report
    
# ggsave(filename = here::here("results", paste0("policy_", tax, "_scatter.png")), 
#        plot = p, 
#        width=8, height=8)

sredf %>%
  pivot_wider(names_from = vartype, values_from = coeff)

# which state had the largest change in volatility??
txs <- c("tottax", "iit", "gst", "mft", "cit", "proptax", "sevtax")
txs <- c("tottax", "iit", "gst", "cit", "sevtax")
changes %>%
  filter(taxtype %in% txs) %>%
  arrange(desc(abs(change))) %>%
  filter(row_number() < 15)


```



# Capital gains
```{r eval=TRUE}
# % change in gdp, tax revenue, cg
cgtax <- gdpfy %>%
  full_join(capgains,
            by="year") %>%
  full_join(censustax %>%
              filter(name %in% c("tottax", "iit", "gst")) %>%
              rename(tax=value),
            by="year") %>%
  drop_na() %>%
  group_by(stabbr, name) %>%
  arrange(year) %>%
  mutate(gdp_pch=pchya(gdp, year),
         cg_pch=pchya(capgains, year),
         tax_pch=pchya(tax, year),
         cggdp=capgains / gdp * 100) %>%
  ungroup
  
```


```{r gdpcgtax_plot}
cgtax %>%
  filter(stabbr=="US") %>%
  select(year, gdp_pch, cg_pch, tax_pch) %>%
  pivot_longer(-year) %>%
  ggplot(aes(year, value, colour=name)) +
    geom_point() +
    geom_line() +
    geom_hline(yintercept = 0) +
  ggtitle("Percent change in GDP, capital gains, and tax revenue for the United States",
          subtitle = "Not adjusted for inflation")
  
```


```{r cggdp_plot, eval=TRUE, include=TRUE, fig.height=6, fig.width=6}
# fix to use calendar year??
rec <- recessions %>%
  filter(rec_year > 1953)

anno <- cgtax %>%
  filter(stabbr=="US") %>%
  mutate(cggdp=cggdp / 100)

p <- cgtax %>%
  filter(stabbr=="US") %>%
  select(year, cggdp) %>%
  ggplot(aes(year, cggdp / 100)) +
  geom_point(colour="blue") +
  geom_line(colour="blue") +
  scale_y_continuous(name="% of GDP", labels=scales::percent_format(accuracy = .1), limits=c(0, NA)) +
  scale_x_continuous(name=NULL, breaks=seq(1950, 2030, 5)) +
  gband(rec$peak_decimal, rec$trough_decimal) +
  geom_hline(aes(yintercept=median(cggdp, na.rm=TRUE) / 100)) +
  annotate("text", x=1955, y=median(anno$cggdp, na.rm=TRUE) +.001, label = "median", size=2.5) +
  annotate("text", x=1987, y=.0765, label = "Federal\ntax reform", size=2.5) +
  annotate("text", x=2000.25, y=.0675, label = "Dot-com\nbubble", size=2.5) +
  annotate("text", x=2007.25, y=.0685, label = "Real estate\nbubble", size=2.5) +
  ggtitle("Capital gains as a percent of GDP",
          subtitle = "Recession periods are shaded") +
  labs(caption="Horizontal line marks median for the period shown") +
  theme_report
p
ggsave(filename = here::here("results", "cggdp_us.png"), plot = p, width=8, height=6)


# theme(
#   plot.title = element_text(),
#   plot.subtitle.title = element_text(),
#   plot.caption = element_text()
# )

# p + theme(
#   plot.caption = element_text(hjust = 0)
#   )

  
```





# Construct Seegert volatility measure
```{r}
df <- censustax %>%
  filter(stabbr=="US", name=="tottax", year >= 1959) %>%
  select(year, tax=value) %>%
  left_join(gdpfy, by="year") %>% # nominal gdp
  arrange(year)

df2 <- df %>%
  pivot_longer(-year) %>%
  group_by(name) %>%
  arrange(year) %>%
  mutate(trend=hptrend(value, smooth=100),
         vol1=log(abs(value - trend)),
         vol2=(value - trend) / trend * 100)

df2 %>%
  filter(year >= 1970) %>%
  ggplot(aes(year, vol2, colour=name)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0) +
  scale_x_continuous(name=NULL) +
  scale_y_continuous(name="volatility (% difference from smoothed trend)",
                     breaks=seq(-20, 20, 2)) +
  ggtitle("Volatility of State Government Tax Revenue and Gross Domestic Product, United States",
          subtitle="Volatility measured as % difference from smoothed trend (HP filter)") +
  theme_bw()

# Colorado-type approach
df2 <- df %>%
  mutate(return=tax / lag(tax),
         dreturn=log(return)*100)

df2 %>%
  filter(year >= 1970) %>%
  ggplot(aes(year, vol)) +
  geom_line() +
  geom_point()

```


## Introduction and motivation

### Setting the scene - year over year % change
```{r pchplot, fig.height=4.5, fig.width=8, eval=TRUE, include=TRUE}
# sd10c
#  pchplot, fig.height=4.5, fig.width=8, eval=FALSE}
# summary(sgtax.a)
irecs <- which(recessions$rec_year >= 1960)

p2 <- taxgdpus %>%
  filter(year >= 1960) %>%
  ggplot(aes(year, vpch, colour=name)) +
  geom_line(size=1) +
  geom_point(size=1.15) +
  scale_y_continuous(name="% change", breaks=seq(-20, 20, 2)) +
  scale_x_continuous(name=NULL, breaks=seq(1960, 2030, 5)) +
  scale_colour_manual(values=c("red", "blue")) +
  geom_hline(yintercept = 0) +
  annotate("rect",
           xmin = recessions$peak[irecs] %>% decimal_date(),
           xmax = recessions$trough[irecs] %>% decimal_date(),
           fill = "lightgrey",
           alpha = .5, # larger alpha is darker rectangle
           ymin = -Inf, ymax = Inf) +
  ggtitle("Year over year growth in state government tax revenue and GDP for the United States",
          subtitle="Recession periods are shaded.") +
  labs(caption="Source: U.S. Bureau of the Census and Bureau of Economic Analysis") +
  theme_bw() +
  theme(plot.caption = element_text(hjust=0, size=9)) +
  theme(legend.title = element_blank())
p2
ggsave(filename = here::here("results", "tax_gdp_us.png"), plot = p2)


```


## Ways to measure tax revenue volatility

* Past research
  + Rolling variance (Dauchy 2013)
  + Deviation from trend
    + Kwak 2013 says White 1983; Dye and McGuire 1991; Braun and Otsuka 1998; Aisen and Jose´ Veiga 2008 all used deviation from trend
    + Orthogonal Distance Regression (Kwak 2013)
     

### Selected measures for the United States




#### Rolling standard deviation


#### Deviation from trend


## Recent trends in revenue volatility


## Drivers of state revenue volatility

### The nature of national economic changes

### State economic structure

### State tax structure

#### The portfolio of state taxes

#### The structure of individual taxes


##	Implications for states in current recession and beyond

### Special characteristics of the Covid-19 recession and state tax volatility

### How much can states reduce volatility with alternative tax structures?




# APPENDICES BELOW HERE


# Project management and organization
## Deadlines and schedule

* July 3, 2020 Project start
* October 3, 2020 Provide list of external reviewers
* June 3, 2021  Draft report due (plan on sooner)
* August 3, 2021 Project end and final report due


## Links

+  [Zotero folder](https://www.zotero.org/groups/2541602/pew_tax_revenue_volatility/library)
+  [Google drive folder](https://drive.google.com/drive/folders/1fM7HIP7qrKbIdndF_wQ30WxVb1j_Cq06)
+  [Google sheet](https://docs.google.com/spreadsheets/d/1hSWy_7ep4RbM9lfSA-LYnDeGI1c6p9Ud0GqcSlR9xJ8/edit?usp=sharing) (same as link below)
+   Draft report: [Google docs](https://docs.google.com/document/d/13MaAnySHpE8ODvvRjraMPtR7l6GFCX_hk7qDP2KC_s8/edit?usp=sharing) is current version


## Goals and issues

* Ways to measure tax revenue volatility

* Recent trends in revenue volatility
  + Has tax revenue become more volatile over time?
  + If so, why?

* Drivers of state revenue volatility - to what extent driven by
  + The nature of national economic changes (e.g., characteristics of recession)?
  + State economic structure?
  + State tax structure? 
    + The portfolio of taxes? 
    + The structure of individual taxes?

*	Implications for states in current recession and beyond
  +  What does the Covid-19 recession imply for state tax volatility?
  +  How much can states reduce volatility with alternative tax structures?
    

Note: For details on references see [this](https://docs.google.com/spreadsheets/d/1hSWy_7ep4RbM9lfSA-LYnDeGI1c6p9Ud0GqcSlR9xJ8/edit?usp=sharing).


## Notes to reviewers

Use renv.



## Data tasks and status
### Done

* State tax revenue from Census by tax, state, and year, 1951-2019, collapsed to the following categories (currently):
  + total taxes (tottax)
  + individual income tax (iit)
  + general sales tax (gst)
  + selective sales taxes such as tobacco, motor fuel, motor vehicle fees, alcohol… (selsalestax)
  + corporate income tax (cit)
  + severance taxes (sevtax)
  + all other taxes (othertax)
* Pew data on policy changes  
* Economic data
  + gross domestic product, by state fiscal year, 1951-2019 (still must adjust for states with fiscal years different from July 1)
  + gross domestic price index (chained), by state fiscal year, 1951-2019 (same adjustment required)
  + official NBER dates for recessions during relevant time periods
  + obtained NIPA details for description of recessions
  + obtained state-level consumption data from BEA

### TODO

* organize NIPA data to measure recessions
* organize NIPA data for consumption details
* organize state consumption data
* Possibly get data from Dauchy
* Graph pch tax vs econ
* IRS data components of income
* Measures of income tax progressivity
* Similar measures sales tax?
* Consider base = revenue / rate as robustness check

## Details about the project

### Overall goals

Examine tax revenue volatility and its impacts on state governments

Provide deep and broad understanding of:

* Ways to measure tax revenue volatility
* Recent trends in revenue volatility
* Drivers of state revenue volatility
*	Implications for states in current recession and beyond

Ultimate goal: Insights that can help policymakers manage budgets


### Important sub-questions

* To what extent is volatility driven by:
    + The nature of national economic changes (e.g., characteristics of recession)?
    + State economic structure?
    + State tax structure? The portfolio of taxes? The structure of individual taxes?
* Has tax revenue become more volatile over time? If so, why?
* How much can states reduce volatility with alternative tax structures?

All have implications for states in current recession and beyond.


### Overall plan

1. Review relevant literature
2. Build database for analyzing volatility
    + Annual state tax revenue, all states, by tax
    + Auxiliary data to allow population & inflation adjustment
    + Selected other data
3. Multiple methods
    + Descriptive analytic review of history
    + Decomposition
    + Simulations of portfolio effects
    + Exploratory econometrics for certain topics


### Core data source

* Census Bureau Annual Survey of State Finances
  + State tax revenue, annual, all states, many taxes
  + Comprehensive (not just general fund)
  + Relatively consistent definitions over time and states
  + Annual period consistent with policymaking cycle
  + Includes multiple recessions (some data back to mid-1950s):
      + 1980-81, 1990, 2001, and 2007 (2020 won't be available)
      + may be able to include 1960, 1969, 1973 recessions

### Selected technical & data issues

* Tax data implicitly include embedded policy changes. Use Pew tax-change data to help answer: How much does adjusting for policy changes affect conclusions?
* How much do alternative volatility measures affect conclusions?
* Data on structure of individual taxes are limited. To extent practical, use these limited data to help answer: How does structure of individual taxes affect volatility?
  

### Topics and approaches
+----------------------------------+----------------------------------------------+
| Topic                            | Methods                                      |
+==================================+==============================================+
| Ways to measure                  | - Literature review                          |
| revenue volatility               | - Conceptual differences among measures      |
|                                  | - Trends with different measures             |
+----------------------------------+----------------------------------------------+
| Recent revenue volatility trends | - Descriptive analysis                       |
+----------------------------------+----------------------------------------------+
| Drivers of revenue volatility    | - Descriptive analysis                       |
|   (Nature of economic change,    | - Decomposition                              |
|    State economic structure,     | - Econometric analysis                       |
|    State tax structure)          | - Simulation of alternative structures       |
+----------------------------------+----------------------------------------------+
| Implications in current          | - Synthesis of above                         |
| recession and beyond             |                                              |
+----------------------------------+----------------------------------------------+


### Comments from the August 4, 2020 methods review

* Clarify key terms with thorough definitions and describe the applied metrics, highlighting areas of standardization. Absent those details, some confusion may arise. Specifically,
    +	Convey the practical differences between the proposed metrics of revenue volatility for state and other policymakers. These metrics can vary considerably depending on the underlying data distribution, meaningfully changing what is observed as volatility. 
    + Elaborate on how revenue streams may need to be combined/collapsed for analytical tractability. If doing so, highlight the thresholds for those modification, if these differ by state or source, and if those decisions affect the results.  
    + Account for the possible interpretations of concepts like “state economic structures” and economies “at risk”, and how those choices shift the meaning of the analysis and findings. If these concepts are applied differently by state, explain how these may be standardized.  

* Ensure that the selected time frames (periods and trends) and their decision rules are transparent in the final report. Specifically: 
    + Discuss how partial recessionary periods (i.e. quarters) will be addressed since the study will generally leverage annual data, to speak to policymaker budgeting cycles/priorities.
    +	Clarify that “recent trends” may reflect multi-decade periods of fiscal management, which may not be self-evident. Explain how the bounds of “trend periods” are chosen, whether through an assessment of business cycles, regression analyses, inductive analyses or other techniques.

* Ensure consistency between this and prior Pew studies, or provide a crosswalk, particularly if the methods have been modified or augmented. Focus on what sources of uncertainty remain embedded in each measure of volatility. 
    + Assess whether correcting for tax policy adjustments changes estimated volatility. Also, since the policy adjusted time series is shorter, estimate whether a shorter series will complicate the development of robust volatility estimates using the more technical/complex approaches. 
    + Highlight — for more technical methods — how trends are being estimated, and how differences in those trends might change the interpretation of volatility, as the embedded sources of uncertainty differ (e.g. by extracting cyclicality). 







```{r tax_data, eval=TRUE}

# tax_stack <- tax2 %>%
#   rename(nominal=value) %>%
#   left_join(gdppi %>% select(year, igdppi), by="year") %>%
#   mutate(real=nominal * igdppi) %>%
#   select(-igdppi) %>%
#   pivot_longer(c(nominal, real), names_to="type")

```


```{r}
# data checking
# what to do about negative tax values?? they will become nan in logs, and we need to be robust to that
# what about zero values in the middle that look like they must be missing
# OH cit 2014 is negative (-$118)
# NM cit has pos 1951-1954, zero 1955-1966, then positive

tax_stack %>%
  filter(value <= 0) %>%
  as.data.frame()

tax_stack %>%
  filter(stabbr=="OH", name=="cit", type=="nominal")

tax_stack %>%
  filter(stabbr=="NM", name=="cit", type=="nominal")

measures %>%
  filter(stabbr=="OH", name=="cit", type=="nominal")
```


### Basic tax measures
```{r}


measures <- tax_stack %>%
  group_by(stabbr, name, type) %>%
  mutate(lvalue=log(value))

#   nest() %>%

```


```{r vol_measures}
# hp_df <- function(df, vname){
#   vts <- ts(df[vname])
#   vts_hp <- hpfilter(vts, freq=NULL, type=c("lambda"), drift=FALSE)
#   df %>% mutate(trend=as.numeric(vts_hp$trend),
#                 cycle=as.numeric(vts_hp$cycle))
# }


# takes a LONG LONG time
# vol1 <- tax_real %>%
#   filter(name=="tottax") %>%
#   group_by(stabbr, name, type) %>%
#   arrange(year) %>%
#   hp_df(., "value")
#   



# df <- df %>%
#   f(., "iit") %>%
#   mutate(dtrend=cycle / trend * 100,
#          pch=iit / iit[match(year - 1, year)] * 100 - 100,
#          dtrend_sd10=rollsd_p(dtrend, 10),
#          pch_sd10=rollsd_p(pch, 10))


```



```{r explore}
# remove trend
df <- sgtax.a %>%
  filter(year >= 1951, stabbr=="NY", item=="T40") %>%
  select(year, stabbr, iit=value)
df %>% ht  

# iit <- ts(df$iit, start = 1951, end=2019)
# iit <- ts(df$iit)
# plot(iit)
# 
# stl(iit, s.window = "periodic")
# # hpfilter(iit, freq=NULL, type=c("lambda","frequency"),drift=FALSE)
# iit_hp <- hpfilter(iit, freq=NULL, type=c("lambda"), drift=FALSE)
# iit_hp$trend
# iit_hp$cycle

f <- function(df, vname){
  vts <- ts(df[vname])
  vts_hp <- hpfilter(vts, freq=NULL, type=c("lambda"), drift=FALSE)
  df %>% mutate(trend=as.numeric(vts_hp$trend),
                cycle=as.numeric(vts_hp$cycle))
}

df <- df %>%
  f(., "iit") %>%
  mutate(dtrend=cycle / trend * 100,
         pch=iit / iit[match(year - 1, year)] * 100 - 100,
         dtrend_sd10=rollsd_p(dtrend, 10),
         pch_sd10=rollsd_p(pch, 10))

glimpse(df)


df %>%
  select(year, iit, trend) %>%
  pivot_longer(cols=c(iit, trend)) %>%
  ggplot(aes(year, value, colour=name)) +
  geom_line() +
  geom_point()


df %>%
  ggplot(aes(year, dtrend)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0)
  
df %>%
  ggplot(aes(year, pch)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0)

# df %>%
#   select(year, dtrend, pch) %>%
#   pivot_longer(cols=-year) %>%
#   ggplot(aes(year, value, colour=name)) +
#   geom_line() +
#   geom_point() +
#   geom_hline(yintercept = 0)
# 
# df %>%
#   select(year, dtrend, pch) %>%
#   mutate(pch=pch - mean(pch, na.rm=TRUE)) %>%
#   pivot_longer(cols=-year) %>%
#   ggplot(aes(year, value, colour=name)) +
#   geom_line() +
#   geom_point() +
#   geom_hline(yintercept = 0)

# another change

df %>%
  select(year, dtrend_sd10, pch_sd10) %>%
  pivot_longer(cols=-year) %>%
  ggplot(aes(year, value, colour=name)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0)
 
```


## Calculate volatility measures

```{r vfunctions}
hp_df <- function(df, vname, smooth=NULL){
  vts <- ts(df[vname])
  vts_hp <- hpfilter(vts, freq=smooth, type=c("lambda"), drift=FALSE)
  df %>% 
    mutate(trend=as.numeric(vts_hp$trend),
           cycle=as.numeric(vts_hp$cycle))
}

```


```{r}
tax_stacked
count(tax_stacked, name)
tax_stacked %>% 
  filter(type=="nominal", name=="tottax") %>%
  group_by(stabbr) %>%
  summarise(firstyear=min(year),
            lastyear=max(year))  # 1951-2019

# differences from actual revenue
# prep the tax file -- determine first year of each series!!! and start there
taxprep <- tax_stacked %>%
  arrange(stabbr, name, type, year) %>%
  group_by(stabbr, name, type) %>%
  mutate(ratio=value / value[match(year - 1,  year)],
         logpch=log(ratio) * 100,
         pch=value / value[match(year - 1,  year)] * 100 - 100)
taxprep
# note logpch warning in  stabbr = "OH", name = "cit",

# separate files with the different measures
# hpfilter all items

hp2 <- function(df, vname, smooth=NULL){
  vts <- ts(df[vname])
  vts_hp <- hpfilter(vts, freq=smooth, type=c("lambda"), drift=FALSE)
  tibble(trend=as.numeric(vts_hp$trend),
         cycle=as.numeric(vts_hp$cycle))
}

a <- proc.time()
hpall <- tax_stacked %>%
  # filter(stabbr=="OH") %>%
  group_by(stabbr, name, type) %>%
  nest()  %>%
  mutate(hp=map(data, hp2, "value", smooth=100))
b <- proc.time() 
b - a

glimpse(hpall)

hpall %>%
  filter(stabbr=="OH", type=="nominal", name=="cit") %>%
  unnest(cols = c(data, hp))


```




# Appendix

## Colorado revenue volatility review
```{r covol}


# lambda
freq <- 1
lam <- (2 * sin(pi / freq))^(-4)  # lambda w/freq=1: 2.779035e+62


conom <- slgfin %>%
  filter(stabbr=="CO", level==2, aggvar=="osr", year %in% 1992:2017) %>%
  arrange(year)

cohp <- conom %>%
  mutate(return=value / lag(value),
         dreturn=log(return) * 100) %>%
  filter(year %in% 1993:2017) %>%
  hp_df(., "dreturn", smooth=1600) %>%
  mutate(vol=cycle)
sum(cohp$cycle)

cohp_dtrend <- conom %>%
  filter(year %in% 1993:2017) %>%
  hp_df(., "value", smooth=1600) %>%
  mutate(vol=cycle / trend * 100)

comp <- bind_rows(cohp %>% select(year, vol) %>% mutate(name="dpch"),
                  cohp_dtrend %>% select(year, vol) %>% mutate(name="dtrend"))

comp %>% 
  pivot_wider(values_from = vol)
  

cohp %>%
  mutate(vol=cycle) %>%
  ggplot(aes(year, vol)) +
  geom_bar(stat="identity", fill="blue") +
  scale_x_continuous(breaks=1993:2017) +
  scale_y_continuous(breaks=seq(-20, 20, 2.5)) +
  theme(axis.text.x = element_text(angle = 45, vjust=0, hjust=0))

cohp %>%
  select(year, dreturn, trend) %>%
  pivot_longer(cols=-year) %>%
  ggplot(aes(year, value, colour=name)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 0)

comp %>%
  ggplot(aes(year, vol, colour=name)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 0)

# djb new start ----
param <- 1600

cohp <- conom %>%
  mutate(return=value / lag(value),
         dreturn=log(return) * 100) %>%
  filter(year %in% 1993:2017) %>%
  hp_df(., "dreturn", smooth=param) %>%
  mutate(vol=cycle)

cohp_dtrend <- conom %>%
  filter(year %in% 1993:2017) %>%
  hp_df(., "value", smooth=param) %>%
  mutate(vol=cycle / trend * 100)

comp <- bind_rows(cohp %>% select(year, vol) %>% mutate(name="dpch"),
                  cohp_dtrend %>% select(year, vol) %>% mutate(name="dtrend"))

p <- comp %>%
  filter(year %in% 1993:2017) %>%
  mutate(name=factor(name,
                     levels=c("dpch", "dtrend"),
                     labels=c("Calculated as footnote describes",
                              "Calculated as graph is labeled"))) %>%
  ggplot(aes(year, vol)) +
  geom_bar(stat="identity", fill="blue") +
  scale_x_continuous(name=NULL, breaks=1993:2017) +
  scale_y_continuous(name="Percent", breaks=seq(-20, 20, 5), limits=c(-15, 10)) +
  facet_wrap(~name, ncol=1) +
  ggtitle("Colorado's revenue deviation from trend calculated two ways",
          subtitle=paste0("HP filter smoothing parameter: ", param)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=0.5))
p



cotaxhp <- slgfin %>%
  filter(stabbr=="CO", level==2, aggvar=="tottax", year %in% 1992:2017) %>%
  arrange(year) %>%
  mutate(return=value / lag(value),
         dreturn=log(return) * 100) %>%
  filter(year %in% 1993:2017) %>%
  hp_df(., "dreturn", smooth=1600)

cohp %>%
  select(year, cycle) %>%
  mutate(name="osr") %>%
  bind_rows(cotaxhp %>% 
              select(year, cycle) %>%
              mutate(name="tax")) %>%
  ggplot(aes(year, cycle, colour=name)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 0)

co_vol <- tax_stacked %>%
  filter(name=="tottax", type=="real", stabbr=="CO", year %in% 1993:2017) %>%
  group_by(stabbr, name, type) %>%
  arrange(year) %>%
  hp_df(., "value")


co_vol %>%
  mutate(vol= cycle / trend * 100) %>%
  ggplot(aes(year, vol)) +
  geom_bar(stat="identity")


co_vol2 <- tax_stacked %>%
  filter(name=="tottax", type=="nominal", stabbr=="CO", year %in% 1992:2017) %>%
  group_by(stabbr, name, type) %>%
  arrange(year) %>%
  mutate(return=value / lag(value),
         dreturn=log(return) * 100) %>%
  filter(year %in% 1993:2017) %>%
  hp_df(., "dreturn")

co_vol2

co_vol2 %>%
  mutate(vol= cycle / trend * 100) %>%
  ggplot(aes(year, vol)) +
  geom_bar(stat="identity")

tax_stacked %>%
  filter(name=="tottax", type=="nominal", stabbr=="CO", year %in% 1992:2017) %>%
  group_by(stabbr, name, type) %>%
  arrange(year) %>%
  mutate(pch=value / lag(value) * 100 - 100) %>%
  filter(year %in% 1993:2017) %>%
  mutate(diff=pch - median(pch))


data(package="bdata")

tmp <- slgfin %>%
  filter(stabbr=="CO", level==2, aggvar=="osr", year %in% 1992:2017)

cov3 <- tmp %>%
  arrange(year) %>%
  mutate(return=value / lag(value),
         dreturn=log(return) * 100) %>%
  filter(year %in% 1993:2017) %>%
  hp_df(., "dreturn") %>%
  filter(year %in% 1993:2017)


```



# Earlier notes



```{r pchplot_old, fig.height=4.5, fig.width=8, eval=FALSE}
# summary(sgtax.a)
irecs <- which(recessions$rec_year >= 1960)

p1 <- sgt1 %>%
  filter(year >= 1960) %>%
  ggplot(aes(year, vpch)) +
  geom_line(colour="blue", size=1) +
  geom_point(colour="blue", size=1.15) +
  scale_y_continuous(name="% change", breaks=seq(-20, 20, 2)) +
  scale_x_continuous(name=NULL, breaks=seq(1960, 2030, 5)) +
  geom_hline(yintercept = 0) +
  annotate("rect",
           xmin = recessions$peak[irecs] %>% decimal_date(),
           xmax = recessions$trough[irecs] %>% decimal_date(),
           fill = "lightgrey",
           alpha = .5, # larger alpha is darker rectangle
           ymin = -Inf, ymax = Inf) +
  ggtitle("Year over year growth in state government tax revenue for the United States",
          subtitle="Recession periods are shaded.") +
  labs(caption="Source: U.S. Bureau of the Census") +
  theme_bw() +
  theme(plot.caption = element_text(hjust=0, size=9))
p1

```



```{r vplot, eval=FALSE}
# summary(sgtax.a)

irecs <- which(recessions$rec_year >= 1960)

p <- sgt1 %>%
  filter(year >= 1960) %>%
  ggplot(aes(year, sd10)) +
  geom_line(colour="blue") +
  geom_point(colour="blue", size=0.75) +
  scale_y_continuous(name="Standard deviation (%)", breaks=seq(0, 10, .5), limits=c(0, NA)) +
  scale_x_continuous(name=NULL, breaks=seq(1960, 2030, 5)) +
  annotate("rect",
           xmin = recessions$peak[irecs] %>% decimal_date(),
           xmax = recessions$trough[irecs] %>% decimal_date(),
           fill = "lightgrey",
           alpha = .5, # larger alpha is darker rectangle
           ymin = -Inf, ymax = Inf) +
  ggtitle("Volatility in aggregate state government tax revenue for the United States",
          subtitle="Trailing 10-year standard deviation of annual % change. Recession periods are shaded.") +
  theme_bw()
p    


```


# Nest play area
```{r}

# how correlated are measures
# summarize(correlation = cor(age, circumference))
# pchsd  hpsd       sre sre_stderr
# measures %>%
#   group_by(taxtype) %>%
#   summarise(pch_hp=cor(pchsd, hpsd, use="pairwise.complete.obs"),
#             pch_sre=cor(pchsd, sre, use="pairwise.complete.obs"),
#             pch_srese=cor(pchsd, sre_stderr, use="pairwise.complete.obs"),
#             hp_sre=cor(hpsd, sre, use="pairwise.complete.obs"),
#             hp_srese=cor(hpsd, sre_stderr, use="pairwise.complete.obs"),
#             sre_srese=cor(sre, sre_stderr, use="pairwise.complete.obs"))


# vol2 <- vol %>%
#   mutate(temp = map(sremod, function(mod) unname(mod$result$coefficients["dlgdp"])))
# 
# 
# vol2 <- vol %>%
#   unnest_auto(sremod)
# 
# vol3 <- vol2 %>%
#   tidy(result)
# 
# names(vol2$result[[1]])
#   
#   
#   mutate(mod2 = map(sreunlist(sremod))
# 
# 
# coef %>% 
#                          as.list %>%
#                          as_tibble)
# 
# vol2 %>%vol2 %>%tidy()
#   as_tibble() %>%
#   tidy %>% dplyr::select(term, estimate) %>% spread(term, estimate)))  %>% dplyr::select(id, model1)
# 
# 
# vol2 %>%
#   mutate(coeff=as.double(temp[[1]]))
# 
# d1 <- vol %>%
#   unnest_wider(sremod)
# names(d1$result[[1]])
# 
# d1 %>%
#   hoist(result, "coefficients")
#   unnest_auto(result)
#   
#   hoist(sremod, "coefficients")
# 
# df2 %>%
#   
#   hoist(sre, coeff=c("result", "coeff"))  # reach within result to get coeff
# 
# tmp <- df2 %>%
#   filter(stabbr=="NY", name=="iit")
# str(tmp$sre)
# str(tmp$sre[[1]])
# names(tmp$sre[[1]]$result)
# 
# coef(tmp$sre[[1]])
# 
# tmp$sre[[1]]
# names(tmp$sre[[1]])
# names(tmp$sre[[1]]$result)
# names(tmp$sre[[1]]$result$coeff)
# 
# tmp %>%
#   select(stabbr, name, sre) %>%
#   unnest(cols = c(sre))
# 
# tmp %>%
#   hoist(sre, coeff=list("result", 1))
# 
# tmp %>%
#   hoist(sre, coeff=c("result", "coeff"))
# 
# 
# tmp %>%
#   hoist(sre, "result", "mod", "coefficients", "dlgdp")
# 
# tmp %>%
#   unnest_wider(sre) %>%
#   unnest_wider(result)
# 
# 
# 
# tmp %>%
#   select(stabbr, name, hpf) %>%
#   unnest(cols = c(hpf))
# 
# 
# 
# tmp %>%
#   ungroup %>%
#   hoist(sre,
#         mod="mod")
# 
# 
# tmp %>%
#   ungroup %>%
#   hoist(sre,
#         # mod="mod",
#         coeff="coeff")
# 
# 
# names(tmp2$sre[[1]])
# tmp2 %>%
#   ungroup %>%
#   hoist(sre,
#         # mod="mod",
#         coeff="coeff")
# 
# 
# tmp <- df2 %>%
#   filter(stabbr=="NY", name=="iit", year >= 1965) %>%
#   nest()
# 
# tmp2 <- tmp %>%
#   mutate(sre=map(data, sredf))
# tmp2$sre
# 
# mod <- lm(dlvalue ~ dlgdp, data=tmp$data[[1]])
# str(mod)
# str(mod$coefficients["dlgdp"])
# mod$coefficients["dlgdp"]
# unname(mod$coefficients["dlgdp"])
# str(coef(mod))
# 
# 
# # names(belchers$info[[1]])
# # belchers %>% hoist(info,
# #   name = "name",
# #   age = "age",
# #   dad = "father",
# #   firstborn = list("children", 1L)
# # )
# 
# names(tmp2$sre[[1]])
# tmp2 %>%
#   ungroup %>%
#   hoist(sre,
#         # mod="mod",
#         coeff="coeff")
# 
# tmp2 %>%
#   unnest(sre)
# tmp2 %>%
#   unnest_wider(sre)
# 
# tmp2 %>%
#   hoist(sre, "coeff")
# 
# tmp2 %>%
#   unnest_wider(sre$coeff)
# 
# 
# df3 <- df2 %>%
#   unnest(cols = c(data, mod)) %>%
#   mutate(vol=value / trend * 100 - 100,
#          absvol=abs(vol)) %>%
#   group_by(stabbr, name) %>%
#   mutate(mdn=median(vol),
#          absmdn=median(absvol)) %>%
#   ungroup
# 
# volsum <- df3 %>%
#   group_by(name, stabbr) %>%
#   summarise(hpmdn=median(abs(vol)),
#             hpmean=mean(abs(vol)),
#             hpsd=sd(vol),
#             pchsd=sd(pch, na.rm=TRUE),
#             .groups="drop") %>%
#   arrange(-hpmdn)
# 
# volsum %>%
#   arrange(-hpmdn)
# 
# volsum %>%
#   filter(stabbr=="US")
# 
# volsum %>%
#   filter(name=="tottax")
# 
# # dev.off()
# 
# # US facet plot, one panel for each tax
# usdata <- df3 %>%
#   filter(stabbr=="US") %>%
#   group_by(stabbr, year) %>%
#   mutate(comp=vol[name=="tottax"]) %>%
#   select(stabbr, name, year, vol, comp) %>%
#   pivot_longer(cols=c(vol, comp), names_to = "type")
# 
# usdata %>%
#   ggplot(aes(year, value, colour=type)) +
#   geom_point() +
#   geom_line() +
#   geom_hline(yintercept = 0) +
#   facet_wrap(~name, ncol=3, scales="free")
# 
# stdata <- df3 %>%
#   group_by(name, year) %>%
#   mutate(us=vol[stabbr=="US"]) %>%
#   select(stabbr, name, year, vol, us) %>%
#   pivot_longer(cols=c(vol, us), names_to = "geo")
# 
# stdata %>%
#   filter(name=="tottax") %>%
#   filter(stabbr %in% c("CA", "NY", "AL", "MS", "MA", "CT", "OH")) %>%
#   ggplot(aes(year, value, colour=geo)) +
#   geom_point() +
#   geom_line() +
#   geom_hline(yintercept = 0) +
#   facet_wrap(~stabbr, ncol=3, scales="free")
# 
# 
# sts <- c("AL", "CA", "CT", "FL", "MA", "MS", "NY", "OH","OR", "PA", "TN", "TX", "UT", "VA", "WA")
# df3 %>%
#   filter(name=="tottax") %>%
#   filter(stabbr %in% sts) %>%
#   select(stabbr, name, year, absvol, absmdn) %>%
#   pivot_longer(cols=c(absvol, absmdn), names_to = "measure") %>%
#   ggplot(aes(year, value, colour=measure)) +
#   geom_point() +
#   geom_line() +
#   geom_hline(yintercept = 0) +
#   facet_wrap(~stabbr, ncol=3, scales="free")
# 
# sts <- c("AL", "CA", "CT", "FL", "MA", "MS", "NY", "OH","OR", "PA", "TN", "TX", "UT", "VA", "WA")
# df3 %>%
#   filter(stabbr=="US") %>%
#   select(stabbr, name, year, absvol, absmdn) %>%
#   pivot_longer(cols=c(absvol, absmdn), names_to = "measure") %>%
#   ggplot(aes(year, value, colour=measure)) +
#   geom_point() +
#   geom_line() +
#   geom_hline(yintercept = 0) +
#   facet_wrap(~name, ncol=3, scales="free")



```

