---
title: "State tax revenue volatility"
author: "Don Boyd"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_notebook:
    df_print: paged
    fig_height: 6
    fig_width: 8
    toc: yes
    toc_depth: 3
    toc_float: true
    number_sections: yes
editor_options:
  chunk_output_type: console
---


<!-- <style> -->
<!-- body .main-container { -->
<!--   max-width: 1500px; -->
<!-- } -->
<!-- </style> -->



```{r runall, eval=FALSE, echo=FALSE} 
# When we want a final report, run the following code selectively "by hand" (interactively)
# -- NEVER using Knit with eval=TRUE
# note <br> breaks line for html output but \n should break for pdf; also could try |

rmdfn <- "./Pew_Tax_Revenue_Volatility.rmd" # this file
outfn <- paste0("tax_volatility_", format(Sys.time(), "%Y-%m-%d"), ".html")
rmarkdown::render(rmdfn, output_format="html_document", output_file=outfn)

# library("RCurl")
# ftpUpload(outfn, "ftp://kffrig:boyd4812@files.000webhost.com/public_html/KFF_RIG_MedicaidCuts_new.html")

# Note may b safest to fully exit RStudio and restart it before running whole thing. Otherwise knitr may get confused
# and include repetitive information in the output html file.

```




```{r setup, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(echo = FALSE, include=FALSE, eval=FALSE, message = FALSE, warning = FALSE)
```


```{r css, echo=FALSE}
# blockquote {
#     padding: 20px;
#     margin: 0 0 20px;
#     font-size: 14px;
#     border-left: 5px solid #eee;
# }

```



<!-- comments -->

<!--
# code folding ----
#   alt-L, alt-shift-L  one section
#   alt-O, alt-shift-O  all sections
#   ctrl-D run current code section (remapped from Ctrl-alt-T)


always_allow_html: yes

  html_document:
    toc: yes
    toc_depth: '2'
    df_print: paged
    number_sections: yes



# renv ----
vignette("renv")
documentation online at https://rstudio.github.io/renv

The general workflow when working with renv is:

1. Call renv::init() to initialize a new project-local environment with a private R library,

2. Work in the project as normal, installing and removing new R packages as they are needed in the project,

3. Call renv::snapshot() to save the state of the project library to the lockfile (called renv.lock),

4. Continue working on your project, installing and updating R packages as needed.

5. Call renv::snapshot() again to save the state of your project library if your attempts to update R packages were successful, or call renv::restore() to revert to the previous state as encoded in the lockfile if your attempts to update packages introduced some new problems.


-->


<!-- git notes:
https://happygitwithr.com/git-commands.html
https://www.geos.ed.ac.uk/~smudd/NMDM_Course/html/more_advanced_github.html


git clone https://github.com/USERNAME/REPOSITORY.git  # get a repository (probably mine)

Check the remote was cloned successfully:
git remote --verbose

Stage local changes, commit:
git add foo.txt
git commit --message "A commit message"

Check on the state of the Git world:
git status
git log
git log --oneline

Compare versions:

git diff

Add a remote to existing local repo:

git remote add origin https://github.com/jennybc/happy-git-with-r
git remote --verbose
git remote show origin

Push local master to GitHub master and have local master track master on GitHub:

git push --set-upstream origin master
## shorter form
git push -u origin master
## you only need to set upstream tracking once!

$ git fetch remotename
# Fetches updates made to a remote repository

$ git merge remotename/branchname
# Merges updates made online with your local work

git pull is a convenient shortcut for completing both git fetch and git merge in the same command:

$ git pull remotename branchname
# Grabs online updates and merges them with your local work

git checkout -b issue-5  # create branch and checkout branch

work on changes, then
git commit --all -m "WIP"
git checkout master

Once you have done your work and committed it to the feature branch, you can switch back to master and merge the feature branch.

git checkout master
git merge issue-5


etc. one last change
-->


```{r renv_notes, eval=FALSE}
# https://rstudio.github.io/renv/articles/renv.html

# Custom R Package Repositories
# Custom and local R package repositories are supported as well. The only requirement is that these repositories
# are set as part of the repos R option, and that these repositories are named. For example, you might use:
# 
# repos <- c(CRAN = "https://cloud.r-project.org", WORK = "https://work.example.org")
# options(repos = repos)
# to tell renv to work with both the official CRAN package repository, as well as a package repository you have hosted and set up in your work environment.

# install local packages
# local_package_sources <- r"(C:\RPrograms PC\ProjectsCurrent\Pew_Revenue_Volatility\local\)"
# repos <- c(CRAN = "https://cloud.r-project.org", LOCAL = local_package_sources)
# options(repos = repos)

# note that tarballs are at: C:\RPrograms PC\Packages\bdata_0.6.tar.gz
# create tarballs of the desired packages and save them in the renv\local directory here, so that
# they will be portable. users will then have to install them from the renv\local directory
# steps below

# 1) Define directories
# pkgsource_dir is relevant for Boyd only - it is the location of source files for any packages I created
# pkgsource_dir <- r"(C:\RPrograms PC\Packages\)"  # raw string
# pkgdir is where "tarballs" of local packages will be placed -- this will be...
# pkgdir <- r"(C:\RPrograms PC\ProjectsCurrent\Pew_Revenue_Volatility\renv\local\)"

# 2) This step is for Boyd machine only:
# uncomment code below and build and save my local packages tarballs in the renv local directory
devtools::build(pkg = paste0(pkgsource_dir, "btools"),  # bdata BEAData btools
                path = pkgdir,
                binary = FALSE,
                vignettes = FALSE,
                manual = FALSE,
                args = NULL,
                quiet = FALSE)
# do the above for any other needed local packages

# 3) On Boyd machine and on user machine, install these packages from tarballs
# you may have to define the pkgdir location on your machine
# if the package is in use and you are updating it, then first: detach(package:bdata)
# pkgs <- c("bdata_0.7.tar.gz", "BEAData_0.6.0.tar.gz", "btools_0.9.tar.gz")

renv::install(paste0(pkgdir, "bdata_0.7.tar.gz"))
renv::install(paste0(pkgdir, "BEAData_0.6.0.tar.gz"))
renv::install(paste0(pkgdir, "btools_0.9.tar.gz"))
renv::snapshot()


# further notes
# to get it all started:
# renv::init()

# to remove it all:
# renv::deactivate()
# then delete the renv folder

# periodically execute:
# renv::snapshot()

# to see what directory is being used by renv for local package sources, execute:
# renv:::renv_paths_local()

# ${RENV_PATHS_LOCAL}/<package>_<version>.tar.gz

```


```{r get_libs_and_functions, include=FALSE, eval=TRUE}
source(here::here("r", "libs_base.r"))
source(here::here("r", "libs_ts.r"))
source(here::here("r", "functions.r"))
devtools::session_info()
# devtools::package_info()

```



# Project management and organization
## Deadlines and schedule

* July 3, 2020 Project start
* October 3, 2020 Provide list of external reviewers
* June 3, 2021  Draft report due (plan on sooner)
* August 3, 2021 Project end and final report due


## Links

+  [Zotero folder](https://www.zotero.org/groups/2541602/pew_tax_revenue_volatility/library)
+  [Google sheet](https://docs.google.com/spreadsheets/d/1hSWy_7ep4RbM9lfSA-LYnDeGI1c6p9Ud0GqcSlR9xJ8/edit?usp=sharing) (barely started) to summarize important articles
+  [Google drive folder](https://drive.google.com/drive/folders/1fM7HIP7qrKbIdndF_wQ30WxVb1j_Cq06)
    +   Draft report is stored here (latest version has highest number in parentheses)


## Goals and issues

* Ways to measure tax revenue volatility

* Recent trends in revenue volatility
  + Has tax revenue become more volatile over time?
  + If so, why?

* Drivers of state revenue volatility - to what extent driven by
  + The nature of national economic changes (e.g., characteristics of recession)?
  + State economic structure?
  + State tax structure? 
    + The portfolio of taxes? 
    + The structure of individual taxes?

*	Implications for states in current recession and beyond
  +  What does the Covid-19 recession imply for state tax volatility?
  +  How much can states reduce volatility with alternative tax structures?
    

Note: For details on references see [this](https://docs.google.com/spreadsheets/d/1hSWy_7ep4RbM9lfSA-LYnDeGI1c6p9Ud0GqcSlR9xJ8/edit?usp=sharing).


## Notes to reviewers

Use renv.



## Data tasks and status
### Done

* State tax revenue from Census by tax, state, and year, 1951-2019, collapsed to the following categories (currently):
  + total taxes (tottax)
  + individual income tax (iit)
  + general sales tax (gst)
  + selective sales taxes such as tobacco, motor fuel, motor vehicle fees, alcohol… (selsalestax)
  + corporate income tax (cit)
  + severance taxes (sevtax)
  + all other taxes (othertax)
* Pew data on policy changes  
* Economic data
  + gross domestic product, by state fiscal year, 1951-2019 (still must adjust for states with fiscal years different from July 1)
  + gross domestic price index (chained), by state fiscal year, 1951-2019 (same adjustment required)
  + official NBER dates for recessions during relevant time periods
  + obtained NIPA details for description of recessions
  + obtained state-level consumption data from BEA

### TODO

* organize NIPA data to measure recessions
* organize NIPA data for consumption details
* organize state consumption data
* Possibly get data from Dauchy
* Graph pch tax vs econ
* IRS data components income
* Measures of income tax progressivity
* Similar measures sales tax?
* Consider base = revenue / rate as robustness check

## Details about the project

### Overall goals

Examine tax revenue volatility and its impacts on state governments

Provide deep and broad understanding of:

* Ways to measure tax revenue volatility
* Recent trends in revenue volatility
* Drivers of state revenue volatility
*	Implications for states in current recession and beyond

Ultimate goal: Insights that can help policymakers manage budgets


### Important sub-questions

* To what extent is volatility driven by:
    + The nature of national economic changes (e.g., characteristics of recession)?
    + State economic structure?
    + State tax structure? The portfolio of taxes? The structure of individual taxes?
* Has tax revenue become more volatile over time? If so, why?
* How much can states reduce volatility with alternative tax structures?

All have implications for states in current recession and beyond.


### Overall plan

1. Review relevant literature
2. Build database for analyzing volatility
    + Annual state tax revenue, all states, by tax
    + Auxiliary data to allow population & inflation adjustment
    + Selected other data
3. Multiple methods
    + Descriptive analytic review of history
    + Decomposition
    + Simulations of portfolio effects
    + Exploratory econometrics for certain topics


### Core data source

* Census Bureau Annual Survey of State Finances
  + State tax revenue, annual, all states, many taxes
  + Comprehensive (not just general fund)
  + Relatively consistent definitions over time and states
  + Annual period consistent with policymaking cycle
  + Includes multiple recessions (some data back to mid-1950s):
      + 1980-81, 1990, 2001, and 2007 (2020 won't be available)
      + may be able to include 1960, 1969, 1973 recessions

### Selected technical & data issues

* Tax data implicitly include embedded policy changes. Use Pew tax-change data to help answer: How much does adjusting for policy changes affect conclusions?
* How much do alternative volatility measures affect conclusions?
* Data on structure of individual taxes are limited. To extent practical, use these limited data to help answer: How does structure of individual taxes affect volatility?
  

### Topics and approaches
+----------------------------------+----------------------------------------------+
| Topic                            | Methods                                      |
+==================================+==============================================+
| Ways to measure                  | - Literature review                          |
| revenue volatility               | - Conceptual differences among measures      |
|                                  | - Trends with different measures             |
+----------------------------------+----------------------------------------------+
| Recent revenue volatility trends | - Descriptive analysis                       |
+----------------------------------+----------------------------------------------+
| Drivers of revenue volatility    | - Descriptive analysis                       |
|   (Nature of economic change,    | - Decomposition                              |
|    State economic structure,     | - Econometric analysis                       |
|    State tax structure)          | - Simulation of alternative structures       |
+----------------------------------+----------------------------------------------+
| Implications in current          | - Synthesis of above                         |
| recession and beyond             |                                              |
+----------------------------------+----------------------------------------------+


### Comments from the August 4, 2020 methods review

* Clarify key terms with thorough definitions and describe the applied metrics, highlighting areas of standardization. Absent those details, some confusion may arise. Specifically,
    +	Convey the practical differences between the proposed metrics of revenue volatility for state and other policymakers. These metrics can vary considerably depending on the underlying data distribution, meaningfully changing what is observed as volatility. 
    + Elaborate on how revenue streams may need to be combined/collapsed for analytical tractability. If doing so, highlight the thresholds for those modification, if these differ by state or source, and if those decisions affect the results.  
    + Account for the possible interpretations of concepts like “state economic structures” and economies “at risk”, and how those choices shift the meaning of the analysis and findings. If these concepts are applied differently by state, explain how these may be standardized.  

* Ensure that the selected time frames (periods and trends) and their decision rules are transparent in the final report. Specifically: 
    + Discuss how partial recessionary periods (i.e. quarters) will be addressed since the study will generally leverage annual data, to speak to policymaker budgeting cycles/priorities.
    +	Clarify that “recent trends” may reflect multi-decade periods of fiscal management, which may not be self-evident. Explain how the bounds of “trend periods” are chosen, whether through an assessment of business cycles, regression analyses, inductive analyses or other techniques.

* Ensure consistency between this and prior Pew studies, or provide a crosswalk, particularly if the methods have been modified or augmented. Focus on what sources of uncertainty remain embedded in each measure of volatility. 
    + Assess whether correcting for tax policy adjustments changes estimated volatility. Also, since the policy adjusted time series is shorter, estimate whether a shorter series will complicate the development of robust volatility estimates using the more technical/complex approaches. 
    + Highlight — for more technical methods — how trends are being estimated, and how differences in those trends might change the interpretation of volatility, as the embedded sources of uncertainty differ (e.g. by extracting cyclicality). 






# Data preparation and checking

Certain chunk names in this section start with "ONETIME". In those chucks, I create and save data that will be later read from files. The data only need to be created one time (or updated and re-saved if I decide to update the data).

## Create data frame indicating the start of each state's fiscal year

So that we can adjusting certain data to correspond to state fiscal years.

*   46 states have a July 1 fiscal year
*   NY has April 1
*   TX has Sept 1
*   AL, MI have Oct 1
*   For example, see [this](https://www.ncsl.org/research/fiscal-policy/fy-2021-state-budget-status.aspx)


```{r state_fiscal_years}
# state fiscal year starts  https://www.ncsl.org/research/fiscal-policy/fy-2021-state-budget-status.aspx
# NY April, TX Sept, AL & MI Oct, 46 states July
sfy_startmonths <- tibble(stabbr=state.abb) %>%
  mutate(startmonth=case_when(stabbr=="NY" ~ 4,
                              stabbr=="TX" ~ 9,
                              stabbr %in% c("AL", "MI") ~ 10,
                              TRUE ~ 7))


```

## NIPA national data

*   I have a package called `BEAData` that includes virtually all major data from the Bureau of Economic Analysis (BEA) in relation to the National Income and Products Accounts (NIPA) and in relation to regional data.
*   For details on data creation see the [github site](https://github.com/donboyd5/BEAData).
*   The package is stored locally in the renv/local folder and can be installed from there.
*   `data(package="BEAData")` shows the files available
*   NIPA national data are in dataframe `nipa`
*   `comment(nipa)` shows creation date

In this section I create data frames with national GDP (nominal) and the GDP price index, on a state fiscal year basis.


```{r nipa_national}
glimpse(nipa)
nipa %>%
  filter(freq=="Q", str_detect(vdesc, "Gross domestic product, Current")) %>%
  group_by(vname, vdesc) %>%
  filter(date==max(date))
# A191RC US nominal gdp
# "A191RG" US GDP price index

gdpfy <- nipa %>%
  filter(vname=="A191RC", freq=="Q") %>%
  # fix NY AL MI TX
  mutate(fyear=ifelse(month(date) >= 7, year + 1, year) %>% as.integer) %>%
  group_by(vname, fyear) %>%
  summarise(gdp=mean(value, na.rm=TRUE), .groups="drop") %>%
  filter(fyear < 2020) %>%
  select(year=fyear, gdp)
gdpfy %>% ht  # 1947, 2019

 
baseyear <- 2019
gdppi <- nipa %>%
  filter(vname=="A191RG", freq=="Q") %>%
  # fix NY AL MI TX
  mutate(fyear=ifelse(month(date) >= 7, year + 1, year) %>% as.integer) %>%
  group_by(vname, fyear) %>%
  summarise(gdppi=mean(value, na.rm=TRUE), .groups="drop") %>%
  filter(fyear < 2020) %>%
  select(year=fyear, gdppi) %>%
  mutate(igdppi=gdppi[year==baseyear] / gdppi) 

recessions # 1857, ..., 2020

```


```{r nipa_national_data_for_recessions}
# https://apps.bea.gov/iTable/iTable.cfm
# we want two kinds of breakdowns:
#    C + I + G + (X - M)
#    Industrial breakdown

#.. CIGXM ----
# Table 1.1.6. Real Gross Domestic Product, Chained Dollars
tab <- NIPAvars %>%
  filter(tabnum == "1.1.6") %>%
  arrange(line)

cigxm_vars <- tab %>%
  filter(line %in% c(1:6, 7, 8, 14, 15, 15, 22))
cigxm_vars

cigxm <- nipa %>%
  right_join(cigxm_vars %>% 
               select(vname), 
             by = "vname")


#.. Industrial breakdown ----
# A150RX
NIPAvars %>%
  filter(vname == "A150RX") 

tmp <- NIPAvars %>%
  filter(str_detect(vdesc, coll("manuf", ignore_case = TRUE))) 
 
mtables <- count(tmp, tabname)
# https://apps.bea.gov/industry/Release/XLS/GDPxInd/ValueAdded.xlsx
# https://apps.bea.gov//industry/iTables%20Static%20Files/AllTablesHist.zip



# Table 5.8.6B. Real Private Inventories and Real Domestic Final Sales by Industry, Chained Dollars
# Table 6.16D. Corporate Profits by Industry
# Table 6.5D. Full-Time Equivalent Employees by Industry

#..Employment breakdown ----
# https://www.bls.gov/web/empsit/cesseriespub.htm
# https://download.bls.gov/pub/time.series/ce/ce.data.01a.CurrentSeasAE  
# https://download.bls.gov/pub/time.series/ce/ce.txt says:
# ce.data.01a.CurrentSeasAE - contains every seasonally adjusted all employee series and its complete history.    

#..Causes and characteristics ----
# https://www.investopedia.com/articles/economics/08/past-recessions.asp
# inflation: 1969
# oil and energy crises: 1973, 1980, 1981
# war: 1990 (Gulf)
# 9/11:  2001
# financial crisis: 2007 Great Recession
# real estate: 2007 GR
# 
#
# variables:
# oil prices
# real estate prices
# stock market
# manufacturing, tourism, other sectors
# unemployment rate

```
    
    
## Census Bureau tax data

### Clean tax data

Steps:

1.    Get Census state govt tax data from bdata (previously created)
2.    Keep desired taxes and create an other taxes category
3.    Identify and explore anomalies

```{r get_tax}
# tax1 has untouched data from Census, only the items we want, long format
# tax2 is same, but with the other tax variable


glimpse(sgtax.a)  # 2019 end year; most values from ~1951+
# count(sgtax.a, year) %>% as.data.frame()
# count(sgtax.a, item, desc)

# determine the tax items we want
taxnames <- sgtax.a %>% select(item, desc) %>% distinct()
items <- c("C105", "T09", "C109", "T40", "T41", "T53")
varnames <- c("tottax", "gst", "selsalestax", "iit", "cit", "sevtax")
sumvars <- setdiff(varnames, "tottax")


tax1 <- sgtax.a %>%
  filter(year >= 1951) %>%
  filter(item %in% items) %>%
  mutate(name=factor(item, levels=items, labels=varnames)) %>%
  select(stabbr, name, year, value) %>%
  arrange(stabbr, name, year)
summary(tax1)  # no missing values

# tax2 creates a wide version
tax2 <- tax1 %>%
  # now calculate other tax
  pivot_wider(names_from = name) %>%
  mutate(nonother = rowSums(across(all_of(sumvars)), na.rm=TRUE),
         othertax=tottax - nonother) %>%
  select(-nonother) %>%
  pivot_longer(-c(year, stabbr)) %>%
  filter(!is.na(value))  # we can safely drop NA values because we didn't have any in tax1
summary(tax2)

```

Identify and explore anomalies:

+   missing
+   zero
+   missing or zero at start (always assumed bad)
+   missing or zero at end (always assumed bad)
+   missing in middle
+   zero in middle
+   negative in middle (logs won't work -- only 1 found)
+   short runs (fewer than X positive values in a row)


```{r anomalies}
# identify different anomalies
# CO sevtax  2017 0
# OH cit 2014 -118 
tax2 %>% filter(stabbr=="AK", name=="iit")
tax2 %>% filter(stabbr=="AK", name=="gst")
tax2 %>% filter(stabbr=="AZ", name=="sevtax")
tax2 %>% filter(stabbr=="CO", name=="sevtax")
tax2 %>% filter(stabbr=="OH", name=="cit")
tax2 %>% filter(stabbr=="NH", name=="sevtax")
tax2 %>% filter(stabbr=="SD", name=="sevtax")

# vectors used to test the functions below
a1 <- c(0, 0, NA, 1, 2, 3, 0, 4, 5, 6, 0, 0)
a2 <- c(0, 0, 0, 0)
a3 <- c(0, NA, 0, 0)
a4 <- c(7, 9, 0, 0, NA, 1, 2, 3, 0, 4, 5, 6)

firstnz <- function(x) {
  # find index of first nonzero value, to help us identify zeroes at the start
  ifelse(all(x==0, na.rm=TRUE), 
         Inf, 
         which(x != 0)[[1]])
}

# test firstnz
firstnz(a1)
firstnz(a2)
firstnz(a3)
firstnz(a4)

lastnz <- function(x) {
  # find index of last nonzero value, to help us identify zeroes at the end
  ifelse(all(x==0, na.rm=TRUE),
         -Inf, 
         tail(which(x != 0), 1))
}

# test lastnz
lastnz(a1)
lastnz(a2)
lastnz(a3)
lastnz(a4)


firstgap <- function(gap) {
  # gap is a a vector of 0 or 1 indicators for a series
  # 0 indicates the observation is not part of a gap and 1 indicates it is part of a gap
  # firstgap finds the index at which a gap in a series starts (if a gap exists)
  ifelse(all(gap !=TRUE, na.rm=TRUE),
         Inf, # return Inf if there is no gap in the series
         which(gap==TRUE)[[1]]) # index of the start of the first gap
}


lastgap <- function(gap) {
  # index of the final observation in last gap
  ifelse(all(gap !=TRUE, na.rm=TRUE),
         -Inf, # return -Inf if there is no gap in the series
         tail(which(gap==TRUE), 1)) # index of the final observation in the last gap of a series
}


# identify anomalies within each state-tax combination
# also, for series that have gaps in them, identify observations that occur before the first gap or
# after the last gap, as we will only want to keep one or the other
anomalies <- tax2 %>%
  group_by(stabbr, name) %>%
  mutate(missval = is.na(value),  # we should not have any missing values but check anyway
         zeroval = value == 0,
         negval = value < 0,
         leadzero = row_number() < firstnz(value),   # identify all leading-zero rows
         trailzero = row_number() > lastnz(value),  # identify all trailing-zero rows
         gap = zeroval & !(leadzero | trailzero),  # a gap occurs if we have a zero that is not leading or trailing
         pregap = row_number() < firstgap(gap),   # identify data that occurs before the first gap
         postgap = row_number() > lastgap(gap))   # identify data that occurs after the last gap
summary(anomalies)
# what should we do if a series has multiple gaps?? look to make sure it doesn't happen; I don't think so

anomalies %>% filter(stabbr=="AK", name=="iit")
anomalies %>% filter(stabbr=="AK", name=="gst")
anomalies %>% filter(stabbr=="AZ", name=="sevtax")
anomalies %>% filter(stabbr=="AZ", name=="cit") # gap in 1957
anomalies %>% filter(stabbr=="CO", name=="sevtax")
anomalies %>% filter(stabbr=="OH", name=="cit")
anomalies %>% filter(stabbr=="NH", name=="sevtax")
anomalies %>% filter(stabbr=="NV", name=="sevtax")
anomalies %>% filter(stabbr=="SD", name=="sevtax")
anomalies %>% filter(gap)


# new line added here

anomalies %>% 
  filter(stabbr=="NV", name=="sevtax") %>%
  lastgap()

# rules:
# - drop all lead zero and trail zero
# - take the post gap periods
# individually and decide - take the post gap

anomalies2 <- anomalies %>%
  filter(!(leadzero | trailzero))


ibad <- which(a %in% c(0, NA_real_))
Position(f, x, right = FALSE, nomatch = NA_integer)


anomalies %>%
  select(stabbr, name, firstyear) %>%
  unique() %>%
  filter(firstyear != 1951)

anomalies %>%
  select(stabbr, name, fy2) %>%
  unique() %>%
  filter(fy2 != 1951)

# text added from winbranch

# text added from linbranch

```





```{r tax_data, eval=TRUE}
glimpse(sgrev)


glimpse(sgtax.a)  # 2019 end year; most values from ~1951+
# count(sgtax.a, year) %>% as.data.frame()
# count(sgtax.a, item, desc)

# data(package="BEAData")


# collapse taxes
taxnames <- sgtax.a %>% select(item, desc) %>% distinct()
items <- c("C105", "T09", "C109", "T40", "T41", "T53")
varnames <- c("tottax", "gst", "selsalestax", "iit", "cit", "sevtax")
sumvars <- setdiff(varnames, "tottax")

tax1 <- sgtax.a %>%
  filter(year >= 1951) %>%
  filter(item %in% items) %>%
  mutate(name=factor(item, levels=items, labels=varnames)) %>%
  select(stabbr, name, year, value) %>%
  
  # now calculate other tax
  pivot_wider(names_from = name) %>%
  mutate(nonother = rowSums(across(all_of(sumvars)), na.rm=TRUE),
         othertax=tottax - nonother) %>%
  select(-nonother) %>%
  pivot_longer(-c(year, stabbr))
tax1
summary(tax1) # good no NA values
count(tax1, stabbr)
count(tax1, name)
tax1 %>% filter(is.na(value))
tax1 %>% filter(stabbr=="AK")

# determine first and last nonzero and nonmissing year for every state-tax combination
# and keep just the years we want
good_firstlast <- tax1 %>%
  filter(value != 0, !is.na(value)) %>%
  group_by(stabbr, name) %>%
  summarise(first=min(year), last=max(year), .groups="drop") %>%
  mutate(nyears=last - first + 1)
summary(good_firstlast)

good_firstlast %>% filter(stabbr=="OH")
sum(good_firstlast$nyears)

tax2 <- tax1 %>%
  left_join(good_firstlast, by = c("stabbr", "name")) %>%
  filter(year >= first, year <=  last) %>%
  select(stabbr, year, name, value)
  
tax1 %>% filter(stabbr=="CT", name=="iit")
tax1 %>% filter(stabbr=="DE", name=="gst") # this looks wrong; make sure we don't calc volatility
tax2 %>% filter(stabbr=="DE", name=="gst")

tax_stack <- tax2 %>%
  rename(nominal=value) %>%
  left_join(gdppi %>% select(year, igdppi), by="year") %>%
  mutate(real=nominal * igdppi) %>%
  select(-igdppi) %>%
  pivot_longer(c(nominal, real), names_to="type")

```


```{r}
# data checking
# what to do about negative tax values?? they will become nan in logs, and we need to be robust to that
# what about zero values in the middle that look like they must be missing
# OH cit 2014 is negative (-$118)
# NM cit has pos 1951-1954, zero 1955-1966, then positive

tax_stack %>%
  filter(value <= 0) %>%
  as.data.frame()

tax_stack %>%
  filter(stabbr=="OH", name=="cit", type=="nominal")

tax_stack %>%
  filter(stabbr=="NM", name=="cit", type=="nominal")

measures %>%
  filter(stabbr=="OH", name=="cit", type=="nominal")
```


### Basic tax measures
```{r}


measures <- tax_stack %>%
  group_by(stabbr, name, type) %>%
  mutate(lvalue=log(value))

#   nest() %>%

```



## Pew data on enacted revenue changes

*   Pew constructed estimates of the percent change in major taxes, by year, state, and major tax type for work it did in ___
*   In a 10/8/2015 email, on my request Brenna Erford <BErford@pewtrusts.org> provided the file "Boyd Data Request 100815.xlsx" to me (donald.boyd@rockinst.suny.edu), which had data underlying the project
*   Pew's short description of the data, in the "Cover" tab of the workbook, was:
    
<blockquote style="color:grey;font-size:100%;margin-left: 4em">
  <p>Volatility scores for each state’s total tax revenue and specific tax sources were calculated using the U.S. Census Bureau’s State Government Tax Collections historical data series for 1994 to 2014, accessed April 21, 2015. Data were adjusted to control for the effects of tax policy change using the National Conference of State Legislatures (NCSL) State Tax Action reports for 1994 to 2013, accessed in January 2015.</p>
</blockquote>


```{r ONETIME_pew_revenue_data, eval=FALSE}
# data file Boyd Data Request 100815.xlsx
xlfn <- here::here("data", "Boyd Data Request 100815.xlsx")
sheets <- excel_sheets(xlfn)
sheets

get_pewsheet <- function(sheet){
  print(sheet)
  range <- ifelse(str_detect(sheet, "Major"), "A5:V190", "A5:V56")
  read_excel(xlfn, sheet=sheet, range=range) %>%
    mutate(sheet=sheet)
}

pewdf <- map_dfr(sheets[-1], get_pewsheet)
glimpse(pewdf)
count(pewdf, sheet)
count(pewdf, `Major Revenue Source`)

#    `Major Revenue Source`               n
#    <chr>                            <int>
#  1 Amusements                           9
#  2 Corporation                          9
#  3 Corporation net income              90
#  4 Documentary and stock transfer       6
#  5 General sales and gross receipts   135
#  6 Individual income                  123
#  7 Motor fuels                         99
#  8 Motor vehicle                       12
#  9 Occupation and business, NEC         6
# 10 Property taxes                      27
# 11 Public utilities                     9
# 12 Severance                           27
# 13 Tobacco products                     3
# 14 Total                              153

type_map <- read_delim("revsource; taxtype
Amusements; amusetax
Corporation; corptax
Corporation net income; cit
Documentary and stock transfer; doctax
General sales and gross receipts; gst
Individual income; iit
Motor fuels; mft
Motor vehicle; mvl
Occupation and business, NEC; occtax
Property taxes; proptax
Public utilities; utiltax
Severance; sevtax
Tobacco products; cigtax
Total; tottax", 
delim=";", trim_ws = TRUE)

pewlong <- pewdf %>%
  select(stname=1, revsource=2, everything()) %>%
  left_join(bdata::stcodes %>% select(stabbr, stname), by="stname") %>%
  left_join(type_map, by="revsource") %>%
  mutate(vartype=word(sheet, 3)) %>%
  pivot_longer(-c(stabbr, stname, revsource, sheet, vartype, taxtype), names_to = "year") %>%
  mutate(year=as.integer(year)) %>%
  select(stabbr, stname, sheet, vartype, revsource, taxtype, year, value) %>%
  arrange(stabbr, vartype, taxtype, year)
  
glimpse(pewlong)

# caution: US data do not have major taxes, just the total
count(pewlong, stabbr, stname)

pewlong %>% filter(stabbr=="US")
saveRDS(pewlong, here::here("data", "pewlong.rds"))


```





```{r vol_measures}
# hp_df <- function(df, vname){
#   vts <- ts(df[vname])
#   vts_hp <- hpfilter(vts, freq=NULL, type=c("lambda"), drift=FALSE)
#   df %>% mutate(trend=as.numeric(vts_hp$trend),
#                 cycle=as.numeric(vts_hp$cycle))
# }


# takes a LONG LONG time
# vol1 <- tax_real %>%
#   filter(name=="tottax") %>%
#   group_by(stabbr, name, type) %>%
#   arrange(year) %>%
#   hp_df(., "value")
#   



# df <- df %>%
#   f(., "iit") %>%
#   mutate(dtrend=cycle / trend * 100,
#          pch=iit / iit[match(year - 1, year)] * 100 - 100,
#          dtrend_sd10=rollsd_p(dtrend, 10),
#          pch_sd10=rollsd_p(pch, 10))


```



```{r explore}
# remove trend
df <- sgtax.a %>%
  filter(year >= 1951, stabbr=="NY", item=="T40") %>%
  select(year, stabbr, iit=value)
df %>% ht  

# iit <- ts(df$iit, start = 1951, end=2019)
# iit <- ts(df$iit)
# plot(iit)
# 
# stl(iit, s.window = "periodic")
# # hpfilter(iit, freq=NULL, type=c("lambda","frequency"),drift=FALSE)
# iit_hp <- hpfilter(iit, freq=NULL, type=c("lambda"), drift=FALSE)
# iit_hp$trend
# iit_hp$cycle

f <- function(df, vname){
  vts <- ts(df[vname])
  vts_hp <- hpfilter(vts, freq=NULL, type=c("lambda"), drift=FALSE)
  df %>% mutate(trend=as.numeric(vts_hp$trend),
                cycle=as.numeric(vts_hp$cycle))
}

df <- df %>%
  f(., "iit") %>%
  mutate(dtrend=cycle / trend * 100,
         pch=iit / iit[match(year - 1, year)] * 100 - 100,
         dtrend_sd10=rollsd_p(dtrend, 10),
         pch_sd10=rollsd_p(pch, 10))

glimpse(df)


df %>%
  select(year, iit, trend) %>%
  pivot_longer(cols=c(iit, trend)) %>%
  ggplot(aes(year, value, colour=name)) +
  geom_line() +
  geom_point()


df %>%
  ggplot(aes(year, dtrend)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0)
  
df %>%
  ggplot(aes(year, pch)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0)

# df %>%
#   select(year, dtrend, pch) %>%
#   pivot_longer(cols=-year) %>%
#   ggplot(aes(year, value, colour=name)) +
#   geom_line() +
#   geom_point() +
#   geom_hline(yintercept = 0)
# 
# df %>%
#   select(year, dtrend, pch) %>%
#   mutate(pch=pch - mean(pch, na.rm=TRUE)) %>%
#   pivot_longer(cols=-year) %>%
#   ggplot(aes(year, value, colour=name)) +
#   geom_line() +
#   geom_point() +
#   geom_hline(yintercept = 0)

df %>%
  select(year, dtrend_sd10, pch_sd10) %>%
  pivot_longer(cols=-year) %>%
  ggplot(aes(year, value, colour=name)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0)
 
```


## Calculate volatility measures

```{r vfunctions}
hp_df <- function(df, vname, smooth=NULL){
  vts <- ts(df[vname])
  vts_hp <- hpfilter(vts, freq=smooth, type=c("lambda"), drift=FALSE)
  df %>% 
    mutate(trend=as.numeric(vts_hp$trend),
           cycle=as.numeric(vts_hp$cycle))
}

```


```{r}
tax_stacked
count(tax_stacked, name)
tax_stacked %>% 
  filter(type=="nominal", name=="tottax") %>%
  group_by(stabbr) %>%
  summarise(firstyear=min(year),
            lastyear=max(year))  # 1951-2019

# differences from actual revenue
# prep the tax file -- determine first year of each series!!! and start there
taxprep <- tax_stacked %>%
  arrange(stabbr, name, type, year) %>%
  group_by(stabbr, name, type) %>%
  mutate(ratio=value / value[match(year - 1,  year)],
         logpch=log(ratio) * 100,
         pch=value / value[match(year - 1,  year)] * 100 - 100)
taxprep
# note logpch warning in  stabbr = "OH", name = "cit",

# separate files with the different measures
# hpfilter all items

hp2 <- function(df, vname, smooth=NULL){
  vts <- ts(df[vname])
  vts_hp <- hpfilter(vts, freq=smooth, type=c("lambda"), drift=FALSE)
  tibble(trend=as.numeric(vts_hp$trend),
         cycle=as.numeric(vts_hp$cycle))
}

a <- proc.time()
hpall <- tax_stacked %>%
  # filter(stabbr=="OH") %>%
  group_by(stabbr, name, type) %>%
  nest()  %>%
  mutate(hp=map(data, hp2, "value", smooth=100))
b <- proc.time() 
b - a

glimpse(hpall)

hpall %>%
  filter(stabbr=="OH", type=="nominal", name=="cit") %>%
  unnest(cols = c(data, hp))


```



# Materials for report, organized by topic

## Introduction and motivation

### Setting the scene - year over year % change
```{r pchplot, fig.height=4.5, fig.width=8, eval=TRUE, include=TRUE}
# sd10c
#  pchplot, fig.height=4.5, fig.width=8, eval=FALSE}
# summary(sgtax.a)
irecs <- which(recessions$rec_year >= 1960)

p2 <- taxgdpus %>%
  filter(year >= 1960) %>%
  ggplot(aes(year, vpch, colour=name)) +
  geom_line(size=1) +
  geom_point(size=1.15) +
  scale_y_continuous(name="% change", breaks=seq(-20, 20, 2)) +
  scale_x_continuous(name=NULL, breaks=seq(1960, 2030, 5)) +
  scale_colour_manual(values=c("red", "blue")) +
  geom_hline(yintercept = 0) +
  annotate("rect",
           xmin = recessions$peak[irecs] %>% decimal_date(),
           xmax = recessions$trough[irecs] %>% decimal_date(),
           fill = "lightgrey",
           alpha = .5, # larger alpha is darker rectangle
           ymin = -Inf, ymax = Inf) +
  ggtitle("Year over year growth in state government tax revenue and GDP for the United States",
          subtitle="Recession periods are shaded.") +
  labs(caption="Source: U.S. Bureau of the Census and Bureau of Economic Analysis") +
  theme_bw() +
  theme(plot.caption = element_text(hjust=0, size=9)) +
  theme(legend.title = element_blank())
p2
ggsave(filename = here::here("results", "tax_gdp_us.png"), plot = p2)


```


## Ways to measure tax revenue volatility

* Past research
  + Rolling variance (Dauchy 2013)
  + Deviation from trend
    + Kwak 2013 says White 1983; Dye and McGuire 1991; Braun and Otsuka 1998; Aisen and Jose´ Veiga 2008 all used deviation from trend
    + Orthogonal Distance Regression (Kwak 2013)
     

### Selected measures for the United States




#### Rolling standard deviation


#### Deviation from trend


## Recent trends in revenue volatility


## Drivers of state revenue volatility

### The nature of national economic changes

### State economic structure

### State tax structure

#### The portfolio of state taxes

#### The structure of individual taxes


##	Implications for states in current recession and beyond

### Special characteristics of the Covid-19 recession and state tax volatility

### How much can states reduce volatility with alternative tax structures?


# Appendix

## Colorado revenue volatility review
```{r covol}


# lambda
freq <- 1
lam <- (2 * sin(pi / freq))^(-4)  # lambda w/freq=1: 2.779035e+62


conom <- slgfin %>%
  filter(stabbr=="CO", level==2, aggvar=="osr", year %in% 1992:2017) %>%
  arrange(year)

cohp <- conom %>%
  mutate(return=value / lag(value),
         dreturn=log(return) * 100) %>%
  filter(year %in% 1993:2017) %>%
  hp_df(., "dreturn", smooth=1600) %>%
  mutate(vol=cycle)
sum(cohp$cycle)

cohp_dtrend <- conom %>%
  filter(year %in% 1993:2017) %>%
  hp_df(., "value", smooth=1600) %>%
  mutate(vol=cycle / trend * 100)

comp <- bind_rows(cohp %>% select(year, vol) %>% mutate(name="dpch"),
                  cohp_dtrend %>% select(year, vol) %>% mutate(name="dtrend"))

comp %>% 
  pivot_wider(values_from = vol)
  

cohp %>%
  mutate(vol=cycle) %>%
  ggplot(aes(year, vol)) +
  geom_bar(stat="identity", fill="blue") +
  scale_x_continuous(breaks=1993:2017) +
  scale_y_continuous(breaks=seq(-20, 20, 2.5)) +
  theme(axis.text.x = element_text(angle = 45, vjust=0, hjust=0))

cohp %>%
  select(year, dreturn, trend) %>%
  pivot_longer(cols=-year) %>%
  ggplot(aes(year, value, colour=name)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 0)

comp %>%
  ggplot(aes(year, vol, colour=name)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 0)

# djb new start ----
param <- 1600

cohp <- conom %>%
  mutate(return=value / lag(value),
         dreturn=log(return) * 100) %>%
  filter(year %in% 1993:2017) %>%
  hp_df(., "dreturn", smooth=param) %>%
  mutate(vol=cycle)

cohp_dtrend <- conom %>%
  filter(year %in% 1993:2017) %>%
  hp_df(., "value", smooth=param) %>%
  mutate(vol=cycle / trend * 100)

comp <- bind_rows(cohp %>% select(year, vol) %>% mutate(name="dpch"),
                  cohp_dtrend %>% select(year, vol) %>% mutate(name="dtrend"))

p <- comp %>%
  filter(year %in% 1993:2017) %>%
  mutate(name=factor(name,
                     levels=c("dpch", "dtrend"),
                     labels=c("Calculated as footnote describes",
                              "Calculated as graph is labeled"))) %>%
  ggplot(aes(year, vol)) +
  geom_bar(stat="identity", fill="blue") +
  scale_x_continuous(name=NULL, breaks=1993:2017) +
  scale_y_continuous(name="Percent", breaks=seq(-20, 20, 5), limits=c(-15, 10)) +
  facet_wrap(~name, ncol=1) +
  ggtitle("Colorado's revenue deviation from trend calculated two ways",
          subtitle=paste0("HP filter smoothing parameter: ", param)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=0.5))
p



cotaxhp <- slgfin %>%
  filter(stabbr=="CO", level==2, aggvar=="tottax", year %in% 1992:2017) %>%
  arrange(year) %>%
  mutate(return=value / lag(value),
         dreturn=log(return) * 100) %>%
  filter(year %in% 1993:2017) %>%
  hp_df(., "dreturn", smooth=1600)

cohp %>%
  select(year, cycle) %>%
  mutate(name="osr") %>%
  bind_rows(cotaxhp %>% 
              select(year, cycle) %>%
              mutate(name="tax")) %>%
  ggplot(aes(year, cycle, colour=name)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 0)

co_vol <- tax_stacked %>%
  filter(name=="tottax", type=="real", stabbr=="CO", year %in% 1993:2017) %>%
  group_by(stabbr, name, type) %>%
  arrange(year) %>%
  hp_df(., "value")


co_vol %>%
  mutate(vol= cycle / trend * 100) %>%
  ggplot(aes(year, vol)) +
  geom_bar(stat="identity")


co_vol2 <- tax_stacked %>%
  filter(name=="tottax", type=="nominal", stabbr=="CO", year %in% 1992:2017) %>%
  group_by(stabbr, name, type) %>%
  arrange(year) %>%
  mutate(return=value / lag(value),
         dreturn=log(return) * 100) %>%
  filter(year %in% 1993:2017) %>%
  hp_df(., "dreturn")

co_vol2

co_vol2 %>%
  mutate(vol= cycle / trend * 100) %>%
  ggplot(aes(year, vol)) +
  geom_bar(stat="identity")

tax_stacked %>%
  filter(name=="tottax", type=="nominal", stabbr=="CO", year %in% 1992:2017) %>%
  group_by(stabbr, name, type) %>%
  arrange(year) %>%
  mutate(pch=value / lag(value) * 100 - 100) %>%
  filter(year %in% 1993:2017) %>%
  mutate(diff=pch - median(pch))


data(package="bdata")

tmp <- slgfin %>%
  filter(stabbr=="CO", level==2, aggvar=="osr", year %in% 1992:2017)

cov3 <- tmp %>%
  arrange(year) %>%
  mutate(return=value / lag(value),
         dreturn=log(return) * 100) %>%
  filter(year %in% 1993:2017) %>%
  hp_df(., "dreturn") %>%
  filter(year %in% 1993:2017)


```



# Older stuff below here



```{r pchplot_old, fig.height=4.5, fig.width=8, eval=FALSE}
# summary(sgtax.a)
irecs <- which(recessions$rec_year >= 1960)

p1 <- sgt1 %>%
  filter(year >= 1960) %>%
  ggplot(aes(year, vpch)) +
  geom_line(colour="blue", size=1) +
  geom_point(colour="blue", size=1.15) +
  scale_y_continuous(name="% change", breaks=seq(-20, 20, 2)) +
  scale_x_continuous(name=NULL, breaks=seq(1960, 2030, 5)) +
  geom_hline(yintercept = 0) +
  annotate("rect",
           xmin = recessions$peak[irecs] %>% decimal_date(),
           xmax = recessions$trough[irecs] %>% decimal_date(),
           fill = "lightgrey",
           alpha = .5, # larger alpha is darker rectangle
           ymin = -Inf, ymax = Inf) +
  ggtitle("Year over year growth in state government tax revenue for the United States",
          subtitle="Recession periods are shaded.") +
  labs(caption="Source: U.S. Bureau of the Census") +
  theme_bw() +
  theme(plot.caption = element_text(hjust=0, size=9))
p1

```



```{r vplot, eval=FALSE}
# summary(sgtax.a)

irecs <- which(recessions$rec_year >= 1960)

p <- sgt1 %>%
  filter(year >= 1960) %>%
  ggplot(aes(year, sd10)) +
  geom_line(colour="blue") +
  geom_point(colour="blue", size=0.75) +
  scale_y_continuous(name="Standard deviation (%)", breaks=seq(0, 10, .5), limits=c(0, NA)) +
  scale_x_continuous(name=NULL, breaks=seq(1960, 2030, 5)) +
  annotate("rect",
           xmin = recessions$peak[irecs] %>% decimal_date(),
           xmax = recessions$trough[irecs] %>% decimal_date(),
           fill = "lightgrey",
           alpha = .5, # larger alpha is darker rectangle
           ymin = -Inf, ymax = Inf) +
  ggtitle("Volatility in aggregate state government tax revenue for the United States",
          subtitle="Trailing 10-year standard deviation of annual % change. Recession periods are shaded.") +
  theme_bw()
p    


```



