---
title: "State tax revenue volatility"
author: "Don Boyd"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_notebook:
    df_print: paged
    fig_height: 6
    fig_width: 8
    toc: yes
    toc_depth: 3
    toc_float: true
    number_sections: yes
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include=FALSE)
# renv::record("renv@0.12.2")
```

<!-- comments -->

<!--
# code folding ----
#   alt-L, alt-shift-L  one section
#   alt-O, alt-shift-O  all sections
#   ctrl-D run current code section (remapped from Ctrl-alt-T)


always_allow_html: yes

  html_document:
    toc: yes
    toc_depth: '2'
    df_print: paged
    number_sections: yes



# renv ----
vignette("renv")
documentation online at https://rstudio.github.io/renv

The general workflow when working with renv is:

1. Call renv::init() to initialize a new project-local environment with a private R library,

2. Work in the project as normal, installing and removing new R packages as they are needed in the project,

3. Call renv::snapshot() to save the state of the project library to the lockfile (called renv.lock),

4. Continue working on your project, installing and updating R packages as needed.

5. Call renv::snapshot() again to save the state of your project library if your attempts to update R packages were successful, or call renv::restore() to revert to the previous state as encoded in the lockfile if your attempts to update packages introduced some new problems.


-->


<!-- I put this section after first slide header or else beamer adds a blank slide

"`r format(Sys.time(), '%B %d, %Y')`"
 
-->


```{r renv_notes, eval=FALSE}
# install local packages
# pkgdir <- r"(C:\RPrograms PC\ProjectsCurrent\Pew_Revenue_Volatility\renv\local\)"
# pkgs <- c("bdata_0.6.tar.gz", "BEAData_0.6.0.tar.gz", "btools_0.9.tar.gz")
# install.packages(paste0(pkgdir, pkgs[3]), repos=NULL, type="source")

# to get it all started:
# renv::init()

# to remove it all:
# renv::deactivate()
# then delete the renv folder

# periodically execute:
# renv::snapshot()

# renv:::renv_paths_local()

# ${RENV_PATHS_LOCAL}/<package>_<version>.tar.gz

```



```{r libs_fns, include=FALSE}
source(here::here("r", "libs_base.r"))
# source(here::here("r", "libs_ts.r"))
source(here::here("r", "functions.r"))
devtools::session_info()

```


```{r build_data}
glimpse(sgtax.a)
# df <- sgtax.a
# count(df, year) %>% as.data.frame()
# count(df, vname)

recessions

sgt1 <- sgtax.a %>%
  mutate(year=as.integer(year)) %>%
  filter(stabbr=="US", item=="C105", year>=1950) %>%
  mutate(vpch=pchya(value, year),
         sd10=rollsd_p(vpch, 10),
         sd10c=rollsd_p2(vpch, 10)) # population sd
```




# Goals and issues

* Ways to measure tax revenue volatility

* Recent trends in revenue volatility
    +  Has tax revenue become more volatile over time?
    +  If so, why?

* Drivers of state revenue volatility - to what extent driven by
    +  The nature of national economic changes (e.g., characteristics of recession)?
    +  State economic structure?
    +  State tax structure? 
         +  The portfolio of taxes? 
         +  The structure of individual taxes?

*	Implications for states in current recession and beyond
    +  What does the Covid-19 recession imply for state tax volatility?
    +  How much can states reduce volatility with alternative tax structures?


# Ways to measure tax revenue volatility


# Recent trends in revenue volatility


# Drivers of state revenue volatility

## The nature of national economic changes

## State economic structure

## State tax structure

### The portfolio of state taxes

### The structure of individual taxes


#	Implications for states in current recession and beyond

## Special characteristics of the Covid-19 recession and state tax volatility

## How much can states reduce volatility with alternative tax structures?




# Appendix


## Project description

### Overall goals

Examine tax revenue volatility and its impacts on state governments

Provide deep and broad understanding of:

* Ways to measure tax revenue volatility
* Recent trends in revenue volatility
* Drivers of state revenue volatility
*	Implications for states in current recession and beyond

Ultimate goal: Insights that can help policymakers manage budgets


### Important sub-questions

* To what extent is volatility driven by:
    + The nature of national economic changes (e.g., characteristics of recession)?
    + State economic structure?
    + State tax structure? The portfolio of taxes? The structure of individual taxes?
* Has tax revenue become more volatile over time? If so, why?
* How much can states reduce volatility with alternative tax structures?

All have implications for states in current recession and beyond.


### Overall plan

1. Review relevant literature
2. Build database for analyzing volatility
    + Annual state tax revenue, all states, by tax
    + Auxiliary data to allow population & inflation adjustment
    + Selected other data
3. Multiple methods
    + Descriptive analytic review of history
    + Decomposition
    + Simulations of portfolio effects
    + Exploratory econometrics for certain topics


### Core data source

* Census Bureau Annual Survey of State Finances
  + State tax revenue, annual, all states, many taxes
  + Comprehensive (not just general fund)
  + Relatively consistent definitions over time and states
  + Annual period consistent with policymaking cycle
  + Includes multiple recessions (some data back to mid-1950s):
      + 1980-81, 1990, 2001, and 2007 (2020 won't be available)
      + may be able to include 1960, 1969, 1973 recessions

### Selected technical & data issues

* Tax data implicitly include embedded policy changes. Use Pew tax-change data to help answer: How much does adjusting for policy changes affect conclusions?
* How much do alternative volatility measures affect conclusions?
* Data on structure of individual taxes are limited. To extent practical, use these limited data to help answer: How does structure of individual taxes affect volatility?
  

### Topics and approaches
+----------------------------------+----------------------------------------------+
| Topic                            | Methods                                      |
+==================================+==============================================+
| Ways to measure                  | - Literature review                          |
| revenue volatility               | - Conceptual differences among measures      |
|                                  | - Trends with different measures             |
+----------------------------------+----------------------------------------------+
| Recent revenue volatility trends | - Descriptive analysis                       |
+----------------------------------+----------------------------------------------+
| Drivers of revenue volatility    | - Descriptive analysis                       |
|   (Nature of economic change,    | - Decomposition                              |
|    State economic structure,     | - Econometric analysis                       |
|    State tax structure)          | - Simulation of alternative structures       |
+----------------------------------+----------------------------------------------+
| Implications in current          | - Synthesis of above                         |
| recession and beyond             |                                              |
+----------------------------------+----------------------------------------------+


### Deadlines and schedule

* July 3, 2020 Project start
* October 3, 2020 Provide list of external reviewers
* June 3, 2021  Draft report due (plan on sooner)
* August 3, 2020 Project end and final report due


### Comments from the August 4, 2020 methods review

* Clarify key terms with thorough definitions and describe the applied metrics, highlighting areas of standardization. Absent those details, some confusion may arise. Specifically,
    +	Convey the practical differences between the proposed metrics of revenue volatility for state and other policymakers. These metrics can vary considerably depending on the underlying data distribution, meaningfully changing what is observed as volatility. 
    + Elaborate on how revenue streams may need to be combined/collapsed for analytical tractability. If doing so, highlight the thresholds for those modification, if these differ by state or source, and if those decisions affect the results.  
    + Account for the possible interpretations of concepts like “state economic structures” and economies “at risk”, and how those choices shift the meaning of the analysis and findings. If these concepts are applied differently by state, explain how these may be standardized.  

* Ensure that the selected time frames (periods and trends) and their decision rules are transparent in the final report. Specifically: 
    + Discuss how partial recessionary periods (i.e. quarters) will be addressed since the study will generally leverage annual data, to speak to policymaker budgeting cycles/priorities.
    +	Clarify that “recent trends” may reflect multi-decade periods of fiscal management, which may not be self-evident. Explain how the bounds of “trend periods” are chosen, whether through an assessment of business cycles, regression analyses, inductive analyses or other techniques.

* Ensure consistency between this and prior Pew studies, or provide a crosswalk, particularly if the methods have been modified or augmented. Focus on what sources of uncertainty remain embedded in each measure of volatility. 
    + Assess whether correcting for tax policy adjustments changes estimated volatility. Also, since the policy adjusted time series is shorter, estimate whether a shorter series will complicate the development of robust volatility estimates using the more technical/complex approaches. 
    + Highlight — for more technical methods — how trends are being estimated, and how differences in those trends might change the interpretation of volatility, as the embedded sources of uncertainty differ (e.g. by extracting cyclicality). 



```{r pchplot, fig.height=4.5, fig.width=8, eval=FALSE}
# summary(sgtax.a)
irecs <- which(recessions$rec_year >= 1960)

p1 <- sgt1 %>%
  filter(year >= 1960) %>%
  ggplot(aes(year, vpch)) +
  geom_line(colour="blue", size=1) +
  geom_point(colour="blue", size=1.15) +
  scale_y_continuous(name="% change", breaks=seq(-20, 20, 2)) +
  scale_x_continuous(name=NULL, breaks=seq(1960, 2030, 5)) +
  geom_hline(yintercept = 0) +
  annotate("rect",
           xmin = recessions$peak[irecs] %>% decimal_date(),
           xmax = recessions$trough[irecs] %>% decimal_date(),
           fill = "lightgrey",
           alpha = .5, # larger alpha is darker rectangle
           ymin = -Inf, ymax = Inf) +
  ggtitle("Year over year growth in state government tax revenue for the United States",
          subtitle="Recession periods are shaded.") +
  labs(caption="Source: U.S. Bureau of the Census") +
  theme_bw() +
  theme(plot.caption = element_text(hjust=0, size=9))
p1

```



```{r vplot, eval=FALSE}
# summary(sgtax.a)

irecs <- which(recessions$rec_year >= 1960)

p <- sgt1 %>%
  filter(year >= 1960) %>%
  ggplot(aes(year, sd10)) +
  geom_line(colour="blue") +
  geom_point(colour="blue", size=0.75) +
  scale_y_continuous(name="Standard deviation (%)", breaks=seq(0, 10, .5), limits=c(0, NA)) +
  scale_x_continuous(name=NULL, breaks=seq(1960, 2030, 5)) +
  annotate("rect",
           xmin = recessions$peak[irecs] %>% decimal_date(),
           xmax = recessions$trough[irecs] %>% decimal_date(),
           fill = "lightgrey",
           alpha = .5, # larger alpha is darker rectangle
           ymin = -Inf, ymax = Inf) +
  ggtitle("Volatility in aggregate state government tax revenue for the United States",
          subtitle="Trailing 10-year standard deviation of annual % change. Recession periods are shaded.") +
  theme_bw()
p    


```



