---
title: "State tax revenue volatility"
author: "Don Boyd"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_notebook: 
    df_print: paged
    fig_height: 6
    fig_width: 8
    toc: yes
    toc_float: yes
    number_sections: yes
  html_document:
    toc: yes
    toc_depth: '3'
    df_print: paged
editor_options:
  chunk_output_type: console
---

<!-- alt-O collapse all folds -->

<!-- alt-shift-O expand all folds -->

<!-- <style> -->

<!-- body .main-container { -->

<!--   max-width: 1500px; -->

<!-- } -->

<!-- </style> -->

<!-- Google drive -->

<!-- https://drive.google.com/drive/folders/1fM7HIP7qrKbIdndF_wQ30WxVb1j_Cq06 -->

# PROJECT MANAGEMENT

```{r runall, eval=FALSE, echo=FALSE}
# When we want a final report, run the following code selectively "by hand" (interactively)
# -- NEVER using Knit with eval=TRUE
# note <br> breaks line for html output but \n should break for pdf; also could try |

rmdfn <- "./Pew_Tax_Revenue_Volatility.rmd" # this file
outfn <- paste0("tax_volatility_", format(Sys.time(), "%Y-%m-%d"), ".html")
rmarkdown::render(rmdfn, output_format="html_document", output_file=outfn)

# library("RCurl")
# ftpUpload(outfn, "ftp://kffrig:boyd4812@files.000webhost.com/public_html/KFF_RIG_MedicaidCuts_new.html")

# Note may be safest to fully exit RStudio and restart it before running whole thing. Otherwise knitr may get confused
# and include repetitive information in the output html file.

```

# Links

-   [Google drive folder](https://drive.google.com/drive/folders/1fM7HIP7qrKbIdndF_wQ30WxVb1j_Cq06)
-   [Zotero folder](https://www.zotero.org/groups/2541602/pew_tax_revenue_volatility/library)
-   [Google sheet](https://docs.google.com/spreadsheets/d/1hSWy_7ep4RbM9lfSA-LYnDeGI1c6p9Ud0GqcSlR9xJ8/edit?usp=sharing) (same as link below)

# LOAD OPTIONS, LIBRARIES, FUNCTIONS, CONSTANTS

This section doesn't produce any output other than some informational messages. It also contains some workflow notes as markdown comments.

## Workflow notes

```{=html}
<!--
Workflow notes:

# code folding:
#   alt-L, alt-shift-L  one section
#   alt-O, alt-shift-O  all sections
#   ctrl-D run current code section (remapped from Ctrl-alt-T)

# yaml notes:
always_allow_html: yes

  html_document:
    toc: yes
    toc_depth: '2'
    df_print: paged
    number_sections: yes

# renv workflow:
# to get it all started:
# renv::init()

# to remove it all:
# renv::deactivate()
# then delete the renv folder

# periodically execute:
# renv::status()
# renv::snapshot()

# to see what directory is being used by renv for local package sources, execute:
# renv:::renv_paths_local()


# git workflow:
# when starting work from a different computer
# - pull project from remote
# git checkout -b issue-5  # create branch and checkout branch
# Stage local changes, commit, check status:
# git add foo.txt
# git commit --message "A commit message"
# git status

# now merge
# git checkout master
# git merge issue-5

# then push
# git push

-->
```
## Preliminaries

```{r setup, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(include=FALSE, eval=TRUE,
                      echo=FALSE, message = FALSE, warning = FALSE)

# opts_chunk is a merger of local options and default options
# opts_current is the current chunk's options - from within any given chunk
# knitr::opts_chunk$get()  # copy to console to get proper list of options
# knitr::opts_current$get()
str(knitr::opts_chunk$get()) # for a list of default chunk options.

```

```{r css, echo=FALSE}
# blockquote {
#     padding: 20px;
#     margin: 0 0 20px;
#     font-size: 14px;
#     border-left: 5px solid #eee;
# }

```

```{r renv_notes, eval=FALSE}

# vignette("renv")
# documentation online at https://rstudio.github.io/renv
# 
# The general workflow when working with renv is:
# 
# 1. Call renv::init() to initialize a new project-local environment with a private R library,
# 
# 2. Work in the project as normal, installing and removing new R packages as they are needed in the project; periodically do renv::status()
# 
# 3. Call renv::snapshot() to save the state of the project library to the lockfile (called renv.lock),
# 
# 4. Continue working on your project, installing and updating R packages as needed.
# 
# 5. Call renv::snapshot() again to save the state of your project library if your attempts to update R packages were successful, or call renv::restore() to revert to the previous state as encoded in the lockfile if your attempts to update packages introduced some new problems.

# https://rstudio.github.io/renv/articles/renv.html

# Custom R Package Repositories
# Custom and local R package repositories are supported as well. The only requirement is that these repositories
# are set as part of the repos R option, and that these repositories are named. For example, you might use:
#
# repos <- c(CRAN = "https://cloud.r-project.org", WORK = "https://work.example.org")
# options(repos = repos)
# to tell renv to work with both the official CRAN package repository, as well as a package repository you have hosted and set up in your work environment.

# 1) Define directories

# pkgsource_dir - not needed by Pew: location of source files for packages I created
# not in this project, but avail on github if Pew is interested
pkgsource_dir <- r"(C:\Users\donbo\Documents\R_projects\packages)"

# where Boyd puts tar.gz files (tarballs) for packages created by Boyd - in this project
# Pew will install the packages from here
pkgdir <- r"(C:\Users\donbo\Documents\R_projects\Pew_Revenue_Volatility\renv\local\)"

# 2) This step is for Boyd only:
# uncomment code below, build Boyd packages, save tarballs in the renv local directory
devtools::build(pkg = file.path(pkgsource_dir, "BEAData"),  # bdata BEAData BLSdata btools qtax
                path = pkgdir,
                binary = FALSE,
                vignettes = FALSE,
                manual = FALSE,
                args = NULL,
                quiet = FALSE)
# do the above for any other needed local packages

# 3) For Boyd and Pew: uncomment code and install tarballs
# Pew may have to define the pkgdir location on your machine
# if the package is in use and you are updating it, then first detach it
#  example: detach(package:bdata)

# renv:::renv_paths_local()
# pkgdir <- here::here("renv", "local", "/")  # can define package

# Now, install each package, being sure to use full tarball name based on files in the
# pkgdir directory (tarball names may be updated from those shown below)
# renv::install(file.path(pkgdir, "bdata_0.9.tar.gz"))
# renv::install(file.path(pkgdir, "BEAData_0.7.4.tar.gz"))
# renv::install(file.path(pkgdir, "BLSdata_0.3.1.tar.gz"))
# renv::install(file.path(pkgdir, "btools_0.9.1.tar.gz"))
# renv::install(file.path(pkgdir, "qtax_0.3.0.tar.gz"))


# key renv commands
# to get it all started:
# renv::init()

# renv::status()

# periodically update the renv lockfile to save current versions of packages
# renv::snapshot() 

# to remove it all:
# renv::deactivate()
# then delete the renv folder

# to see what directory is being used by renv for local package sources, execute:
# renv:::renv_paths_local()

# ${RENV_PATHS_LOCAL}/<package>_<version>.tar.gz

```

# Libraries and functions

```{r get_libs_and_functions, eval=TRUE}
# clear objects from the workspace
rm(list=ls(all=TRUE))
source(here::here("r", "libs_base.r"))
source(here::here("r", "libs_ts.r"))
source(here::here("r", "functions.r"))
source(here::here("r", "functions_3panel.r"))
devtools::session_info()
# devtools::package_info()
# knitr::opts_current$get()

```

```{r functions, eval=TRUE}

is.even <- function(x) {(x %% 2) == 0}

savefig <- function(basename, .p=p, .pdata=pdata, width=8, height=6, scale=1) {
  
  ggsave(filename = here::here("results", 
                               paste0(basename, ".png")),
         plot = .p, 
         width=width, height=height,
         scale=scale)
         
  write_csv(.pdata, file= here::here("results", paste0(basename, "_data.csv")))
}

savetab <- function(basename, .tab=tab, .tabdata=tabdata, zoom=2, expand=5) {
  # zoom and expand are passed to the underlying webshot::webshot() function
  # for PNG saving zoom defaults to a scale level of 2 and
  # expand (adds whitespace pixels around cropped table image) has a default value of 5)
  
  # expand: A numeric vector specifying how many pixels to expand the clipping
  # rectangle by. If one number, the rectangle will be expanded by that many
  # pixels on all sides. If four numbers, they specify the top, right, bottom,
  # and left, in that order. When taking screenshots of multiple URLs, this
  # parameter can also be a list with same length as url with each element of
  # the list containing a single number or four numbers to use for the
  # corresponding URL.
  
  # zoom: A number specifying the zoom factor. A zoom factor of 2 will result in
  # twice as many pixels vertically and horizontally. Note that using 2 is not
  # exactly the same as taking a screenshot on a HiDPI (Retina) device: it is
  # like increasing the zoom to 200 doubling the height and width of the browser
  # window. This differs from using a HiDPI device because some web pages load
  # different, higher-resolution images when they know they will be displayed on
  # a HiDPI device (but using zoom will not report that there is a HiDPI
  # device).
  
  gtsave(.tab, here::here("results", paste0(basename, ".png")),
         zoom = zoom, expand = expand)
         
  write_csv(.tabdata, file= here::here("results", paste0(basename, "_data.csv")))
}

corr_tidy <- function(vars){
  # return a tidy data frame with correlations
  mat <- as.matrix((vars))
  Hmisc::rcorr(mat) %>%
    tidy
}

dl <- function(vec) {
  # difference in log values
  # ASSUME data are sorted by year and there are no gaps and no zeros
  lvec = log(vec)
  laglvec = c(NA, head(lvec, -1))
  lvec - laglvec
}

episodes <- function(x, f="mean"){
  # create measures of how long and severe episodes are
  # x: a vector of deviations from trend, in date order
  # returns: list(len, asum) where
  #   len is the mean length of episodes that are above or below trend
  #   asum is the mean of the sums, across episodes, of absolute deviations from trend
  
  if(f=="mean") {
    f <- mean
  } else if(f=="median"){
    f <- median
  } else print("ERROR!!!")
  
  
  df <- tibble(x) %>%
    mutate(group = data.table::rleid(sign(x))) %>%
    group_by(group) %>%
    summarise(len=n(), asum=sum(abs(x), na.rm=TRUE),
              .groups="drop") %>%
    summarise(len=f(len), asum=f(asum), .groups="drop")
  
  list(elength=df$len, esum=df$asum)
}


hpdf <- function(data, colname="value") {
  # when type="lambda" freq is the lambda parameter for the HP filter
  # 6.25 is commonly recommended for annual data; see:
  # Ravn, Morten O., and Harald Uhlig. “On Adjusting the Hodrick-Prescott Filter for the Frequency of Observations.” Review of Economics and Statistics 84, no. 2 (May 2002): 371–76. https://doi.org/10.1162/003465302317411604.
  mod <- hpfilter(data[colname], freq=6.25, type=c("lambda"), drift=FALSE)
  tibble(trend=as.numeric(mod$trend),
         cycle=as.numeric(mod$cycle))
}


hptrend <- function(vec, smooth=NULL){
  # Hodrick-Prescott (HP) filter
  # vec: vector of time series data, in order, no missing values
  # returns: trend vector
  # when type="lambda" freq is the lambda parameter for the HP filter
  # 6.25 is commonly recommended for annual data; see:
  # Ravn, Morten O., and Harald Uhlig. “On Adjusting the Hodrick-Prescott Filter for the Frequency of Observations.” Review of Economics and Statistics 84, no. 2 (May 2002): 371–76. https://doi.org/10.1162/003465302317411604.
  vts_hp <- hpfilter(ts(vec), freq=smooth, type=c("lambda"), drift=FALSE)
  as.numeric(vts_hp$trend)
}

check <- vbase %>%
  filter(stabbr=="IN", name=="tottax", realnom=="nominal")

hptrend2 <- function(vec, smooth=NULL){
  # Hodrick-Prescott (HP) filter
  # vec: vector of time series data, in order, no missing values
  # returns: trend vector
  # when type="lambda" freq is the lambda parameter for the HP filter
  # 6.25 is commonly recommended for annual data; see:
  # Ravn, Morten O., and Harald Uhlig. “On Adjusting the Hodrick-Prescott Filter for the Frequency of Observations.” Review of Economics and Statistics 84, no. 2 (May 2002): 371–76. https://doi.org/10.1162/003465302317411604.
  # if statements address percent change, which won't be avail for item 1
  if(is.na(vec[1])){
    vts <- ts(vec[-1])
  } else vts <- ts(vec)
  vts_hp <- hpfilter(vts, freq=smooth, type=c("lambda"), drift=FALSE)
  trend <- as.numeric(vts_hp$trend)
  if(is.na(vec[1])){
    trend <- c(NA, trend)
  } 
  trend
}


sre <- function(data) {
  # short-run elasticity relative to national gdp
  mod = lm(dlvalue ~ dlgdp, data=data)
  list(mod=mod, coeff=unname(mod$coefficients["dlgdp"]))
}


sre_gsp <- function(data) {
  # short-run elasticity with both national and state gdp log differences on rhs
  mod = lm(dlvalue ~ dlgdp + dlgsp, data=data)
  list(mod=mod, coeff=unname(mod$coefficients["dlgdp"]))
}


rollsre <- function(y, x, nobs) {
  # rolling short run elasticity
  # print("in rollsre")
  sremod <- function(df){
    # print("in sremod")
    mod <- lm(y ~ x, na.action=na.exclude, data=df)
    sre <- coef(summary(mod))[2, "Estimate"]
    sre
  }
  safe_sremod <- safely(sremod)
  # print("before df")
  df <- data.frame(y=y, x=x)
  # print("after df")
  zoo::rollapply(df, nobs, function(df) sremod(as.data.frame(df)),
                 by.column = FALSE,
                 fill=NA, align="right")
}


get_measures <- function(df){
  # df data frame with:
  #     stabbr
  #     name
  #     realnom
  #     year
  #     value
  #     trend
  #     pch
  #     pdtrend
  #   assumed to already be winnowed down to desired states, names, years
  
  vol_basic <- df %>%
    group_by(stabbr, realnom, name) %>%
    summarise(n=n(),
              pchsd = sd(pch, na.rm=TRUE),
              hpsd = sd(pdtrend, na.rm=TRUE),
              .groups = "drop")
  # vol_basic
  
  # now construct the more complicated measures
  vol_other <- df %>%
    # bring in real gdp or nominal gdp from vbase, which must exist
    left_join(vbase %>% 
                filter(name=="gdp") %>% 
                select(stabbr, realnom, year, gdp=value),
              by=c("stabbr", "realnom", "year")) %>%
    group_by(stabbr, name, realnom) %>%
    mutate(dlvalue=dl(value), dlgdp=dl(gdp)) %>%
    nest() %>%
    # note that this does not filter out the case where GDP is on the lhs and
    # the rhs so we will get warnings for that! they will arise in the next step
    # when we unpack
    mutate(sremod = purrr::map(data, 
                               safely(function(df) lm(dlvalue ~ dlgdp, data=df))),
           epi = purrr::map(data, 
                            safely(function(df) episodes(df$pdtrend)))) %>%
    ungroup
  # vol_other
  # vol_basic
  
  # sre -- short-run elasticity -- unpack into sre and sresum unpack the summary
  # measures from sremod to get coefficient (short-run elasticity, sre) and its
  # standard error
  sredf <- vol_other %>%
    select(stabbr, name, realnom, data, sremod) %>% 
    # we want to keep data from one set
    unnest_wider(col=sremod) %>%  # we need the column "result"
    mutate(tidied=purrr::map(result, tidy)) %>%
    unnest(cols=tidied) %>%
    filter(term=="dlgdp") %>%
    select(stabbr, name, realnom, data, sre=estimate, srese=std.error)
  # the warnings are for gdp on lhs and rhs

  # epi -- episodes -- unpack into elength and esum
  epidf <- vol_other %>%
    select(stabbr, name, realnom, epi) %>%
    unnest_wider(col=epi) %>%
    unnest_wider(col=result)

# unpack the summary measures from sregspmod to get coefficient (short-run
# elasticity, sre) and its standard error
  measures <- vol_basic %>%
    left_join(sredf, by = c("stabbr", "name", "realnom")) %>%
    left_join(epidf, by = c("stabbr", "name", "realnom")) %>%
    select(stabbr, name, realnom, data, everything())
  measures
}

```

## Constants

```{r constants, eval=TRUE}
# Constants, themes, etc.

#.. constants ----
# vorder, taxnames
stdc <- c(state.abb, "DC")


#.. colors ----
# https://colorbrewer2.org/
bluered <- c("blue", "#a50f15")  # #a50f15 darkred
bluegreen <- c("blue", "darkgreen")
# color2 <- c("blue", "red")
# color3 <- c("blue", "red", "darkgreen")
# color4 <- display.brewer.pal(4,"Set3")
# colors <- list()
# colors[[1]] <- "blue"
# colors[[2]] <- color2
# colors[[3]] <- color3
# colors[[4]] <- color4
# colors

# examine colors with the following functions
# RColorBrewer colors:
# brewer.pal.info
# brewer.pal.info[brewer.pal.info$category == "seq",]
# display.brewer.pal(7,"Accent")
# display.brewer.pal(3,"Dark2")
# display.brewer.pal(4,"Dark2")
# display.brewer.pal(4,"Set1")
# display.brewer.pal(4,"Set2")
# display.brewer.pal(4,"Set3")

# scale_color_viridis()


#.. graph theme items ----
legend_none <- theme(legend.position = "None")
legend_notitle <- theme(legend.title = element_blank())
caption_left <- theme(plot.caption = element_text(hjust = 0))

theme_report <- theme_bw() + 
  theme(
    plot.caption = element_text(hjust = 0)
    )

theme_map_boyd <- function(base_size=9, base_family="") {
  # https://socviz.co/maps.html
  # https://github.com/kjhealy/socviz
  require(grid)
  theme_bw(base_size=base_size, base_family=base_family) %+replace%
    theme(axis.line=element_blank(),
          axis.text=element_blank(),
          axis.ticks=element_blank(),
          axis.title=element_blank(),
          panel.background=element_blank(),
          panel.border=element_blank(),
          panel.grid=element_blank(),
          panel.spacing=unit(0, "lines"),
          plot.background=element_blank(),
          legend.justification = c(0,0) # ,
          # legend.position = c(0,0)
    )
}


#.. miscellaneous constants ---
mlabels <- tribble(
  ~measure, ~mlabel,
  "pchsd", "Standard deviation of annual % change",
  "hpsd", "Standard deviation of % difference from trend",
  "sre", "Short-run elasticity vs. state GDP",
  "esum", "Average cumulative % difference from trend, per episode above or below trend",
  "srese", "Standard error short-run elasticity",
  "elength", "Average length of deviation-from-trend episodes"
)

taxnames <- tribble(
  ~name, ~namef,
  "tottax", "Total taxes",
  "iit", "Personal income tax",
  "gst", "General sales tax",
  "selsalestax", "Selective sales taxes",
  "cit", "Corporate income tax",
  "sevtax", "Severance taxes",
  "othertax", "Other taxes"
)

#.. source notes ----
source_acs <- "American Community Survey 5-year summary file"
source_bea <- "U.S. Bureau of Economic Analysis"
source_cbo <- "U.S. Congressional Budget Office"
source_decennialpop <- "U.S. Census Bureau Decennial Census"
source_soi <- "IRS Statistics of Income"

```

# LOAD AND PREPARE DATA

## Load previously created data

```{r get-data, include=FALSE, eval=TRUE}
load(file = here::here("data", "general.RData"), verbose=TRUE)
load(file = here::here("data", "econ_national.RData"), verbose=TRUE)
load(file = here::here("data", "gdp_state.RData"), verbose=TRUE)
load(file = here::here("data", "qcew_state.RData"), verbose=TRUE)
load(file = here::here("data", "taxdata.RData"), verbose=TRUE)
load(file = here::here("data", "bmaps.RData"), verbose=TRUE)

```

## Enhance data

```{r tax_shares}
# slim data files down if desired
# pewtax  <- pewtax %>%
#   select(stabbr, year, taxtype, vartype, pch, level)

# get tax shares
taxshares <- censustax %>%
  pivot_wider(values_fill=0) %>%
  mutate(across(-c(stabbr, year, tottax), ~ .x / tottax))

# do a quick check to make sure the shares add approximately to 1
# sharesums <- taxshares %>%
#   select(-tottax) %>%
#   pivot_longer(-c(stabbr, year)) %>%
#   group_by(stabbr, year) %>%
#   summarise(sum=sum(value), .groups="drop")
# all.equal(sharesums$sum, rep(1, nrow(sharesums)))  # checks whether the sums==1, within a tolerance
# quantile(sharesums$sum)
# sharesums %>% filter(sum < .99)  # we have 19 that are < .99; CO 2019 is the only concerning one
# rm(sharesums)

```

## Stack the data

-   Stack files so that we can construct sdpch and hppch volatility measures for any item.

-   We need stabbr, name, year, value for all files

-   Then bring in any rhs vars we need

```{r stack_prep_us, eval=TRUE}
# gdpfy gdppi
gdpfy_prep <- gdpfy %>%
  mutate(stabbr="US", name="gdp", realnom="nominal") %>%
  select(stabbr, name, realnom, year, value=gdp)

gdppi_prep <- gdppi %>%
  mutate(stabbr="US", name="gdppi", realnom="price") %>%
  select(stabbr, name, realnom, year, value=gdppi)


cg_prep <- capgains %>%
  mutate(stabbr="US", name="capgains", realnom="nominal") %>%
  select(stabbr, name, realnom, year, value=capgains)
glimpse(cg_prep)

summary(cg_prep) # 1954-2031

```

```{r stack_prep_econstates, eval=TRUE}

# sgdpfy   count(sgdpfy, name)
# sgdpgrouped_sfy
# qcew_grouped

sgdp_prep <- sgdpfy %>%  # has gdp, rgdp
  # indicate if data were real in original source
  mutate(name=ifelse(name=="rgdp", "rgdp_source", name),  
         realnom=case_when(name=="gdp" ~ "nominal",
                           name=="rgdp_source" ~ "real",
                           TRUE ~ "ERROR")) %>%
  select(stabbr, name, realnom, year, value)
summary(sgdp_prep)  # 1964-2020
count(sgdp_prep, name)
count(sgdp_prep, realnom)
count(sgdp_prep, stabbr) # 50 states

gspsectors_prep <- sgdpgrouped_sfy %>%  # 2005+
  select(-c(total, sum)) %>%
  pivot_longer(-c(stabbr, year)) %>%
  mutate(name=paste0("sgdp_", name),
         realnom="nominal") %>%
  select(stabbr, name, realnom, year, value)
summary(gspsectors_prep)
count(gspsectors_prep, stabbr)  # 50 states, DC, US
count(gspsectors_prep, name)

# qcew_prep <- qcew_grouped %>%
#   mutate(realnom=ifelse(name=="emp", "real", "nominal")) %>%
#   pivot_longer(-c(stabbr, year, name, realnom), names_to = "sector") %>%
#   unite(name, c(name, sector))
# glimpse(qcew_prep)
# summary(qcew_prep)  # 1990 2020
# count(qcew_prep, name, realnom)

```

```{r stack_prep_taxstates}
tax_prep <- bind_rows(censustax, census_gstiitadj) %>%
  # we do not have gstadj data for the period excluded, so it is safe to do this
  filter(!(stabbr == "DE" & name == "gst")) %>% # only 3 obs, 1951-53
  # negative in 2014, maybe later figure out how to salvage this
  filter(!(stabbr == "OH" & name == "cit")) %>% 
  mutate(realnom="nominal")
glimpse(tax_prep)
summary(tax_prep)  # 1951-2020

# gstbase, created in prep_data, maps BEA consumption components to state
# sales tax bases
gst_prep <- gstbase %>%
  filter(stabbr != "DC") %>%
  mutate(realnom="nominal") %>%
  select(stabbr, name, realnom, year, value)
glimpse(gst_prep)
summary(gst_prep)  # 1929-2020

```

```{r stack_combined}
stack1 <- bind_rows(
  gdpfy_prep, gdppi_prep, cg_prep,  # US economy
  sgdp_prep, gspsectors_prep, # maybe qcew_slim?? qcew_prep,  # state economies
  tax_prep, gst_prep  # state tax-related variables
  ) %>%
  filter(year %in% 1950:2020)  # gdppi price starts in 1947, 1950 seems like a good start
glimpse(stack1)
summary(stack1)
count(stack1, stabbr)
count(stack1, name)

# create real versions of the nominal variables, 2021 $
gdppi
summary(gdppi)
rstack1 <- stack1 %>%
  filter(realnom=="nominal") %>%
  left_join(gdppi %>% select(year, igdppi),
            by="year") %>%
  mutate(value=value * igdppi,
         realnom="real") %>%
  select(stabbr, name, realnom, year, value)
summary(rstack1)

# combine into a single stacked file just US states
stack <- bind_rows(stack1, rstack1) %>%
  filter(stabbr %in% c("US", state.abb))
glimpse(stack)
summary(stack)
count(stack, realnom)
check <- count(stack, name)

stack %>%
  filter(is.na(value))

stack %>%
  filter(stabbr=="US") %>%
  count(name)

stack %>%
  filter(stabbr=="US", name=="gdp") %>% ht
```

## Create vbase - THIS IS A CRUCIAL DATA FRAME

vbase has:

-   stabbr
-   name (type of data item - econ variables, tax variables)
-   realnom (categorical - nominal, price, real)
-   value
-   pch (vs year ago)
-   trend (computed as hptrend)
-   pdtrend (pct difference from trend)



```{r vbase}
# lambda=6.25 is commonly recommended for annual data
# see OECD World Library and cites within

vbase <- stack %>%
  arrange(stabbr, name, realnom, year) %>% 
  group_by(stabbr, name, realnom) %>%
  mutate(pch=value / value[match(year - 1, year)] - 1,
         pchtrend=hptrend2(pch, smooth=6.25),
         dpchtrend=pch - pchtrend,
         trend=hptrend2(value, smooth=6.25),
         pdtrend=value / trend - 1) %>%
  ungroup
summary(vbase)

```

## Function to compute volatility measures from vbase
```{r}
#.. function to create summary vol measures for a time period and set of variables ----

f_vol <- function(startend, vars, vbase){
  year1 <- startend[1]
  year2 <- startend[2]
  period <- paste0(year1, "-", year2)
  print(period)
  
  # vol1 is our starting point for creating measures
  vol1 <- vbase %>%
    filter(name %in% vars,
         year %in% (year1 - 1):year2) %>% # minus 1 so we can get lagged value
    # bring in nominal gdp
    left_join(vbase %>% 
                filter(name=="gdp") %>%
                select(stabbr, year, realnom, gdp=value),
              by=c("stabbr", "year", "realnom")) %>%
    group_by(stabbr, name, realnom) %>%
    arrange(year) %>%
    mutate(dlvalue=dl(value), dlgdp=dl(gdp)) %>% # first difference of log values
    # get rid of unnecessary year1 - 1 value
    filter(year >= year1) %>%
    nest() %>%
    ungroup
  
  # create model-based measures
  volmods <- vol1 %>%
    # we must use "safely" because some state-tax combinations might be bad -- for example,
    # having a zero value for tax revenue which, when we take the log, became NA
    mutate(sremod = purrr::map(data, safely(function(df) lm(dlvalue ~ dlgdp, data=df))))
  
  # sre -- short-run elasticity -- unpack into sre and srese
  # unpack the summary measures from sremod to get:
  #   coefficient (short-run elasticity, sre), and
  #   its standard error
  # this will generate warnings where we have models that are not proper
  # we will weed them out later
  # this can be diome consuming
  sredf <- volmods %>%
    select(stabbr, name, realnom, sremod) %>%
    mutate(result=purrr::map(sremod, "result"),
           tidied=purrr::map(result, tidy),
           stats=purrr::map(result, glance)) %>%
    # pull specific statistics out of stats
    hoist(stats, nobs="nobs", dfr="df.residual", r2adj="adj.r.squared") %>%
    select(-c(sremod, result)) %>%
    unnest(cols=tidied) %>%
    filter(term=="dlgdp") %>%
    select(stabbr, name, realnom, nobs, dfr, r2adj, sre=estimate, srese=std.error)
  # note - we lost MO sevtax; it has some junky values like 0, 1, etc.
  # count(sredf, dfr)
  
  # get the other volatility measures and combine them with sredf
  volmeas <- vol1 %>%
    # calc the other volatility measures
    select(stabbr, name, realnom, data) %>%
    unnest(cols=data) %>% # data won't be in the output
    group_by(stabbr, name, realnom) %>% 
    summarise(growthmdn=median(pch, na.rm=TRUE),
              pdtrendiqr=IQR(pdtrend, na.rm=TRUE),
              pchiqr=IQR(pch, na.rm=TRUE),
              dpchtrendiqr=IQR(dpchtrend, na.rm=TRUE),
              pchsd=sd(pch, na.rm=TRUE),
            .groups="drop") %>%
    ungroup 
  
  # put the pieces back together:
  #   vol1 -- original data
  #   sredf -- model-based short-run elasticity estimates
  #   volmeas -- other summary measures
  
  vol <- vol1 %>%
    # bring in the model-based volatility measures
    left_join(sredf, by = c("stabbr", "name", "realnom")) %>%
    left_join(volmeas, by = c("stabbr", "name", "realnom")) %>%
    mutate(namef=factor(name,
                        levels=taxnames$name,
                        labels=taxnames$namef),
           year1=!!year1,
           year2=!!year2,
           period=!!period) %>%
    select(stabbr, name, namef, year1, year2, period, data, everything()) %>%
    arrange(namef)
  
  vol
}

```


```{r measures_decades}

# now construct a stacked volatility file with several different periods
count(vbase, name)
vtax <- c("tottax", "iit", "gst", "selsalestax", "cit", "sevtax", "othertax")
vnontax <- c("gdp", "capgains")
vars <- c(vtax, vnontax)

check <- f_vol(c(1995, 2009), vars, vbase)

# get several volatility time periods with different start and end years
# create a list of start-end pairs for which we want volatility estimates
ylist <- list(
  c(1970, 2020),
  c(1960, 1969),
  c(1970, 1979),
  c(1980, 1989),
  c(1990, 1999),
  c(2000, 2009),
  c(2010, 2019)
)
volall <- map_dfr(ylist, f_vol, vars=vars, vbase=vbase)
saveRDS(volall, here::here("data", "volall.rds"))

glimpse(volall)
count(volall, period)
count(volall, stabbr)
count(volall, name, namef)
count(volall, realnom)

volall %>%
  filter(stabbr=="US", name=="tottax", realnom=="nominal")

```


```{r measures_recession_overlaps}
# create a list of start-ends for recession periods
recessions

# create a list of start-end pairs for which we want volatility estimates
recyears <- c(1969, 1973, 1980, 1990, 2001, 2007)

plusminus <- 5
recylist <- purrr::map(recyears, function(y) c(y - plusminus, y + plusminus))
recylist
# create vector of periods that we can map back to recyears
recperiods <- purrr::map_chr(recylist, function(years) paste0(years[1], "-", years[2]))
cbind(recyears, recperiods)

vtax <- c("tottax", "iit", "gst", "selsalestax", "cit", "sevtax", "othertax")
vnontax <- c("gdp", "capgains")
vars <- c(vtax, vnontax)

volrecs <- map_dfr(recylist, f_vol, vars=vars, vbase=vbase)
volrecs <- volrecs %>%
  mutate(recyear=factor(period, levels=recperiods, labels=recyears),
         recyear=as.integer(as.character(recyear)))
glimpse(volrecs)
count(volrecs, recyear, period)  
saveRDS(volrecs, here::here("data", "volrecs.rds"))

```


## Mapping volatility measures
```{r}
#.. function to map vol measures ----
# Sequential color palettes available in colorbrewer
# Blues, BuGn, BuPu, GnBu, Greens, Greys, Oranges, OrRd, PuBu, PuBuGn, PuRd,
# Purples, RdPu, Reds, YlGn, YlGnBu, YlOrBr, YlOrRd


f_volmap <- function(tax="tottax", measure="sre", 
                     gtitle, data=volrecs, facetvar="recyear") {
  
  # titles
  if(measure=="sre"){
    mname="short-run elasticity"
  } else if(measure=="pchiqr"){
    mname="interquartile range of percent change"
  } else if(measure=="dpchtrendiqr"){
    mname="interquartile range of difference in percent change from trend"
  }else if(measure=="pdtrendiqr"){
    mname="interquartile range of percent deviation from trend"
  }
  gsubtitle <- paste0("Volatility measure: ", mname)
  
  # build the data
  voldata <- data %>%
    filter(name==tax, realnom=="nominal") %>%
    select(facet=all_of(facetvar), stabbr, value=all_of(measure))
  
  # cut the measure into quartiles, over all recession years
    
  f_breaks <- function(value){
    # function to define quartile cuts
    qtiles <- quantile(value, p=c(0, .25, .5, .75, 1), na.rm=TRUE)
    # breaks replace min and max with infinite
    c(-Inf, qtiles[-c(1, length(qtiles))], Inf)
  }
  breakpoints <- f_breaks(voldata$value)
  
  # create decent labels for the groups defined by these breakpoints
  f_cutlabs <- function(breakpoints){
    if(measure %in% c("pdtrendiqr", "pchiqr", "dpchtrendiqr")){
      fmt_breaks <- percent(breakpoints[2:4], accuracy=.1)
    } else if(measure %in% c("sre")){
      fmt_breaks <- number(breakpoints[2:4], accuracy=.01)
    }
    lab1 <- paste0("Q1: <= ", fmt_breaks[1])
    lab2 <- paste0("Q2: > ", fmt_breaks[1], " - ", fmt_breaks[2])
    lab3 <- paste0("Q3: > ", fmt_breaks[2], " - ", fmt_breaks[3])
    lab4 <- paste0("Q4: > ", fmt_breaks[3])
    c(lab1, lab2, lab3, lab4)
  }
  vlabs <- f_cutlabs(breakpoints)
  
  # create input data frame with all states plus DC for all periods
  # then cut into groups and label
  mbase <- expand_grid(stabbr=c(state.abb, "DC"),
                       facet=unique(voldata$facet)) %>%
    left_join(voldata, by = c("stabbr", "facet")) %>%
    mutate(vcut=cut(value, breakpoints),
           vlab=factor(as.integer(vcut), levels=1:4, labels = vlabs))
  
  # now construct the mapdata and create the map
  # we need an input data frame that has all states plus DC for all years
  mdata <- states51 %>%
    left_join(mbase, by="stabbr")

  m <- mdata %>%
    ggplot(aes(x = long, y = lat, group = group, fill = vlab)) +
    geom_polygon(color = "black", size = 0.1) +
    coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
    scale_fill_brewer(palette = "BuPu", na.translate = FALSE) + # , na.value="grey90") +
    # scale_fill_manual(breaks=as.integer(1:4), values=c("red", "green", "yellow", "blue")) +
    theme_map_boyd() + 
    theme(legend.position = "right") +
    labs(fill="Quartile: ") +
    # legend_notitle +
    facet_wrap(~facet, ncol = 2) +
    ggtitle(gtitle, subtitle=gsubtitle)
  
   # return the base map data and the map
   mlist <- list(mbase=mbase, m=m)
}

```


```{r map_decades}
volall <- readRDS(here::here("data", "volall.rds"))

voldecades <- volall %>% filter(period != "1970-2020")
count(voldecades, period)

gtitle <- "Tax revenue volatility by decade"

mlist <- f_volmap(tax="tottax", measure="pdtrendiqr", 
                  gtitle=gtitle, data=voldecades, facetvar="period")
savefig(basename="map_decades_pdtrendiqr", .p=mlist$m, .pdata = mlist$mbase)

mlist <- f_volmap(tax="tottax", measure="pchiqr", 
                  gtitle=gtitle, data=voldecades, facetvar="period")
savefig(basename="map_decades_pchiqr", .p=mlist$m, .pdata = mlist$mbase)

mlist <- f_volmap(tax="tottax", measure="sre", 
                  gtitle=gtitle, data=voldecades, facetvar="period")
savefig(basename="map_decades_sre", .p=mlist$m, .pdata = mlist$mbase)

```


```{r map_recessions}
volrecs <- readRDS(here::here("data", "volrecs.rds"))
glimpse(volrecs)
count(volrecs, recyear, period)  

gtitle <- "Tax revenue volatility: 5 years before recession start though 5 years after"
# tax="tottax", measure="sre",                      gtitle, data=volrecs, facetvar="recyear"

mlist <- f_volmap(tax="tottax", measure="pdtrendiqr", 
                  gtitle=gtitle, data=volrecs, facetvar="recyear")
savefig(basename="map_recessions_pdtrendiqr", .p=mlist$m, .pdata = mlist$mbase)

mlist <- f_volmap(tax="tottax", measure="pchiqr", 
                  gtitle=gtitle, data=volrecs, facetvar="recyear")
savefig(basename="map_recessions_pchiqr", .p=mlist$m, .pdata = mlist$mbase)

mlist <- f_volmap(tax="tottax", measure="dpchtrendiqr", 
                  gtitle=gtitle, data=volrecs, facetvar="recyear")
savefig(basename="map_recessions_dpchtrendiqr", .p=mlist$m, .pdata = mlist$mbase)

mlist <- f_volmap(tax="tottax", measure="sre", 
                  gtitle=gtitle, data=volrecs, facetvar="recyear")
savefig(basename="map_recessions_sre", .p=mlist$m, .pdata = mlist$mbase)

```

```{r tab_rec_features}
rec_features

tab <- rec_features %>%
  gt() %>%
  tab_header(
    title = "Selected characteristics of the last 8 recessions"
    ) %>%
  cols_label(recyear=html("Initial year of<br>the recession"),
             duration=html("Duration<br>(months)"),
             rgdp=html("Real GDP"),             
             rpce=html("Real personal consumption"),
             urchange=html("Unemployment rate<br>% point change"),
             cgpch=html("% change in<br>capital gains")) %>%
  tab_spanner(
    label = "% change, recession peak to trough:",
    columns = c(rgdp, rpce)
    ) %>%
  fmt_percent(
    columns = c(rgdp, rpce, urchange, cgpch),
    decimals = 1) %>%
  fmt_number(
    columns = c(duration),
    decimals = 0) %>%
  tab_footnote(
    footnote = "Highest quarterly value within 1 year of start minus lowest value within 2 years of trough.",
    locations = cells_column_labels(columns = urchange)
    ) %>%
  tab_footnote(
    footnote = "% change from highest annual value within two years of start to lowest value in 2 years after start.",
    locations = cells_column_labels(columns = cgpch)
    ) %>%
  tab_style(
    style = list(
      cell_fill(color = "grey95")
      ),
    locations = cells_body(
      rows=is.even(1:nrow(rec_features))
      )
    )
tab

savetab("tab_rec_features", .tabdata=rec_features)


```



# DATA EXPLORATION

```{r}
volall <- readRDS(here::here("data", "volall.rds"))

```

## make some maps
```{r}
glimpse(volall)
pdata1 <- volall %>%
  filter(period=="1970-2020", realnom=="nominal", stabbr %in% stdc, name=="tottax") %>%
  select(stabbr, pchiqr)
quantile(pdata1$pchiqr)

pdata2 <- pdata1 %>%
  mutate(vcut=cut(pchiqr, 
                  breaks=c(-Inf, .05, .06, .07, .1, Inf)))
count(pdata2, vcut)

mdata <- states51 %>%
  left_join(pdata2, by="stabbr")


m <- mdata %>%
  ggplot(aes(x = long, y = lat, group = group, fill = vcut)) +
  geom_polygon(color = "gray90", size = 0.1) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  scale_fill_brewer(palette = "PuBu", na.translate = FALSE) +
  theme_map()
m + 
  theme(legend.position = "right") +
  legend_notitle

theme_map <- function(base_size=9, base_family="") {
  # see:
  # https://socviz.co/maps.html
  # https://github.com/kjhealy/socviz
  require(grid)
  theme_bw(base_size=base_size, base_family=base_family) %+replace%
    theme(axis.line=element_blank(),
          axis.text=element_blank(),
          axis.ticks=element_blank(),
          axis.title=element_blank(),
          panel.background=element_blank(),
          panel.border=element_blank(),
          panel.grid=element_blank(),
          panel.spacing=unit(0, "lines"),
          plot.background=element_blank(),
          legend.justification = c(0,0),
          legend.position = c(0,0)
    )
}

theme_map2 <- function(scale = "continuous",
                            base_size = 8.5,
                            base_family = "Lato",
                            base_line_size = 0.5,
                            base_rect_size = 0.5) {

  gg <- theme_urbn_print(base_size = base_size,
                         base_family = base_family,
                         base_line_size = base_line_size,
                         base_rect_size = base_rect_size)

  gg <- gg +  ggplot2::theme(

      # make changes for mapping styles
      axis.text = ggplot2::element_blank(),
      axis.ticks = ggplot2::element_blank(),
      axis.title = ggplot2::element_blank(),
      panel.grid = ggplot2::element_blank(),
      axis.line = ggplot2::element_blank(),
      legend.margin = ggplot2::margin(t = 6L, r = 6L, b = 6L, l = 6L, "pt")
    )

  if (scale == "continuous") {
    gg <- gg + ggplot2::theme(
      legend.position = "right",
      legend.direction = "vertical",
      legend.title = ggplot2::element_text(size = base_size)
    )
  }
  # return gg
  gg
}

```




## Nominal tax vs GDP, % change

```{r eval=TRUE, include=TRUE}
df <- censustax %>%
  filter(stabbr=="US", name=="tottax") %>%
  select(year, tax=value) %>%
  inner_join(gdpfy, by="year") %>%
  pivot_longer(-year) %>%
  group_by(name) %>%
  mutate(pch=pchya(value, year)) %>%
  ungroup

recs <- recdata(1953)

p <- df %>%
  # filter(year >= 1970) %>%
  ggplot(aes(year, pch, colour=name)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0) +
  scale_x_continuous(name=NULL, breaks=seq(1950, 2030, 5)) +
  scale_y_continuous(name="% change",
                     breaks=seq(-.20, .20, .02),
                     labels=label_percent(accuracy=.1)) +
  scale_color_manual(values=bluered) +
  gband(recs$peak_decimal, recs$trough_decimal) +
  ggtitle("State Government Tax Revenue and Gross Domestic Product, United States") +
  theme_report +
  legend_notitle
p 
ggsave(filename = here::here("results", "gdp_tax_us.png"),
       plot = p, width=8, height=6)

```

## Inflation and Nominal tax % change

```{r include=TRUE, eval=TRUE}

glimpse(censustax)
df <- censustax %>%
  filter(stabbr=="US", name=="tottax") %>%
  select(year, tax=value) %>%
  inner_join(gdppi %>% select(year, gdppi), by="year") %>%
  pivot_longer(-year) %>%
  group_by(name) %>%
  mutate(pch=pchya(value, year)) %>%
  filter(year > 1951) %>%
  mutate(trend=hptrend(pch)) %>%
  ungroup
 
p <- df %>%
  filter(year >= 1955) %>%
  mutate(name=factor(name, 
                     levels=c("tax", "gdppi"), 
                     labels=c("taxes", "inflation"))) %>%
  ggplot(aes(year, trend, colour=name)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0) +
  scale_x_continuous(name=NULL, breaks=seq(1955, 2025, 5)) +
  scale_y_continuous(name="% change",
                     breaks=seq(-.20, .20, .02),
                     labels=label_percent(accuracy=.1)) +
  ggtitle("% change in trend tax revenue and trend general price level, United States") +
  scale_colour_manual(values=rev(bluered)) +
  theme_report +
  legend_notitle
p 
ggsave(filename = here::here("results", "gdppi_tax_us.png"), 
       plot = p, 
       width=8, height=6)
 
```

# HOW SHOULD WE MEASURE VOLATILITY?


# HOW DOES VOLATILITY VARY BY TAX?

```{r fig_pch_taxtype_box}

pdata <- vbase %>%
  filter(year %in% 2000:2020, 
         realnom=="nominal", 
         name %in% taxnames$name,
         stabbr != "US")
count(pdata, name)


gtitle1 <- "State tax revenue year over year growth rates by tax type,"
gtitle2 <- "\n50 states pooled, ranked by IQR of % change"
gtitle <- paste0(gtitle1, gtitle2)

p <- pdata %>%
  left_join(taxnames, by = "name") %>%
  ggplot(aes(x=pch, y=reorder(namef, pch, IQR, na.rm=TRUE))) +
  geom_boxplot(width=0.5, 
               lwd=.1, 
               outlier.size=.75,
               notch=FALSE, 
               fatten=6,
               fill="blue", 
               outlier.colour="blue",
               alpha=.2) +
  geom_vline(xintercept = 0,) +
  labs(x="percentage", y=NULL) +
  scale_x_continuous(breaks=seq(-2, 2, .1), 
                     labels = label_percent(accuracy=1), 
                     limits=c(-.38, .62)) +
  # scale_x_break(c(.65, 1.3)) +
  # scale_x_break(c(-.58, -.38)) +
  ggtitle(gtitle,
          subtitle="2000-2020") +
  theme_report
savefig(basename="fig_pch_taxtype_box")
```


```{r fig_pch_taxtype_decades_box}
count(volall, period)
# pdata <- volall %>%
#   filter(realnom=="nominal", 
#          !period %in% c("1960-1969"),
#          name %in% taxnames$name,
#          stabbr != "US") %>%
#   select(stabbr, name, namef, period, pchiqr, pdtrendiqr, sre) %>%
#   group_by(name, namef, period) %>%
#   summarise(across(c(pchiqr, pdtrendiqr, sre), ~ median(.x, na.rm=TRUE))) %>%
#   pivot_longer(-c(name, namef, period), names_to = "measure") %>%
#   pivot_wider(names_from=period) %>%
#   arrange(measure, desc(`1970-1979`))

pdata <- volall %>%
  filter(realnom=="nominal", 
         !period %in% c("1960-1969", "1970-2020"),
         name %in% taxnames$name,
         stabbr != "US") %>%
  select(stabbr, name, namef, period, data) %>%
  unnest(data)

gtitle1 <- "State tax revenue year over year growth rates by tax type,"
gtitle2 <- " 50 states pooled, ranked by IQR of % change"
gtitle <- paste0(gtitle1, gtitle2)

p <- pdata %>%
  ggplot(aes(x=pch, y=reorder(namef, pch, IQR, na.rm=TRUE))) +
  geom_boxplot(width=0.5, 
               lwd=.1, 
               outlier.size=.75,
               notch=FALSE, 
               fatten=6,
               fill="blue", 
               outlier.colour="blue",
               alpha=.2) +
  geom_vline(xintercept = 0,) +
  labs(x="percentage", y=NULL) +
  scale_x_continuous(breaks=seq(-2, 2, .1), 
                     labels = label_percent(accuracy=1), 
                     limits=c(-.38, .62)) +
  # scale_x_break(c(.65, 1.3)) +
  # scale_x_break(c(-.58, -.38)) +
  ggtitle(gtitle,
          subtitle="Five decades") +
  theme_report +
  facet_wrap(~period, ncol=1)
p
savefig(basename="fig_pch_taxtype_decades_box", width=6, height=8, scale=1.25)

```


```{r fig_pch_decades_taxtype_box}

pdata <- volall %>%
  filter(realnom=="nominal", 
         !period %in% c("1960-1969", "1970-2020"),
         name %in% taxnames$name,
         stabbr != "US") %>%
  select(stabbr, name, namef, period, data) %>%
  unnest(data)

gtitle <- "State tax revenue year over year growth rates by tax type over 5 decades"
gsubtitle <- "50 states pooled, box widths indicate IQR of % change"
p <- pdata %>%
  # mutate(period=factor(period, levels=unique(pdata$period))) %>%
  ggplot(aes(x=pch, y=reorder(period, desc(period)))) +
  geom_boxplot(width=0.5, 
               lwd=.1, 
               outlier.size=.75,
               notch=FALSE, 
               fatten=6,
               fill="blue", 
               outlier.colour="blue",
               alpha=.2) +
  geom_vline(xintercept = 0,) +
  labs(x="percentage", y=NULL) +
  scale_x_continuous(breaks=seq(-2, 2, .1), 
                     labels = label_percent(accuracy=1), 
                     limits=c(-.38, .62)) +
  # scale_x_break(c(.65, 1.3)) +
  # scale_x_break(c(-.58, -.38)) +
  ggtitle(gtitle,
          subtitle=gsubtitle) +
  theme_report +
  facet_wrap(~namef, ncol=1)
p
savefig(basename="fig_pch_decades_taxtype_box", width=6, height=8, scale=1.25)

```


```{r}
volall <- readRDS(here::here("data", "volall.rds"))
glimpse(volall)
count(volall, name)
volall %>%
  filter(name=="gdp", period=="1970-2020") %>%
  select(stabbr, realnom, value=pchiqr) %>%
  pivot_wider(names_from = realnom) %>%
  mutate(rrank=rank(-real), nrank=rank(-nominal)) %>%
  arrange(desc(nominal))

```


## Comparison of taxes

```{r}
# data check
glimpse(stack)
glimpse(vbase)

count(vbase, name)
check <- vbase %>%
  filter(name %in% c("tottax", "iit", "gst", "selsalestax", "cit", "sevtax", "othertax"),
         realnom=="nominal",
         stabbr=="US") %>%
  select(stabbr, name, year, value) %>%
  pivot_wider() %>%
  mutate(checksum=select(., -c(stabbr, year, tottax)) %>% 
           rowSums(na.rm=TRUE),
         check=checksum - tottax)
summary(check)

# look at tax shares
check %>%
  select(-c(checksum, check)) %>%
  mutate(across(-c(stabbr, year, tottax), ~ .x / tottax),
         big3=iit + gst + selsalestax) %>%
  ht
# big3 = 84% of total, less in early years

```


```{r fig_all_pdtrend}

vorder <- c("tottax", "iit", "gst", "selsalestax", "cit", "sevtax", "othertax")
(vnames <- taxnames$namef[match(vorder, taxnames$name)])

# we want a df with group=(total, detail), type=(iit, cit, ...othertax), and pdtrend
taxtype_order <- c("tottax", "iit", "gst", "selsalestax", "cit", "sevtax", "othertax")
level_order <- c("detail", "tottax")
# we want a df with level=(tottax, detail), taxtype=(iit, cit, ...othertax), and pdtrend
pdata  <- vbase %>%
  filter(name %in% vorder,
         realnom=="nominal",
         stabbr=="US",
         year >= 1970) %>%
  select(year, name, pdtrend) %>%
  # not sure why eval is required for self-join, but it is
  left_join(x=eval(.) %>% filter(name != "tottax") %>% select(year, taxtype=name, detail=pdtrend),
            y=eval(.) %>% filter(name=="tottax") %>% select(year, tottax=pdtrend),
            by="year") %>%
  pivot_longer(cols = c(detail, tottax), names_to = "level") %>%
  # add names for figure
  left_join(taxnames %>% rename(taxtype=name, taxtypef=namef), by = "taxtype") %>%
  mutate(levelf=ifelse(level=="detail", "Comparison tax", "Total taxes")) %>%
  # ensure that the data sort as desired
  mutate(taxtype=factor(taxtype, levels=taxtype_order),
         taxtypef=fct_reorder2(taxtypef, taxtypef, taxtype, .desc=FALSE),
         level=factor(level, levels=level_order),
         levelf=fct_reorder2(levelf, levelf, level, .desc=FALSE))

# verify ordering
count(pdata, taxtypef, taxtype) 
count(pdata, levelf, level)

years <- min(pdata$year):max(pdata$year)
recs <- recdata(years)
lsize <- .75
psize <- .75

ptop <- pdata %>%
  filter(taxtype %in% c("iit", "gst", "selsalestax")) %>%
  ggplot(aes(year, value, colour=levelf)) +
  # put recession bands down before lines and dots
  geom_band(recs) +
  geom_line(size=lsize) +
  geom_point(size=psize) +
  geom_hline(yintercept = 0) +
  scale_x_continuous(name=NULL, breaks=seq(1950, 2030, 5)) +
  scale_y_continuous(breaks=seq(-1, 1, .05),
                     limits=c(-.35, .35),
                     labels=label_percent(accuracy=1)) +
  # scale_colour_brewer(type="qual", palette="Set1") +
  # scale_color_viridis(discrete=TRUE) +
  scale_color_manual(values=bluegreen) +
  ggtitle("Volatility of major taxes",
          subtitle="Percentage deviation from trend") +
  labs(x=NULL, y="% deviation from trend",
       caption=NULL) +
  theme_report +
  legend_notitle +
  caption_left +
  facet_wrap(~taxtypef, ncol=3, scales="fixed")
ptop

pbottom <- pdata %>%
  filter(taxtype %in% c("cit", "sevtax", "othertax")) %>%
  ggplot(aes(year, value, colour=levelf)) +
  # put recession bands down before lines and dots
  geom_band(recs) +
  geom_line(size=lsize) +
  geom_point(size=psize) +
  geom_hline(yintercept = 0) +
  scale_x_continuous(name=NULL, breaks=seq(1950, 2030, 5)) +
  scale_y_continuous(breaks=seq(-1, 1, .05),
                     labels=label_percent(accuracy=1)) +
  # scale_colour_brewer(type="qual", palette="Set1") +
  # scale_color_viridis(discrete=TRUE) +
  scale_color_manual(values=bluegreen) +
  ggtitle("Volatility of smaller taxes",
          subtitle="Percentage deviation from trend") +
  labs(x=NULL, y="% deviation from trend",
       caption=NULL) +
  theme_report +
  legend_notitle +
  caption_left +
  facet_wrap(~taxtypef, ncol=3, scales="fixed")
pbottom

p <- ptop / pbottom
p
savefig("fig_all_pdtrend", .p=p, .pdata=pdata, width=10, height=6, scale=1.5) 

```


```{r tab_all_taxes}

# for each tax, get:
#   share of total tax revenue in latest year
#   summary volatility measures
#     pdtrend
#     pchya
#     sre
vorder <- c("tottax", "iit", "gst", "selsalestax", "cit", "sevtax", "othertax")
tabbase <- vbase %>%
  filter(name %in% vorder,
         realnom=="nominal",
         stabbr=="US",
         year >= 1969) # so we can get lagged value

shares <- tabbase %>%
  filter(year==2020) %>%
  select(name, year, value) %>%
  mutate(share=value / value[name=="tottax"])

vol <- tabbase %>%
  # bring in nominal gdp
  left_join(vbase %>% 
              filter(name=="gdp", realnom=="nominal", stabbr=="US") %>%
              select(year, gdp=value),
            by=c("year")) %>%
  group_by(name) %>%
  mutate(dlvalue=dl(value), dlgdp=dl(gdp)) %>%
  nest() %>%
  # note that this does not filter out the case where GDP is on the lhs and the rhs so we will get warnings for that!
  # they will arise in the next step when we unpack
  mutate(sremod = purrr::map(data, ~ lm(dlvalue ~ dlgdp, data = .x)))


  # mutate(sremod = purrr::map(data, safely(function(df) lm(dlvalue ~ dlgdp, data=df))),
  #        epi = purrr::map(data, safely(function(df) episodes(df$pdtrend)))) %>%
  # ungroup

# sre -- short-run elasticity -- unpack into sre and sresum
# unpack the summary measures from sremod to get:
#   coefficient (short-run elasticity, sre), and
#   its standard error
sredf <- vol %>%
  select(name, sremod) %>%
  mutate(tidied = purrr::map(sremod, tidy)) %>% 
  unnest(cols=tidied) %>%
  filter(term=="dlgdp") %>%
  ungroup %>%
  select(name, sre=estimate, srese=std.error)

sredf %>% arrange(desc(sre))
sredf %>% arrange(desc(srese))

tabdata <- vol %>%
  select(name, data) %>%
  unnest(cols=data) %>%
  group_by(name) %>%
  summarise(pdtrend=IQR(abs(pdtrend)),
            pchiqr=IQR(abs(pch) - median(abs(pch)))) %>%
  ungroup %>%
  left_join(sredf, by = "name") %>%
  left_join(shares %>%
              select(name, share), 
            by = "name") %>%
  mutate(namef=factor(name, 
                      levels=taxnames$name, 
                      labels=taxnames$namef)) %>%
  relocate(c(namef, share), .after = name) %>%
  arrange(namef)
  
tabdata

tab <- tabdata %>%
  gt() %>%
  cols_hide(name) %>%
  tab_header(
    title = "Selected volatility measures by tax type for the U.S. as a whole",
    subtitle = "Estimated over 1970-2020"
    ) %>%
  cols_label(namef="Type of tax",
             share=html("% share of<br>total tax<br>revenue (2020)"),
             pdtrend=html("% deviation<br>from trend (IQR)"),             
             pchiqr=html("annual % change:<br>absolute value minus<br>its median (IQR)"),
             sre=html("short-run<br>elasticity (SRE)"),
             srese=html("standard error<br>of SRE")) %>%
  cols_align(
    align = c("left"),
    columns = namef) %>%
  fmt_percent(
    columns = c(share, pdtrend, pchiqr),
    decimals = 1) %>%
  fmt_number(
    columns = c(sre, srese),
    decimals = 2) %>%
  tab_style(
    style = list(
      cell_fill(color = "grey95")
      ),
    locations = cells_body(
      rows=is.even(1:nrow(tabdata))
      )
    )
tab

savetab("tab_all_taxes")
# 
# level_order <- c("detail", "tottax")
# # we want a df with level=(tottax, detail), taxtype=(iit, cit, ...othertax), and pdtrend
# pdata  <- vbase %>%
#   filter(name %in% vorder,
#          realnom=="nominal",
#          stabbr=="US",
#          year >= 1970)
# 
# f <- function(tabdata, tax){
#   tab <- tabdata %>%
#   left_join(mlabels, by = "measure") %>%
#   mutate(measure=factor(measure, levels=meas)) %>%
#   arrange(measure) %>%
#   select(mlabel, actual, adjust, ratio) %>%
#   gt() %>%
#     tab_header(
#       title = paste0(tax, " volatility measures for median state, with and without adjustment for tax-rate changes"),
#       subtitle = ""
#       ) %>%
#       cols_label(mlabel=html("Volatility<br>measure"),
#                  actual=html("NOT adjusted for<br>tax rate<br>changes"),
#                  adjust=html("Adjusted for<br>tax rate<br>changes"),
#                  ratio=html("Ratio of<br>median adjusted<br>to<br>median unadjusted")) %>%
#       tab_spanner(
#         label = "Rate adjustment:",
#         columns = c(actual, adjust)
#         ) %>%
#       fmt_percent(
#         columns = c(actual, adjust),
#         decimals = 1) %>%
#       # selectively override fmt_percent
#       fmt_number(
#         columns = ratio,
#         decimals = 2) %>%
#       fmt_number(
#         columns = c(actual, adjust),
#         rows = c(4),
#         decimals = 2) %>%
#     tab_style(
#       style = list(
#         cell_fill(color = "lightgrey")
#       ),
#       locations = cells_body(
#         rows=c(2, 4)
#       )
#     ) %>%
#     tab_footnote(
#       footnote = "Adjusted for inflation with GDP price index. Estimated over 1978-2018.",
#       locations = cells_title(groups = "title")
#       )%>%
#     tab_footnote(
#       footnote = "A ratio > 1 means volatility is greater after adjustment for tax-rate changes.",
#       locations = cells_column_labels(columns = ratio)
#       )
#   tab
# }

```



## Personal income tax

```{r fig_capgains_agi}

# get capgains as share of agi -- both datasets already in memory
pdata <- agi %>%
  select(-src) %>%
  left_join(capgains, by = "year") %>%
  mutate(cgagi=capgains / agi)

p <- pdata %>%
  na.omit() %>%
  ggplot(aes(year, cgagi)) +
  geom_line(size=1, colour="blue") +
  geom_point(size=1, colour="blue") +
  scale_x_continuous(name=NULL, breaks=seq(1950, 2020, 5)) +
  scale_y_continuous(name="% of AGI",
                     breaks=seq(-.20, .20, .02),
                     labels=label_percent(accuracy=1),
                     limits=c(0, NA)) +
  recband(recdata(min(pdata$year):max(pdata$year))) +
  ggtitle("Realized capital gains as a percentage of adjusted gross income") +
  labs(caption=paste0("Source: ", source_soi)) +
  theme_report +
  annotate(geom="text", x=1986.5, y=pdata$cgagi[pdata$year==1986],
           hjust=0,
           colour="darkgreen",
           size=2,
           label="Acceleration before 1986\ntax reform took effect")
p

savefig("fig_capgains_agi")

```

```{r fig_iitmod_cgresids}

yearplus <- 1986

moddata1 <- left_join(
  censustax %>%
    filter(name=="iit", stabbr=="US") %>%
    select(year, iit=value),
  gdpfy,
  by = "year") %>%
  pivot_longer(-year) %>%
  group_by(name) %>%
  arrange(year) %>%
  mutate(pchya=pchya(value, year)) %>%
  select(year, name, pchya) %>%
  pivot_wider(values_from = pchya)

moddata2 <- left_join(capgains, 
                 agi %>% select(-src),
                 by = "year") %>%
  arrange(year) %>%
  mutate(cgagi=capgains / agi,
         dcgagi=diffya(cgagi, year),
         ldcgagi=lag(dcgagi))

moddata <- moddata1 %>%
  left_join(moddata2 %>% select(year, cgagi, dcgagi, ldcgagi), by = "year")

modxcg <- lm(iit ~ gdp, na.action = na.exclude, data=moddata)
summary(modxcg)
residuals(modxcg)

pdata <- moddata %>%
  mutate(fit=fitted(modxcg),
         resids=residuals(modxcg),
         ygroup=ifelse(year < yearplus, paste0("pre-", yearplus), paste0(yearplus, "-plus")),
         ygroup=factor(ygroup, 
                       levels=c(paste0("pre-", yearplus), 
                                paste0(yearplus, "-plus"))
                       )) %>%
  arrange(ygroup, year) %>%
  select(year, resids, ldcgagi, ygroup) %>%
  na.omit() %>%
  pivot_longer(-c(ygroup, year)) %>%
  mutate(namef=factor(name, 
                     levels=c("resids", "ldcgagi"),
                     labels=str_wrap(
                       c("Unexplained income tax change",
                         "Capital gains change"),
                       15))) %>%
  arrange(year, namef)

# recessions data frame for bands
recdf <- recdata(pdata$year) %>%
  select(rec_year, peak_decimal, trough_decimal) %>%
  left_join(pdata %>% select(rec_year=year, ygroup), by="rec_year") # to get ygroup as factor

gtitle <- paste0("Portion of income tax growth not explained by GDP")
p <- pdata %>%
  ggplot(aes(year, value, colour=namef)) +
  # put recession bands before other layers
  geom_band(recdf) +
  geom_line(size=1) +
  geom_point(size=1) +
  scale_x_continuous(name="State fiscal year for income tax, and tax year for lagged gains change",
                     breaks=seq(1950, 2030, 5)) +
  scale_y_continuous(name=NULL,
                     breaks=seq(-.20, .20, .05),
                     labels=label_percent(accuracy=.1),
                     limits=c(-.14, .17)) +
  geom_hline(yintercept = 0) +
  scale_color_manual(values=bluered) +
  geom_text(
    data=pdata %>%
      filter(year==1987, name=="ldcgagi"),
    mapping=aes(x=year, y=value + .025),
    label="1986 tax\nreform impact", 
    hjust=0.5,
    colour="darkgreen",
    size=2,
    inherit.aes = FALSE, show.legend = FALSE) +
  ggtitle(gtitle,
          subtitle="Compared to change in capital gains as % of AGI, lagged one year") +
  labs(caption=paste0("\nSource: ", source_bea, " and ", source_cbo)) +
  caption_left +
  theme_report +
  legend_notitle +
  theme(legend.key.size = unit(2, 'lines')) +
  facet_wrap(~ygroup, ncol = 1, scales = "free")
p 
savefig("fig_iitmod_cgresids")

# https://stackoverflow.com/questions/55748379/use-annotaterect-for-one-facet-in-ggplot-when-plotting-time-series-on-x-axis

```

```{r iit_soi}
# get income components data from Urban TPC
dsoi <- r"(E:\data\soi)"
fn <- "historical_source_0.xlsx"
url <- "https://www.taxpolicycenter.org/file/185517/download?token=cYMr2RRH"
fpath <- file.path(dsoi, fn)
download.file(url, fpath, mode="wb")

# year	nret	agi	wages	interest	dividends	busincnet	netcgll	incother	taxbc	taxliab	amt
# 											
vnames <- c("year", "nret", "agi", "wages", "interest", "dividends",
            "busincnet", "netcgll", "incother", "taxbc", "taxliab", "amt")
df <- read_excel(fpath, 
                 col_names = vnames,
                 col_types = "text", 
                 skip=6)
ht(df)

df2 <- df %>%
  select(-c(taxliab, amt)) %>%
  mutate(year=as.integer(year)) %>%
  filter(!is.na(year)) %>%
  mutate(across(-c(year), as.numeric)) %>%
  filter(year >= 1921) %>%
  mutate(intdiv=naz(interest) + naz(dividends)) %>%
  pivot_longer(-year)

df3 <- df2 %>%
  filter(year >= 1954) %>%
  arrange(name, year) %>%
  group_by(name) %>%
  mutate(trend=hptrend(value, smooth=6.25),
         pdtrend=value / trend - 1) %>%
  ungroup

df3 %>%
  filter(year >= 1955, 
         !name %in% c("taxbc", "interest", "dividends", "nret")) %>%
  ggplot(aes(year, pdtrend)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0) +
  facet_wrap(~name, ncol=2)

df3 %>%
  filter(year >= 1955, 
         name %in% c("agi", "wages", "busincnet", "netcgll")) %>%
  ggplot(aes(year, pdtrend, colour=name)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0)

df3 %>%
  filter(year >= 1955, 
         name %in% c("agi", "wages", "busincnet", "netcgll")) %>%
  group_by(name) %>%
  summarise(mabs=mean(abs(pdtrend)), sd=sd(pdtrend))


```


## General sales tax
## Selective sales taxes
## Corporate income tax
## Other taxes
# HOW DOES THE ECONOMY AFFECT TAX REVENUE VOLATILITY?

# HOW HAS VOLATILITY CHANGED OVER TIME?

```{r}
# https://socviz.co/maps.html
us_states <- map_data("state")
head(us_states)
count(us_states, region) # lower 48 plus district of columbia

p <- ggplot(data = us_states,
            mapping = aes(x = long, y = lat,
                          group = group))

p + geom_polygon(fill = "white", color = "black")

p <- us_states %>%
  ggplot(aes(x = long, y = lat, group = group, fill = region))
p
p + geom_polygon(color = "gray90", size = 0.1) + guides(fill = "none")

p + geom_polygon(color = "gray90", size = 0.1) +
    coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
    guides(fill = "none")

# https://github.com/UrbanInstitute/urbnmapr
devtools::install_github("UrbanInstitute/urbnmapr")
library(urbnmapr)
library(sf)

states_sf <- get_urbn_map("states", sf = TRUE)
str(states_sf)
head(states_sf)
class(states_sf)
st2 <- st_crs(states_sf, parameters=TRUE)

states_sf %>% 
  ggplot(aes()) +
  geom_sf(fill = "grey", color = "#ffffff")

glimpse(fifty_states)
count(fifty_states, id)
p <- fifty_states %>%
  ggplot(aes(x = long, y = lat, group = group, fill = id)) +
  geom_polygon(color = "gray90", size = 0.1) + 
  guides(fill = "none") +
  geom_polygon(color = "gray90", size = 0.1) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  guides(fill = "none")
p


```


# HOW DO PORTFOLIOS OF TAXES AFFECT TOTAL REVENUE VOLATILITY?
```{r}
vbase %>%
  filter(year %in% 1970:2020) %>%
  filter(name %in% taxnames$name, name!="sevtax") %>%
  group_by(name) %>%
  summarise(pch_mdn=median(pch, na.rm=TRUE),
            pch_iqr=IQR(pch, na.rm=TRUE)) %>%
  ggplot(aes(pch_iqr, pch_mdn, label=name)) +
  geom_point() +
  geom_text()
  
censustax
summary(sgtax.a)
glimpse(sgtax.a)
count(sgtax.a, item, desc)
count(sgtax.a, year)
keep <- c("C105", "T09", "T10", "T13", "T16", "T40", "T41", "T50", "T53")
df <- sgtax.a %>%
  filter(year %in% 1970:2020, item %in% keep) %>%
  arrange(item, desc, stabbr, year) %>%
  group_by(item, desc, stabbr) %>%
  mutate(pch=pchya(value, year)) %>%
  group_by(item, desc) %>%
  summarise(pch_mdn=median(pch, na.rm=TRUE),
            pch_iqr=IQR(pch, na.rm=TRUE),
            .groups="drop")

df %>%
  # mutate(name=abbreviate(desc, minlength = 6)) %>%
  # mutate(name=str_replace(desc, " ", ""),
  #        name=str_trunc(name, 10, side="right")) %>%
  mutate(name=str_wrap(str_trunc(desc, 30, side="right"), width=10)) %>%
  ggplot(aes(pch_iqr, pch_mdn, label=name)) +
  geom_point() +
  geom_text(position=position_dodge(width=.001)) +
  geom_line(data=. %>% filter(item %in% c("T10", "C105", "T40")))
```



# HOW DOES VOLATILITY VARY ACROSS STATES?

```{r}

# for each tax, get:
#   share of total tax revenue in latest year
#   summary volatility measures
#     pdtrend
#     pchya
#     sre
vorder <- c("tottax", "iit", "gst", "selsalestax", "cit", "sevtax", "othertax")
tabbase <- vbase %>%
  filter(name %in% vorder,
         realnom=="nominal",
         # stabbr=="US",
         year >= 1969) # so we can get lagged value

shares <- tabbase %>%
  filter(year==2020) %>%
  select(stabbr, name, year, value) %>%
  group_by(stabbr) %>%
  mutate(share=value / value[name=="tottax"])

vol <- tabbase %>%
  # bring in nominal gdp
  left_join(vbase %>% 
              filter(name=="gdp", realnom=="nominal") %>%
              select(stabbr, year, gdp=value),
            by=c("stabbr", "year")) %>%
  group_by(stabbr, name) %>%
  mutate(dlvalue=dl(value), dlgdp=dl(gdp)) %>%
  nest() %>%
  # we must use "safely" because some state-tax combinations might be bad -- for example,
  # having a zero value for tax revenue which, when we take the log, became NA
  mutate(sremod = purrr::map(data, safely(function(df) lm(dlvalue ~ dlgdp, data=df)))) %>%
  ungroup

# sre -- short-run elasticity -- unpack into sre and srese
# unpack the summary measures from sremod to get:
#   coefficient (short-run elasticity, sre), and
#   its standard error
sredf <- vol %>%
  ungroup %>%
  select(stabbr, name, sremod) %>%
  mutate(result=purrr::map(sremod, "result"),
         tidied=purrr::map(result, tidy)) %>%
  select(-c(sremod, result)) %>%
  unnest(cols=tidied) %>%
  filter(term=="dlgdp") %>%
  select(stabbr, name, sre=estimate, srese=std.error)
# note - we lost MO sevtax; it has some junky values like 0, 1, etc.

# get the other measures
vol_all <- vol %>%
  select(stabbr, name, data) %>%
  unnest(cols=data) %>%
  group_by(stabbr, name) %>%
  summarise(pdtrend=IQR(abs(pdtrend), na.rm=TRUE),
            pchiqr=IQR(abs(pch) - median(abs(pch)), na.rm=TRUE),
            .groups="drop") %>%
  ungroup %>%
  left_join(sredf, by = c("stabbr", "name")) %>%
  left_join(shares %>%
              select(stabbr, name, share), 
            by = c("stabbr", "name")) %>%
  mutate(namef=factor(name, 
                      levels=taxnames$name, 
                      labels=taxnames$namef)) %>%
  relocate(c(namef, share), .after = name) %>%
  arrange(namef)
  
vol_all

vol_all %>%
  filter(name=="tottax") %>%
  arrange(desc(pdtrend))

# 
# tab <- tabdata %>%
#   gt() %>%
#   cols_hide(name) %>%
#   tab_header(
#     title = "Selected volatility measures by tax type for the U.S. as a whole",
#     subtitle = "Estimated over 1970-2020"
#     ) %>%
#   cols_label(namef="Type of tax",
#              share=html("% share of<br>total tax<br>revenue (2020)"),
#              pdtrend=html("% deviation<br>from trend (IQR)"),             
#              pchiqr=html("annual % change:<br>absolute value minus<br>its median (IQR)"),
#              sre=html("short-run<br>elasticity (SRE)"),
#              srese=html("standard error<br>of SRE")) %>%
#   cols_align(
#     align = c("left"),
#     columns = namef) %>%
#   fmt_percent(
#     columns = c(share, pdtrend, pchiqr),
#     decimals = 1) %>%
#   fmt_number(
#     columns = c(sre, srese),
#     decimals = 2) %>%
#   tab_style(
#     style = list(
#       cell_fill(color = "grey95")
#       ),
#     locations = cells_body(
#       rows=is.even(1:nrow(tabdata))
#       )
#     )
# tab
# 
# savetab("tab_all_taxes")
# 
```


```{r}


pdata <- vbase %>%
  filter(year %in% 2000:2020, realnom=="nominal", name=="tottax") 

p <- pdata %>%
  mutate(stname=stname(stabbr)) %>%
  ggplot(aes(x=pch, y=reorder(stname, pch, IQR, na.rm=TRUE))) +
  geom_boxplot(width=0.5, 
               lwd=.1, 
               outlier.size=.75,
               notch=FALSE, 
               fill="blue", 
               outlier.colour="blue",
               alpha=.2) +
  geom_vline(xintercept = 0) +
  labs(x="percentage", y=NULL) +
  scale_x_continuous(breaks=seq(-2, 2, .1), 
                     labels = label_percent(accuracy=1), 
                     limits=c(-.38, .62)) +
  # scale_x_break(c(.65, 1.3)) +
  # scale_x_break(c(-.58, -.38)) +
  ggtitle("State tax revenue year over year growth rates, ranked by IQR of % change",
          subtitle="2000-2020") +
  theme_report
savefig(basename="fig_pch_box")

# not really the efficient frontier, but rather the attractive frontier!
pdata %>%
  filter(stabbr != "AK") %>%
  group_by(stabbr) %>%
  summarise(pch_mdn=median(pch, na.rm=TRUE),
            pch_iqr=IQR(pch, na.rm=TRUE),
            .groups="drop") %>%
  ggplot(aes(pch_iqr, pch_mdn, label=stabbr)) +
  geom_point(size=1) +
  geom_text_repel(size=1.5) +
  # geom_text(position=position_dodge(width=.001)) +
  geom_line(data=. %>% filter(stabbr %in% c("MD", "ID", "ND")))

```



```{r scatter}
shares <- tabbase %>%
  select(stabbr, name, year, value) %>%
  group_by(stabbr, year) %>%
  mutate(share=value / value[name=="tottax"])

shares_mean <- shares %>%
  group_by(stabbr, name) %>%
  summarise(value=mean(share, na.rm=TRUE), .groups="drop") %>%
  pivot_wider(values_fill = 0)

pdata_wide <- vol_all %>%
  filter(name=="tottax") %>%
  select(-name, -namef, -share) %>%
  left_join(shares_mean %>% select(-tottax), by="stabbr")

pdata <- pdata_wide %>%
  pivot_longer(-c(stabbr, pdtrend, pchiqr, sre, srese))

pdata %>%
  filter(stabbr!="AK") %>%
  ggplot(aes(value, pdtrend, label=stabbr)) +
  geom_point() +
  geom_text() +
  facet_wrap(~name, ncol=3, scales="free")

pdata %>%
  filter(name=="sevtax") %>%
  mutate(stabbr2=ifelse(value >= .05, stabbr, "")) %>%
  ggplot(aes(value, pdtrend, label=stabbr2)) +
  geom_point() +
  geom_text()

pdata %>%
  filter(pdtrend <= .05, name!="sevtax") %>%
  # mutate(stabbr2=ifelse(value >= .05, stabbr, "")) %>%
  ggplot(aes(value, pdtrend, label=stabbr)) +
  geom_point() +
  geom_text() +
  facet_wrap(~name, ncol=3, scales="free")

sts <- c("AK", "ND", "WY")
mod <- lm(pdtrend ~ iit + gst + cit, data=pdata_wide %>% filter(!stabbr %in% sts))
mod <- lm(pdtrend ~ iit + cit + sevtax, data=pdata_wide %>% filter(!stabbr %in% sts))
summary(mod)


```





# COME BACK TO THE ITEMS BELOW
## Alternative volatility measures with made-up data

```{r volatile_data}
set.seed(33)  
# 15 best so far
# 12 some long negatives -- rand much more vol w/pchsd
# 1234 a little too uniform
# 4 quite good
# 14 pretty good
# 28 very good

# create data with different kinds of volatility
nyears <- 30
time <- 0:(nyears - 1)
grate <- .03
init <- 100
trend <- init * (1 + grate)^time
nturns <- 6
x <- seq(0, 1, length.out = nyears)
sinx <- sin(nturns * pi * x)
# plot(sinx)

(noise_prop <- c(0, rnorm(nyears - 1, mean=0, sd=.075)))
cycle_size <- .1
mud <- tibble(time=time) %>%
  mutate(trend=trend, 
         cycle=trend * (1 + sinx * cycle_size),
         cycle=cycle * sum(trend) / sum(cycle),
         random=trend * (1 + noise_prop),
         random=random * sum(trend) / sum(random))

mud %>% summarise(across(c(trend, cycle, random), sum))


# create long version of made-up data
mudl <- mud %>%
  pivot_longer(-time) %>%
  group_by(name) %>%
  mutate(pch=value / value[match(time - 1, time)] - 1) %>%
  group_by(time) %>%
  mutate(pdtrend=value / value[name=="trend"] - 1) %>%
  ungroup

mudl
max(abs(mudl$pch), na.rm=TRUE)
max(mudl$pch, na.rm=TRUE)
min(mudl$pch, na.rm=TRUE)

mudl %>%
  filter(name=="cycle")

dfl %>%
  filter(name=="random")

mudl %>% 
  group_by(name) %>% 
  summarise(across(c(value, pch, pdtrend), 
                   list(mean=~mean(.x, na.rm=TRUE),
                        min=~min(.x, na.rm=TRUE),
                        max=~max(.x, na.rm=TRUE))))
```

## Random and cyclical fluctuation

```{r random_cycle, eval=TRUE, include=TRUE}

colors = c("black", "blue")

# levels
p1 <- mudl %>%
  filter(name %in% c("trend", "random")) %>%
  mutate(name=factor(name, levels=c("trend", "random"))) %>%  # to get right sort order
  ggplot(aes(time, value, colour=name)) +
  geom_line() +
  geom_hline(yintercept = 100) +
  scale_y_continuous(name="$ revenue", limits=c(90, 265), labels=scales::dollar) +
  scale_x_continuous(name="year", breaks=seq(0, 40, 5), limits=c(0, 32)) +
  scale_colour_manual(values=colors) +
  ggtitle("Random revenue fluctuations",
          subtitle="Annual amount of revenue") +
  theme_report +
  legend_notitle
p1

# percent change
p2 <- mudl %>%
  filter(name %in% c("trend", "random")) %>%
  mutate(name=factor(name, levels=c("trend", "random"))) %>%  # to get right sort order
  ggplot(aes(time, pch, colour=name)) +
  geom_line() +
  geom_hline(yintercept = 0, colour="grey") +
  scale_y_continuous(name="% change", 
                     breaks=seq(-.5, .5, .1),
                     limits=c(-.3, .3), 
                     labels=scales::percent_format(accuracy=1)) +
  scale_x_continuous(name="year", breaks=seq(0, 40, 5), limits=c(0, 32)) +
  scale_colour_manual(values=colors) +
  ggtitle("Random revenue fluctuations",
          subtitle="Annual percent change") +
  theme_report +
  legend_notitle
p2

# % diff from trend
p3 <- mudl %>%
  filter(name=="random") %>%
  mutate(posneg=ifelse(pdtrend < 0, "neg", "pos")) %>%
  ggplot(aes(time, pdtrend)) +
  geom_hline(yintercept = 0) +
  geom_bar(stat = "identity", aes(fill = posneg)) + 
  scale_fill_manual(values=rev(bluered)) +
  scale_x_continuous(name="year", breaks=seq(0, 40, 5), limits=c(0, 32)) +
  scale_y_continuous(name="% difference from trend",
                     breaks=seq(-.2, .2, .05), 
                     labels=scales::percent_format(accuracy=1),
                     limits=c(-.2, .2)) +
  ggtitle("Random revenue fluctuations",
          subtitle="Percent difference from trend") +
  theme_report +
  legend_none
p3


p123 <- p1 / p2 / p3
p123
# ggsave(filename = here::here("results", "trend_random_3panel.png"), plot = p123, width=8, height=8)

# Cyclical variation

# levels
p4 <- mudl %>%
  filter(name %in% c("trend", "cycle")) %>%
  mutate(name=factor(name, levels=c("trend", "cycle"))) %>%
  ggplot(aes(time, value, colour=name)) +
  geom_line() +
  geom_hline(yintercept = 100) +
  scale_y_continuous(name="$ revenue", limits=c(90, 265), labels=scales::dollar) +
  scale_x_continuous(name="year", breaks=seq(0, 40, 5), limits=c(0, 32)) +
  scale_colour_manual(values=colors) +
  ggtitle("Cyclical revenue fluctuations",
          subtitle = "Annual amount of revenue") +
  theme_report +
  legend_notitle
p4

# percent change
p5 <- mudl %>%
  filter(name %in% c("trend", "cycle")) %>%
  mutate(name=factor(name, levels=c("trend", "cycle"))) %>%  # to get right sort order
  ggplot(aes(time, pch, colour=name)) +
  geom_line() +
  geom_hline(yintercept = 0, colour="grey") +
  scale_y_continuous(name="% change", 
                     breaks=seq(-.5, .5, .1),
                     limits=c(-.3, .3), 
                     labels=scales::percent_format(accuracy=1)) +
  scale_x_continuous(name="year", breaks=seq(0, 40, 5), limits=c(0, 32)) +
  scale_colour_manual(values=colors) +
  ggtitle("Cyclical revenue fluctuations",
          subtitle = "Annual percent change") +
  theme_report +
  legend_notitle
p5

# % diff from trend
p6 <- mudl %>%
  filter(name=="cycle") %>%
  mutate(posneg=ifelse(pdtrend < 0, "neg", "pos")) %>%
  ggplot(aes(time, pdtrend)) +
  geom_hline(yintercept = 0) +
  geom_bar(stat = "identity", aes(fill = posneg)) + 
  scale_fill_manual(values=rev(bluered)) +
  scale_x_continuous(name="year", breaks=seq(0, 40, 5), limits=c(0, 32)) +
  scale_y_continuous(name="% difference from trend",
                     breaks=seq(-.2, .2, .05), 
                     labels=scales::percent_format(accuracy=1),
                     limits=c(-.2, .2)) +
  ggtitle("Cyclical revenue fluctuations",
          subtitle="Percent difference from trend") +
  theme_report +
  legend_none
p6

p456 <- p4 / p5 / p6

p456
# ggsave(filename = here::here("results", "trend_cycle_3panel.png"), plot = p456, width=8, height=8)

p7 <- p123 | p456
p7
ggsave(filename = here::here("results", "trend_cycle_random.png"),
       plot = p7, width=8, height=8, scale=1)

```

```{r vol_measured}

vmudl <- mudl %>%
  filter(time > 0) %>%
  group_by(name) %>%
  summarise(pchsd = sd(pch, na.rm=TRUE),
            pdtsd = sd(pdtrend, na.rm=TRUE))

vmudl %>% 
  kbl(format="html", digits=2)

mudl %>%
  filter(time > 0) %>%
  group_by(name) %>%
  nest() %>%
  mutate(epi = purrr::map(data, function(df) episodes(df$pdtrend))) %>%
  unnest_wider(col=epi) %>%
  mutate(esum=esum) %>% 
  select(-data) %>%
  kbl(format="html", digits=2)

```

## Reserve fund example

[TO COME: Reserve fund example with real data for the two different volatility measures.]

## Variation relative to the economy

```{r econ_cycle}
# look at growth pdiff faster and pdiff slower than the economy
pdiff <- .25

f <- function(vec){
  # function to get simple constant-growth trend for a vector
  n <- length(vec)
  growth <- vec[n] / vec[1]
  g <- growth ^ (1 / (n - 1))  # this is 1 + the constant growth rate
  trend <- vec[1] * g^(0:(n - 1))
  trend
}

econ1 <- mud %>%
  select(time, econ_actual=cycle) %>%
  mutate(econ_pch=pchya(econ_actual, time),
         econ_pch=ifelse(time==0, 0, econ_pch),
         taxplus_pch=econ_pch * (1 + pdiff),
         taxplus_actual=cumprod(1 + taxplus_pch),
         taxminus_pch=econ_pch * (1 - pdiff),
         taxminus_actual=cumprod(1 + taxminus_pch)) %>%
  # calculate a simple constant-growth trend for each
  mutate(econ_trend=f(econ_actual),
         taxplus_trend=f(taxplus_actual),
         taxminus_trend=f(taxminus_actual)) %>%
  # percent differences from respective trends
  mutate(econ_pdt=econ_actual / econ_trend - 1,
         taxplus_pdt=taxplus_actual / taxplus_trend - 1,
         taxminus_pdt = taxminus_actual / taxminus_trend - 1) %>%
  pivot_longer(-time) %>%
  arrange(name, time) %>%
  separate(name, c("variable", "measure"))

# add a series with the simple constant-growth trend for each
trend_growth <- econ1 %>%
  filter(measure=="actual") %>%
  group_by(variable) %>%
  mutate(trgrowth=(value[time==max(time)] / value[time==0])^(1/(n()-1)),
         trgrowth=trgrowth - 1,
         measure="trgrowth") %>%
  select(time, variable, measure, value=trgrowth)

econ <- bind_rows(econ1, trend_growth)
econ

econwide <- econ %>%
  pivot_wider(names_from = c(variable, measure)) %>%
  mutate(econ_dl=dl(econ_actual), 
         taxplus_dl=dl(taxplus_actual),
         taxminus_dl=dl(taxminus_actual))

m1 <- lm(taxplus_pch ~ econ_pch, data=econwide)
summary(m1)

m2 <- lm(taxminus_pch ~ econ_pch, data=econwide)
summary(m2)

m1a <- lm(taxplus_dl ~ econ_dl, data=econwide %>% filter(time > 0))
summary(m1a)

m1b <- lm(taxminus_dl ~ econ_dl, data=econwide %>% filter(time > 0))
summary(m1b)

```

```{r econ_6panel_graph}

# econ
# summary(econ)
# count(econ, variable)  # econ, taxplus, taxminus
# count(econ, measure)  # actual, pch, pdt, trend
  
colors = c("black", "blue", "darkgreen")

# levels
p1 <- econ %>%
  filter(variable %in% c("econ", "taxminus"), measure=="actual") %>%
  bind_rows(econ %>% filter(variable=="taxminus", measure=="trend")) %>%
  unite("name", variable, measure) %>%
  mutate(name=factor(name, 
                     levels=c("econ_actual", "taxminus_actual", "taxminus_trend"),
                     labels=c("economy", "revenue",  "revenue trend"))) %>%  # to get right sort order
  ggplot(aes(time, value, colour=name)) +
  geom_line() +
  scale_y_continuous(name=NULL, limits=c(100, 300), labels=scales::dollar_format(scale=1)) +
  scale_x_continuous(name="time", limits=c(NA, NA), breaks=seq(0, 30, 5)) +
  scale_colour_manual(values=colors) +
  ggtitle("Revenue that grows 25% slower than economy",
          subtitle="Annual amount indexed to $100 in year 0") +
  theme_report +
  legend_notitle
p1

# percent change
p2 <- econ %>%
  filter(variable %in% c("econ", "taxminus"), measure=="pch") %>%
  bind_rows(econ %>% filter(variable=="taxminus", measure=="trgrowth")) %>%
  unite("name", variable, measure) %>%
  mutate(name=factor(name, 
                     levels=c("econ_pch", "taxminus_pch", "taxminus_trgrowth"),
                     labels=c("economy", "revenue", "revenue trend"))) %>%  # to get right sort order
  ggplot(aes(time, value, colour=name)) +
  geom_line() +
  geom_hline(yintercept = 0, colour="grey") +
  scale_y_continuous(name="% change", 
                     breaks=seq(-.5, .5, .05),
                     limits=c(-.15, .15), 
                     labels=scales::percent_format(accuracy=1)) +
  scale_x_continuous(name="time", breaks=seq(0, 30, 5)) +
  scale_colour_manual(values=colors) +
  ggtitle("Revenue that grows 25% slower than economy",
          subtitle="Annual percent change") +
  theme_report +
  legend_notitle
p2

# percent difference from trend
p3 <- econ %>%
  filter(variable %in% c("taxminus"), measure=="pdt") %>%
  mutate(posneg=ifelse(value < 0, "neg", "pos")) %>%
  ggplot(aes(time, value)) +
  geom_hline(yintercept = 0) +
  geom_bar(stat = "identity", aes(fill = posneg)) + 
  scale_fill_manual(values=rev(bluered)) +
  scale_x_continuous(name="time", breaks=seq(0, 30, 5)) +
  scale_y_continuous(name="% difference from trend",
                     breaks=seq(-.5, .5, .05), 
                     limits=c(-.15, .15),
                     labels=scales::percent_format(accuracy=1)) +
  ggtitle("Revenue that grows 25% slower than economy",
          subtitle="Percent difference from trend") +
  theme_report +
  legend_none
p3

p123 <- p1 / p2 / p3


p4 <- econ %>%
  filter(variable %in% c("econ", "taxplus"), measure=="actual") %>%
  bind_rows(econ %>% filter(variable=="taxplus", measure=="trend")) %>%
  unite("name", variable, measure) %>%
  mutate(name=factor(name, 
                     levels=c("econ_actual", "taxplus_actual", "taxplus_trend"),
                     labels=c("economy", "revenue",  "revenue trend"))) %>%  # to get right sort order
  ggplot(aes(time, value, colour=name)) +
  geom_line() +
  scale_y_continuous(name=NULL, limits=c(100, 300), labels=scales::dollar_format(scale=1)) +
  scale_x_continuous(name="time", limits=c(NA, NA), breaks=seq(0, 30, 5)) +
  scale_colour_manual(values=colors) +
  ggtitle("Revenue that grows 25% faster than economy",
          subtitle="Annual amount indexed to $100 in year 0") +
  theme_report +
  legend_notitle
p4

p5 <- econ %>%
  filter(variable %in% c("econ", "taxplus"), measure=="pch") %>%
  bind_rows(econ %>% filter(variable=="taxplus", measure=="trgrowth")) %>%
  unite("name", variable, measure) %>%
  mutate(name=factor(name, 
                     levels=c("econ_pch", "taxplus_pch", "taxplus_trgrowth"),
                     labels=c("economy", "revenue", "revenue trend"))) %>%  # to get right sort order
  ggplot(aes(time, value, colour=name)) +
  geom_line() +
  geom_hline(yintercept = 0, colour="grey") +
  scale_y_continuous(name="% change", 
                     breaks=seq(-.5, .5, .05),
                     limits=c(-.15, .15),
                     labels=scales::percent_format(accuracy=1)) +
  scale_x_continuous(name="time", breaks=seq(0, 30, 5)) +
  scale_colour_manual(values=colors) +
  ggtitle("Revenue that grows 25% faster than economy",
          subtitle="Annual percent change") +
  theme_report +
  legend_notitle
p5

# percent difference from trend
p6 <- econ %>%
  filter(variable %in% c("taxplus"), measure=="pdt") %>%
  mutate(posneg=ifelse(value < 0, "neg", "pos")) %>%
  ggplot(aes(time, value)) +
  geom_hline(yintercept = 0) +
  geom_bar(stat = "identity", aes(fill = posneg)) + 
  scale_fill_manual(values=rev(bluered)) +
  scale_x_continuous(name="time", breaks=seq(0, 30, 5)) +
  scale_y_continuous(name="% difference from trend",
                     breaks=seq(-.5, .5, .05), 
                     limits=c(-.15, .15),
                     labels=scales::percent_format(accuracy=1)) +
  ggtitle("Revenue that grows 25% faster than economy",
          subtitle="Percent difference from trend") +
  theme_report +
  legend_none
p6


p456 <- p4 / p5 / p6

p7 <- p123 | p456
p7
ggsave(filename = here::here("results", "rev_econ_elasticity_6panel.png"),
       plot = p7, width=8, height=8, scale=1.25)

```

## Volatility measures -- simple case with years 1990-2019

```{r vol_measures}
# create a volatility base file we will use to create all volatility measures

years <- 1990:2020
years <- 1970:2020
# stack: 1950-2020

vbase <- stack %>%
  arrange(stabbr, name, realnom, year) %>% 
  group_by(stabbr, name, realnom) %>%
  mutate(pch=value / value[match(year - 1, year)] - 1,
         trend=hptrend(value, smooth=6.25),
         pdtrend=value / trend - 1) %>%
  ungroup
glimpse(vbase)
summary(vbase)

# the basic measures -- over the entire period
vol_basic <- vbase %>%
  filter(year %in% years) %>%
  group_by(stabbr, realnom, name) %>%
  summarise(pchsd = sd(pch, na.rm=TRUE),
            hpsd = sd(pdtrend, na.rm=TRUE),
            .groups = "drop")
ht(vol_basic)
count(vol_basic, stabbr) # 54 incl US, VI

# now construct the more complicated measures
unique(vbase$name)
count(vbase %>% filter(name=="gdp"), realnom, stabbr) # all 50 states
a <- proc.time()
vol_other <- vbase %>%
  filter(year %in% years) %>%
  # bring in real gdp or nominal gdp
  left_join(vbase %>% filter(name=="gdp") %>% select(stabbr, realnom, year, gdp=value),
            by=c("stabbr", "realnom", "year")) %>%
  group_by(stabbr, name, realnom) %>%
  mutate(dlvalue=dl(value), dlgdp=dl(gdp)) %>%
  nest() %>%
  # note that this does not filter out the case where GDP is on the lhs and the rhs so we will get warnings for that!
  # they will arise in the next step when we unpack
  mutate(sremod = purrr::map(data, safely(function(df) lm(dlvalue ~ dlgdp, data=df))),
         epi = purrr::map(data, safely(function(df) episodes(df$pdtrend)))) %>%
  ungroup
b <- proc.time()
b - a  # 19 secs
vol_other
summary(vol_other)

# unpack these other volatility measures

# sre -- short-run elasticity -- unpack into sre and sresum
# unpack the summary measures from sremod to get coefficient (short-run elasticity, sre) and its standard error
sredf <- vol_other %>%
  select(stabbr, name, realnom, sremod) %>%
  unnest_wider(col=sremod) %>%  # we need the column "result"
  mutate(tidied=purrr::map(result, tidy)) %>%
  unnest(cols=tidied) %>%
  filter(term=="dlgdp") %>%
  select(stabbr, name, realnom, sre=estimate, srese=std.error)
# the warnings are for gdp on lhs and rhs
sredf %>% arrange(desc(sre))
sredf %>% arrange(desc(srese))


# epi -- episodes -- unpack into elength and esum
epidf <- vol_other %>%
  select(stabbr, name, realnom, epi) %>%
  unnest_wider(col=epi) %>%
  unnest_wider(col=result)
epidf

epidf %>%
  arrange(-esum)

# unpack the summary measures from sregspmod to get coefficient (short-run elasticity, sre) and its standard error
measures <- vol_basic %>%
  left_join(sredf, by = c("stabbr", "name", "realnom")) %>%
  left_join(epidf, by = c("stabbr", "name", "realnom")) %>%
  select(stabbr, name, realnom, everything())
summary(measures)

st <- "US"
st <- "NY"
st <- "CA"
measures %>%
  filter(stabbr==st, name %in% c(taxvars, "capgains")) %>%
  arrange(-pchsd)

```

## NY example of the danger of short-run elasticity

```{r include=TRUE}

f <- function(stabbr, tax, realnom, taxname){
  revdf <- stack %>%
    filter(stabbr==!!stabbr, name==!!tax, realnom==!!realnom, year %in% 1990:2019) %>%
    prep_3panel()

  econdf <- stack %>%
    filter(stabbr==!!stabbr, name=="gdp", realnom==!!realnom, year %in% 1990:2019) %>%
    prep_3panel()
  
  p <- revecon4(revdf, econdf,
                revtitle=paste0(stname(stabbr), " ", realnom, " ", taxname),
                econtitle=paste0(stname(stabbr), " real GDP"))
  p
}
p1 <- f("NY", "iit", "real", "income tax")
p1
ggsave(filename = here::here("results", "NY_iit_real_econ4.png"),
       plot = p1, width=8, height=6)

# now compute the regression
df <- stack %>%
  filter(year %in% 1990:2019, stabbr=="NY", name %in% c("iit", "gdp"), realnom=="real") %>%
  group_by(stabbr, name, realnom) %>%
  arrange(year) %>%
  mutate(lvalue=log(value),
         dlvalue=lvalue - lag(lvalue)) %>%
  ungroup %>%
  select(stabbr, name, realnom, year, dlvalue) %>%
  pivot_wider(values_from = dlvalue) %>%
  na.omit()
m1 <- lm(iit ~ gdp, data=df)
summary(m1)
# summary(m2)
tidy(m1)

years <- 1990:2019
p <- stack %>%
  filter(year %in% years, stabbr=="NY", name %in% c("iit", "gdp"), realnom=="real") %>%
  group_by(stabbr, name, realnom) %>%
  arrange(year) %>%
  mutate(pch=pchya(value, year),
         name=factor(name, levels=c("iit", "gdp"), labels=c("income tax", "state GDP"))) %>%
  ggplot(aes(year, pch, colour=name)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0) +
  scale_x_continuous(name=NULL, breaks=seq(1980, 2025, 5)) +
  scale_y_continuous(name="% change",
                     breaks=seq(-20, 20, 2)) + 
  recband(recessions %>% filter(rec_year %in% years)) +
  ggtitle("% change in real NY income tax revenue and state GDP") +
  scale_colour_manual(values=c("blue", "darkgreen")) +
  theme_report +
  legend_notitle
p 
ggsave(filename = here::here("results", "NY_real_iit_gdp.png"), plot = p, width=8, height=6)

```

## Descriptive stats on vol measures

## US totals

```{r}
usmeas <- measures %>%
  filter(stabbr=="US")

```

# ADJUSTING FOR POLICY CHANGES

## Rate adjustment

```{r get_rate_change_measures}

vtemp <- vbase %>%
  filter(name %in% c("iit", "gst", "iitadj", "gstadj")) %>%
  filter(!(stabbr=="AK" & name=="iit")) %>%
  na.omit()

vtemp %>% filter(stabbr=="CT", str_detect(name, "iitadj"))

common_years <- vtemp %>%
  group_by(name, stabbr) %>%
  summarise(minyear=min(year), maxyear=max(year), .groups="drop")

maxcombo <- common_years %>%
  # find most common minyear, most common maxyear
  group_by(name, minyear, maxyear) %>%
  summarise(n=n())
# looks like 1978-2018 is best combination to use

vtemp2 <- vtemp %>%
  filter(year %in% 1978:2018)

rate_measures <- get_measures(vtemp2)
rate_measures

count(rate_measures, stabbr) # 49 states (not AK), plus US
count(rate_measures, n) 
rate_measures %>% filter(n < 41)  # CT iit, real nominal

rate_measures %>% filter(stabbr=="US") # we do NOT have adjusted values for the US

rate_measures %>%
  group_by(realnom, name) %>%
  summarise(pchsd=median(pchsd),
            hpsd=median(hpsd), .groups="drop") %>%
  arrange(-pchsd)

```

### Raw comparisons of unadjusted and adjusted measures

```{r}
# get records where we have measures for iit and gst, actual and adjusted, real and nominal
df <- rate_measures %>%
  group_by(stabbr) %>%
  mutate(nmeas=n()) %>%
  ungroup %>%
  filter(nmeas==8) %>%
  mutate(dtype=ifelse(str_detect(name, "adj"), "adjust", "actual"),
         tax=str_sub(name, 1, 3)) %>%
  select(stabbr, realnom, tax, dtype, pchsd:esum)
df
count(df, stabbr)
df %>% filter(stabbr=="AL")

```

```{r}
df2 <- df %>%
  pivot_longer(cols = pchsd:esum, names_to = "measure") %>%
  pivot_wider(names_from = dtype) %>%
  group_by(realnom, tax, measure) %>%
  summarise(nstates=n(), across(c(actual, adjust), median), .groups="drop") %>%
  mutate(ratio=adjust / actual) %>%
  filter(realnom=="real", measure %in% meas)
count(tabdata, measure)

meas <- c("pchsd", "hpsd", "esum", "sre")

f <- function(tabdata, tax){
  tab <- tabdata %>%
  left_join(mlabels, by = "measure") %>%
  mutate(measure=factor(measure, levels=meas)) %>%
  arrange(measure) %>%
  select(mlabel, actual, adjust, ratio) %>%
  gt() %>%
    tab_header(
      title = paste0(tax, " volatility measures for median state, with and without adjustment for tax-rate changes"),
      subtitle = ""
      ) %>%
      cols_label(mlabel=html("Volatility<br>measure"),
                 actual=html("NOT adjusted for<br>tax rate<br>changes"),
                 adjust=html("Adjusted for<br>tax rate<br>changes"),
                 ratio=html("Ratio of<br>median adjusted<br>to<br>median unadjusted")) %>%
      tab_spanner(
        label = "Rate adjustment:",
        columns = c(actual, adjust)
        ) %>%
      fmt_percent(
        columns = c(actual, adjust),
        decimals = 1) %>%
      # selectively override fmt_percent
      fmt_number(
        columns = ratio,
        decimals = 2) %>%
      fmt_number(
        columns = c(actual, adjust),
        rows = c(4),
        decimals = 2) %>%
    tab_style(
      style = list(
        cell_fill(color = "lightgrey")
      ),
      locations = cells_body(
        rows=c(2, 4)
      )
    ) %>%
    tab_footnote(
      footnote = "Adjusted for inflation with GDP price index. Estimated over 1978-2018.",
      locations = cells_title(groups = "title")
      )%>%
    tab_footnote(
      footnote = "A ratio > 1 means volatility is greater after adjustment for tax-rate changes.",
      locations = cells_column_labels(columns = ratio)
      )
  tab
}

tabdata <- df2 %>%
  filter(tax=="gst")
tabgst <- f(tabdata, "General sales tax")
tabgst
gtsave(tabgst, here::here("results", "gst_adjust.png"), zoom = 1.5, expand = 10)


tabdata <- df2 %>%
  filter(tax=="iit")
tabiit <- f(tabdata, "Individual income tax")
tabiit
gtsave(tabiit, here::here("results", "iit_adjust.png"), zoom = 1.5, expand = 10)
gtsave(tabiit, here::here("results", "iit_adjust.rtf"))


```

### How do rate adjustments affect which tax is most volatile?

```{r}
df2 <- df %>%
  pivot_longer(cols = pchsd:esum, names_to = "measure") %>%
  pivot_wider(names_from = tax) %>%
  mutate(iit_gst=iit / gst) %>%
  group_by(realnom, dtype, measure) %>%
  summarise(nstates=n(), across(c(gst, iit, iit_gst), median), .groups="drop") %>%
  select(-c(gst, iit)) %>%
  pivot_wider(names_from = dtype, values_from = iit_gst) %>%
  mutate(change=adjust - actual) %>%
  filter(realnom=="real", measure %in% meas)

tabdata <- df %>%
  pivot_longer(cols = pchsd:esum, names_to = "measure") %>%
  pivot_wider(names_from = tax) %>%
  mutate(iit_gst=iit / gst) %>%
  group_by(realnom, dtype, measure) %>%
  summarise(ngst=sum(gst > iit), niit=sum(iit > gst), nstates=n(), diff=nstates - ngst - niit, .groups="drop") %>%
  mutate(pctiit=niit / nstates) %>%
  select(realnom, dtype, measure, pctiit) %>%
  pivot_wider(names_from = dtype, values_from = pctiit) %>%
  mutate(change=adjust - actual) %>%
  filter(realnom=="real", measure %in% meas)

tab <- tabdata %>%
  left_join(mlabels, by = "measure") %>%
  mutate(measure=factor(measure, levels=meas)) %>%
  arrange(measure) %>%
  select(mlabel, actual, adjust, change) %>%
  gt() %>%
    tab_header(
      title = "Percentage of income-plus-sales-tax states for which income tax is more volatile",
      subtitle = "With and without adjustment for tax-rate changes"
      ) %>%
      cols_label(mlabel=html("Volatility<br>measure"),
                 actual=html("NOT adjusted for<br>tax rate<br>changes"),
                 adjust=html("Adjusted for<br>tax rate<br>changes"),
                 change=html("Adjusted<br>minus<br>unadjusted")) %>%
      tab_spanner(
        label = "Rate adjustment:",
        columns = c(actual, adjust)
        ) %>%
      fmt_percent(
        columns = c(actual, adjust, change),
        decimals = 1) %>%
    tab_style(
      style = list(
        cell_fill(color = "lightgrey")
      ),
      locations = cells_body(
        rows=c(2, 4)
      )
    ) %>%
    tab_footnote(
      footnote = "Adjusted for inflation with GDP price index. Estimated over 1978-2018. Thirty-eight states included.",
      locations = cells_title(groups = "title")
      )
tab
gtsave(tab, here::here("results", "iit_vs_gst_adjust.png"), zoom = 1.5, expand = 10)

```

```{r}

# within state, nominal and real, which is more volatile, income or sales, under adjusted and actual?
act_v_adjust <- df %>%
  pivot_longer(cols = pchsd:esum, names_to = "measure") %>%
  pivot_wider(names_from = tax) %>%
  mutate(iit_gst_ratio=iit / gst) %>%
  select(-c(gst, iit))

# how many measures put iit as more volatile
act_v_adjust %>%
  group_by(dtype) %>%
  summarise(pctiit=sum(iit_gst_ratio > 1) / n(),
            mdnratio=median(iit_gst_ratio, na.rm=TRUE))
  


rate_measures %>%
  group_by(realnom, name) %>%
  summarise(pchsd=median(pchsd),
            hpsd=median(hpsd), .groups="drop") %>%
  arrange(-pchsd)

```

### How do rate adjustments affect which state is most volatile?

# VOLATILITY OVER TIME

```{r vol_rolling}
summary(vbase)
keepvars <- c("tottax","capgains", "gdp", "iit", "gst", "iitadj", "gstadj")
startyear <- 1960
volroll <- vbase %>%
  # filter(stabbr=="US", name=="gdp", realnom=="nominal") %>%
  filter(name %in% keepvars)  %>%
  filter(year >= startyear) %>%
  group_by(stabbr, name, realnom) %>%
  mutate(
    pchsd_5 = rollsd_p(pch, 5),
    pdtrendsd_5 = rollsd_p(pdtrend, 5),
    pchsd_10 = rollsd_p(pch, 10),
    pdtrendsd_10 = rollsd_p(pdtrend, 10),
    pchsd_15 = rollsd_p(pch, 15),
    pdtrendsd_15 = rollsd_p(pdtrend, 15),
    pchsd_20 = rollsd_p(pch, 20),
    pdtrendsd_20 = rollsd_p(pdtrend, 20)) %>%
  ungroup %>%
  # prepare to make a long file
  select(-c(trend, value)) %>%
  pivot_longer(cols=-c(year, stabbr, name, realnom),
               names_to = "vtype", values_to = "vol") %>%
  separate(vtype, into=c("measure", "volperiod"))
# ,         sre=rollsre(dltax, dlgdp, vperiod)
# warning message is ok, just relates to separate without full # of pieces


volroll %>%
  filter(name=="gdp", 
         realnom=="nominal", 
         stabbr=="US",
         measure=="pchsd") %>%
  na.omit() %>%
  ggplot(aes(year, vol, colour=volperiod)) +
  geom_line()

volroll %>%
  filter(name=="gdp", 
         realnom=="nominal", 
         stabbr=="US",
         measure=="pchsd") %>%
  na.omit() %>%
  ggplot(aes(year, vol, colour=volperiod)) +
  geom_line()

vol_roll %>%
  filter(realnom=="nominal") %>%
  na.omit() %>%
  ggplot(aes(year, pdtrendsdd, colour=name)) +
  geom_line()

vol_roll %>%
  filter(realnom=="nominal", name %in% c("iit", "capgains", "gdp")) %>%
  na.omit() %>%
  ggplot(aes(year, pdtrendsdd, colour=name)) +
  geom_line()

lmdf <- vbase %>%
  left_join(vbase %>% 
              filter(stabbr=="US", name=="capgains") %>% 
              select(realnom, year, cg_pch=pch, cg_pdtrend=pdtrend),
            by=c("realnom", "year")) %>%
  arrange(realnom, stabbr, name, year) %>%
  filter(name %in% c("gst", "iit", "gdp"), realnom=="nominal") %>%
  select(stabbr, name, year, pdtrend, cg_pdtrend) %>%
  pivot_wider(values_from = pdtrend) %>%
  arrange(stabbr, year) %>%
  na.omit()

st <- "CA"
st <- "NY"
summary(lm(iit ~ gdp + cg_pdtrend, data=lmdf %>% filter(stabbr==st)))
summary(lm(gst ~ gdp + cg_pdtrend, data=lmdf %>% filter(stabbr==st)))

lmdf %>%
  filter(stabbr=="CA") %>%
  pivot_longer(cols=c(iit, gdp, cg_pdtrend)) %>%
  ggplot(aes(year, value, colour=name)) +
  geom_line()

```

```{r fig_voltax_pchsd_periods}
pdata <- volroll %>%
  na.omit() %>%
  filter(year >= 1980,
         name=="tottax", 
         realnom=="nominal", 
         stabbr=="US",
         measure=="pchsd",
         volperiod %in% c(10, 20)) %>%
  mutate(volperiod=paste0(volperiod, " years")) %>%
  arrange(year, volperiod)

p <- pdata %>%
  ggplot(aes(year, vol, colour=volperiod)) +
  geom_line(size=1) +
  geom_point(size=1) +
  scale_x_continuous(name=NULL, breaks=seq(1950, 2030, 5)) +
  scale_y_continuous(name="volatility",
                     breaks=seq(-.20, .20, .02),
                     labels=label_percent(accuracy=.1),
                     limits=c(0, NA)) +
  scale_color_manual(values=bluered) +
  recband(recdata(min(pdata$year):max(pdata$year))) +
  ggtitle("Volatility of State Government Tax Revenue",
          subtitle="Rolling standard deviation of annual percentage change") +
  labs(colour = html("SD calculation\nperiod")) +
  theme_report
p

savefig("fig_voltax_pchsd_periods")

```

```{r fig_voltax_pdtrendsd_periods}

pdata <- volroll %>%
  na.omit() %>%
  filter(year >= 1980,
         name=="tottax", 
         realnom=="nominal", 
         stabbr=="US",
         measure=="pdtrendsd",
         volperiod %in% c(10, 20)) %>%
  mutate(volperiod=paste0(volperiod, " years")) %>%
  arrange(year, volperiod)

p <- pdata %>%
  ggplot(aes(year, vol, colour=volperiod)) +
  geom_line(size=1) +
  geom_point(size=1) +
  scale_x_continuous(name=NULL, breaks=seq(1950, 2030, 5)) +
  scale_y_continuous(name="volatility",
                     breaks=seq(-.20, .20, .02),
                     labels=label_percent(accuracy=.1),
                     limits=c(0, NA)) +
  scale_color_manual(values=bluered) +
  recband(recdata(min(pdata$year):max(pdata$year))) +
  ggtitle("Volatility of State Government Tax Revenue",
          subtitle="Rolling standard deviation of percentage difference from trend") +
  labs(colour = html("SD calculation\nperiod")) +
  theme_report
p

savefig("fig_voltax_pdtrendsd_periods")
```

# VOLATILITY AND TYPE OF TAX

# ADDITIONAL ANALYSES

### Real and nominal measures for taxes

```{r}

# how similar are real and nominal measures for a few key variables
taxvars <- c("tottax", "iit", "gst", "cit", "selsalestax", "sevtax", "othertax")

meas <- "pchsd"
meas <- "hpsd"
meas <- "sre"
usmeas %>%
  filter(name %in% taxvars) %>%
  # select(stabbr, name, realnom, measure=all_of(meas)) %>%
  pivot_longer(cols=pchsd:srese, names_to = "measure") %>%
  pivot_wider(names_from = realnom) %>%
  ggplot(aes(x=nominal, y=real, label=name)) +
  geom_point() +
  geom_text() +
  geom_abline(slope=1, intercept = 0) +
  facet_wrap(~measure, scales = "free") +
  theme_report


```

# OLD BELOW HERE

## Formal measures of volatility

In this section we compute summary volatility measures by state and tax. We summarize them over the last approximately 30 years (1990-2018) as follows:

-   Measures of total volatility (NOT controlling for the economy):
    -   annual % change: standard deviation of percent change over the 30 years
    -   \% deviation from trend: standard deviation of the % difference
-   Measures that control for the economy:
    -   national economy: coefficient on a model that estimates the log-difference in tax revenue as a function of the log-difference in national GDP (a la Sobel and Holcombe)
    -   state economy: coefficient from the same kind of model, with log-difference in state GDP on the RHS instead of national GDP
    -   uncertainty of the national model: the standard error of the coefficient on the national model; if the standard error is large, then national GDP alone is not a good predictor of state tax revenue and thus there is a lot of other variation (volatility) besides that captured by the economic measure

```{r}
vol %>%
  filter(stabbr=="NY") %>%
  arrange(-hpsd)

vol %>%
  filter(name=="gstbase") %>%
  arrange(-hpsd)

vol %>%
  filter(name %in% c("gstbase", "gst")) %>%
  arrange(-hpsd)

tmp <- vol %>%
  filter(name %in% c("gstbase", "gst")) %>%
  select(stabbr, name, hpsd) %>%
  pivot_wider(values_from = hpsd) %>%
  mutate(ratio=gst / gstbase) %>%
  arrange(-gst)
summary(tmp)


vol %>%
  filter(name %in% c("gstbase", "gst"), stabbr != "US") %>%
  select(stabbr, name, hpsd) %>%
  pivot_wider(values_from = hpsd) %>%
  ggplot(aes(x=gstbase, y=gst, label=stabbr)) +
  geom_point(colour="blue", size=0.5) +
  geom_text(colour="blue", size=2) +
  geom_abline(slope=1, intercept=0)


```

```{r vol_formal, eval=TRUE}
# prepare the data

# https://cran.r-project.org/web/packages/broom/vignettes/broom_and_dplyr.html
# http://r4stats.com/2017/04/18/group-by-modeling-in-r-made-easy/

# censustax %>% 
#   filter((stabbr == "OH" & name == "cit")) %>%
#   mutate(lvalue=log(value), slvalue=slog(value),
#          dl=dl(value), dsl=dsl(value))

vbase <- censustax %>%
  filter(!(stabbr == "DE" & name == "gst")) %>% # only 3 obs, 1951-53
  filter(!(stabbr == "OH" & name == "cit")) %>% # negative in 2014
  rename(taxtype=name, tax=value) %>%
  left_join(gdpfy, by="year") %>%
  left_join(sgdpfy %>%
              filter(name=="gdp") %>%
              rename(sgdp=value) %>%
              select(year, stabbr, sgdp),
            by=c("stabbr", "year")) %>%
  arrange(stabbr, taxtype, year) %>% 
  group_by(stabbr, taxtype) %>%
  mutate(taxpch=tax / tax[match(year - 1, year)] * -1,
         dltax=dl(tax),
         dlgdp=dl(gdp),
         dlsgdp=dl(sgdp),
         trend=hptrend(tax, smooth=6.25))%>%
  ungroup

summary(vbase)  # 1951-2020
vbase %>% filter(!is.na(dlgdp)) %>% summary() # 1952-2019
vbase %>% filter(!is.na(dlsgdp)) %>% summary() # 1965-2019

vol <- vbase %>%
  filter(year >= 1990) %>%
  group_by(stabbr, taxtype) %>%
  nest() %>%
  mutate(pchsd = map_dbl(data, function(df) sd(df$taxpch, na.rm=TRUE)),
         hpsd = map_dbl(data, function(df) sd(df$tax / df$trend -1, na.rm=TRUE)),
         sremod = purrr::map(data, safely(function(df) lm(dltax ~ dlgdp, data=df))),
         sregspmod = purrr::map(data, safely(function(df) lm(dltax ~ dlsgdp, data=df)))) %>%
  ungroup
vol
summary(vol)


# sre -- short-run elasticity
# unpack the summary measures from sremod to get coefficient (short-run elasticity, sre) and its standard error
sredf <- vol %>%
  select(stabbr, taxtype, sremod) %>%
  unnest_wider(col=sremod) %>%  # we need the column "result"
  mutate(tidied=purrr::map(result, tidy)) %>%
  unnest(cols=tidied) %>%
  filter(term=="dlgdp") %>%
  select(stabbr, taxtype, sre=estimate, srese=std.error)

# unpack the summary measures from sregspmod to get coefficient (short-run elasticity, sre) and its standard error
sregspdf <- vol %>%
  select(stabbr, taxtype, sregspmod) %>%
  unnest_wider(col=c(sregspmod)) %>%  # we need the column "result"
  mutate(tidied=purrr::map(result, tidy)) %>%
  unnest(cols=tidied) %>%
  filter(term=="dlsgdp") %>%
  select(stabbr, taxtype, sregsp=estimate, sregspse=std.error)

measures <- vol %>%
  select(stabbr, taxtype, pchsd, hpsd) %>%
  left_join(sredf, by = c("stabbr", "taxtype")) %>%
  left_join(sregspdf, by = c("stabbr", "taxtype"))

# get all relevant correlations
corr_types <- measures %>% # measures must be ungrouped so that we can drop stabbr!
  filter(stabbr != "US") %>%
  select(-stabbr) %>%  
  group_by(taxtype) %>%
  nest() %>%
  summarise(map_df(data, corr_tidy), .groups="drop")  %>%
  unite("pair", column1, column2, remove=FALSE)
corr_types

corr_types %>%
  select(taxtype, pair, estimate) %>%
  pivot_wider(names_from = pair, values_from = estimate)

corr_states <- measures %>%
  select(-taxtype) %>%  # measures must be ungrouped...
  group_by(stabbr) %>%
  nest() %>%
  # only include columns we want in the correlation
  summarise(map_df(data, corr_tidy), .groups="drop") %>%  # note this will produce some NaNs
  unite("pair", column1, column2, remove=FALSE)
# corr_states

```

## The measures

First we look at the measures for the nation (excluding the state GDP measure), by tax.

A few observations:

-   By all measures, severance taxes are the most volatile, followed by corporate income taxes, personal income, and general sales taxes.
-   Selected sales taxes are the least volatile.
-   The portfolio of taxes -- total taxes -- is less volatile than the income tax but more volatile than sales taxes.

```{r eval=TRUE, include=TRUE}
# hpsd_pchsd	sre_pchsd	sre_hpsd	srese_pchsd	srese_hpsd	srese_sre
vorder <- c("tottax", "iit", "gst", "cit", 
            "selsalestax", "sevtax", "othertax")
vlabs = c("Total taxes", 
          "Individual income tax", "General sales tax", "Corporate income tax", 
          "Selected sales taxes", "Severance taxes", "Other taxes")

tabdata <- measures %>%
  filter(stabbr=="US") %>%
  select(taxtype, pchsd, hpsd, sre, srese) %>%
  mutate(taxtype=factor(taxtype, levels=vorder, labels=vlabs)) %>%
  arrange(taxtype)
shade_rows <- seq(2, nrow(tabdata), 2)

tabdata %>%
  gt() %>%
    tab_header(
      title = "Selected volatility measures by tax type for the U.S. as a whole",
      subtitle = "Estimated over 1990-2018"
      ) %>%
      cols_label(taxtype="Type of tax",
                 pchsd=html("Annual % change<br>(Standard deviation)"),
                 hpsd=html("% Deviation from trend<br>(Standard deviation)"),
                 sre=html("Short-run elasticity<br>(SRE)"),
                 srese=html("Standard error<br>of SRE")) %>%
      cols_align(
        align = c("left"),
        columns = taxtype) %>%
      fmt_number(
        columns = -contains("tax"),
        decimals = 2) %>%
    tab_style(
      style = list(
        cell_fill(color = "lightgrey")
      ),
      locations = cells_body(
        rows=shade_rows
      )
    )

```

## How correlated are the different measures, by tax?

Next we look at how correlated the measures are. To do this, we take each summary measure for each tax in each state, and compute the correlation of these measures. For example, the table shows that the percent-change and deviation-from-trend measures have a 0.818 correlation for the income tax. This suggests that when the percent-change volatility measure is high for the income tax in a given state, the deviation-from-trend measure is usually high in the state as well.

A few observations:

-   The two measures that do NOT control for the economy -- that is, annual percent change, and % deviation from trend - are highly correlated for almost all taxes (1st column of numbers). However, the correlation is very low for severance taxes. That suggests how states think about measures here is particularly important since volatility may be high by one measure but not for another.
-   The measure that controls for the economy (short-run elasticity -- SRE) is not highly correlated with the two "gross" or total volatility measures (columns 2 and 3). Clearly they measure different things. If budget policymakers care about volatility, wherever it comes from, then short-run elasticity is not enough.
-   The standard error of the SRE -- a measure of how well you can predict revenue changes when you know how the economy changes -- is highly correlated with our two measures of total volatity. In other words, if you have a lot of volatility above and beyond what the economy suggests (a higher standard error), then you likely have a lot of total volatility. Another indication that if you care about total volatility, the SRE is not enough.

```{r eval=TRUE, include=TRUE}

# hpsd_pchsd	sre_pchsd	sre_hpsd	srese_pchsd	srese_hpsd	srese_sre
vorder <- c("tottax", "iit", "gst", "cit", 
            "selsalestax", "sevtax", "othertax")
vlabs = c("Total taxes", 
          "Individual income tax", "General sales tax", "Corporate income tax", 
          "Selected sales taxes", "Severance taxes", "Other taxes")
morder <- c("hpsd_pchsd", "sre_pchsd", "sre_hpsd",
            "srese_sre", "srese_hpsd", "srese_pchsd")

tabdata <- corr_types %>%
  filter(pair %in% morder) %>% 
  select(taxtype, pair, estimate) %>%
  mutate(taxtype=factor(taxtype, levels=vorder, labels=vlabs),
         pair=factor(pair, levels=morder)) %>%
  arrange(taxtype, pair) %>%
  pivot_wider(names_from = pair, values_from = estimate)
shade_rows <- seq(2, nrow(tabdata), 2)
# names(tabdata)

tabdata %>%
  gt() %>%
    tab_header(
      title = "Correlation of selected volatility measures by tax type",
      subtitle = "Estimated over 1990-2018"
      ) %>%
      cols_label(taxtype="Type of tax",
                 hpsd_pchsd=html("Percent change<br>-- correlation with --<br>Deviation from trend"),
                 sre_pchsd=html("Percent change<br>-- correlation with --<br>Short-run elasticity (SRE)"),
                 sre_hpsd=html("Deviation from trend<br>-- correlation with --<br>Short-run elasticity (SRE)"),
                 srese_sre=html("Short-run elasticity (SRE)<br>-- correlation with --<br>Standard error of SRE"),
                 srese_pchsd=html("Percent change<br>-- correlation with --<br>Standard error of SRE"),
                 srese_hpsd=html("Deviation from trend<br>-- correlation with --<br>Standard error of SRE")) %>%
      cols_align(
        align = c("left"),
        columns = taxtype) %>%
      fmt_number(
        columns = contains("_"),
        decimals = 3) %>%
    tab_style(
      style = list(
        cell_fill(color = "lightgrey")
      ),
      locations = cells_body(
        rows=shade_rows
      )
    )
	
```

## Scatterplots of measures

Let's visualize some of these conclusions:

```{r eval=TRUE}
# get index by type of tax, for each state
count(measures, stabbr)
scatdata <- measures %>%
  filter(stabbr != "US") %>%
  select(-c(srese, sregspse)) %>%
  group_by(taxtype) %>%
  mutate(across(-c(stabbr), ~ .x / median(.x, n.rm=TRUE)))
scatdata  
```

### Relationship between the total volatility measures

```{r eval=TRUE, include=TRUE}
scatdata %>%
  filter(taxtype=="tottax") %>%
  filter(stabbr != "AK") %>%
  ggplot(aes(x=pchsd, y=hpsd, label=stabbr)) +
   geom_point(colour="blue", size=0.5) +
   geom_text(colour="blue", size=2) +
   geom_abline(slope=1) +
   labs(x="Standard deviation of % change",
        y="Standard deviation of % deviation from trend") +
   ggtitle("Relationship between the two total volatility measures, total taxes",
           subtitle="Indexed so that state median=100; Alaska excluded")

```

### Relationship between total volatility measure and short-run elasticity (controlling for national economy)

Observations:

-   States to the right have a lot of total volatility.
-   States that are high have a lot of volatility beyond what we can see from the national economy.
-   States that are low (below the 45-degree line) and far to the right have a lot of total volatility, but much of it is explained by the national economy.

```{r eval=TRUE, include=TRUE}
scatdata %>%
  filter(taxtype=="tottax") %>%
  filter(stabbr != "AK") %>%
  ggplot(aes(x=hpsd, y=sre, label=stabbr)) +
   geom_point(colour="blue", size=0.5) +
   geom_text(colour="blue", size=2) +
   geom_abline(slope=1) +
   labs(x="Standard deviation of % deviation from trend",
        y="Short run elasticity") +
   ggtitle("Relationship between short-run elasticity (controlling for national GDP) and total volatility (trend deviation), total taxes",
           subtitle="Indexed so that state median=100; Alaska excluded")

```

### Relationship between short-run elasticity controlling for national economy and controlling for state economy

```{r eval=TRUE, include=TRUE}

scatdata %>%
  filter(taxtype=="tottax") %>%
  filter(stabbr != "AK") %>%
  ggplot(aes(x=sre, y=sregsp, label=stabbr)) +
   geom_point(colour="blue", size=0.5) +
   geom_text(colour="blue", size=2) +
   geom_abline(slope=1) +
   labs(x="Short run elasticity - national economy",
        y="Short run elasticity - state economy") +
   ggtitle("Relationship between short-run elasticity controlling for state economy vs. national economy, total taxes",
           subtitle="Indexed so that state median=100; Alaska excluded")

```

```{r include=TRUE}
## Correlation of volatility measures that control for national or state GDP
count(corr_types, pair)
# hpsd_pchsd	sre_pchsd	sre_hpsd	srese_pchsd	srese_hpsd	srese_sre
vorder <- c("tottax", "iit", "gst", "cit", 
            "selsalestax", "sevtax", "othertax")
vlabs = c("Total taxes", 
          "Individual income tax", "General sales tax", "Corporate income tax", 
          "Selected sales taxes", "Severance taxes", "Other taxes")
morder <- c("sregsp_sre", "srese_sre", "sregspse_sregsp", "sregspse_srese")


tabdata <- corr_types %>%
  unite("pair", column1, column2) %>%
  filter(pair %in% morder) %>% 
  select(taxtype, pair, estimate) %>%
  mutate(taxtype=factor(taxtype, levels=vorder, labels=vlabs),
         pair=factor(pair, levels=morder)) %>%
  arrange(taxtype, pair) %>%
  pivot_wider(names_from = pair, values_from = estimate)
shade_rows <- seq(2, nrow(tabdata), 2)
names(tabdata)

# sregsp_sre", "srese_sre", "sregspse_sregsp", "sregspse_sre
tabdata %>%
  gt() %>%
    tab_header(
      title = "Correlation of selected volatility measures by tax type",
      subtitle = "Estimated over 1990-2018"
      ) %>%
      cols_label(taxtype="Type of tax",
                 sregsp_sre=html("Short-run elasticity (SRE)<br>-- with --<br>State-specific GDP SRE"),
                 srese_sre=html("Short-run elasticity (SRE)<br>-- with --<br>Its standard error"),
                 sregspse_sregsp=html("State-specific GDP SRE<br>-- with --<br>Its standard error"),
                 sregspse_srese=html("Short-run elasticity (SRE)<br>-- with --<br>Standard error of state-specific GDP SRE")) %>%
      cols_align(
        align = c("left"),
        columns = taxtype) %>%
      fmt_number(
        columns = contains("_"),
        decimals = 3) %>%
    tab_style(
      style = list(
        cell_fill(color = "lightgrey")
      ),
      locations = cells_body(
        rows=shade_rows
      )
    )

```

```{r include=FALSE, eval=FALSE}
## State maps of volatility
us_states <- map_data("state")

count(us_states, region)  # 49, including DC!

head(us_states)
p <- ggplot(data = us_states,
            mapping = aes(x = long, y = lat,
                          group = group, fill = region))

p + geom_polygon(color = "gray90", size = 0.1) +
    coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
    guides(fill = "none")


```

```{r include=FALSE, eval=FALSE}
## volatility by tax
# vranks <- vol %>%
#   filter(!(stabbr=="AK" & taxtype=="iit")) %>%
#   group_by(taxtype) %>%
#   select(stabbr, taxtype, pchsd, hpsd) %>%
#   mutate(rpchsd = rank(desc(pchsd)),
#          rhpsd = rank(desc(hpsd)))

```

## Total taxes for the U.S.

Observations:

-   Both kinds of measures show an increase in volatility for the U.S. as a whole over the last 20 years.
-   The measure controlling for the economy declined after 2006.
-   The total volatility measure declined after 2013.
-   Both remain far higher than the 1999-prior period.
-   It would be good to add more history to this.

```{r include=TRUE, eval=TRUE}

st <- "US"
tax <- "tottax"
vol_roll %>%
  select(stabbr, taxtype, year, hpsd, sre) %>%
  filter(year >= 1985) %>%
  filter(stabbr==st, taxtype==tax) %>%
  pivot_longer(cols=c(hpsd, sre)) %>%
  mutate(name=factor(name, 
                     levels=c("hpsd", "sre"),
                     labels=c("% deviation\nfrom trend",
                              "short-run elasticity\n(national)"))) %>%
  group_by(stabbr, taxtype, name) %>%
  mutate(ivalue=value / value[year==1999] * 100) %>%
  ggplot(aes(year, ivalue, colour=name)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 100) +
  scale_colour_manual(values=c("blue", "red")) +
  labs(x=NULL, y="Index (1999=100)") +
  scale_y_continuous(limits=c(0, NA)) +
  ggtitle("Volatility measures over time, U.S. as a whole, total taxes",
          subtitle="Indexed to 1999=100") +
  theme_report +
  legend_notitle


```

## Volatility by tax, for the U.S.

```{r eval=TRUE, include=TRUE, fig.width=10}

group1 <- c("tottax", "iit", "gst", "cit")
group2 <- c("tottax", "selsalestax", "othertax", "sevtax")

st <- "US"
p1 <- vol_roll %>%
  select(stabbr, taxtype, year, hpsd, sre) %>%
  filter(year >= 1985) %>%
  filter(stabbr==st, taxtype %in% group1) %>%
  mutate(taxtype=factor(taxtype, levels=group1)) %>%
  arrange(taxtype) %>%
  group_by(stabbr, taxtype) %>%
  mutate(ivalue=hpsd / hpsd[year==1999] * 100) %>%
  ggplot(aes(year, ivalue, colour=taxtype)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 100) +
  # scale_colour_manual(values=c("blue", "red")) +
  labs(x=NULL, y="Index (1999=100)") +
  scale_y_continuous(limits=c(0, NA)) +
  ggtitle("Deviation-from-trend volatility over time, U.S. as a whole, selected major taxes",
          subtitle="Indexed to 1999=100") +
  theme_report +
  legend_notitle
# p1

p2 <- vol_roll %>%
  select(stabbr, taxtype, year, hpsd, sre) %>%
  filter(year >= 1985) %>%
  filter(stabbr==st, taxtype %in% group2) %>%
  mutate(taxtype=factor(taxtype, levels=group2)) %>%
  arrange(taxtype) %>%
  group_by(stabbr, taxtype) %>%
  mutate(ivalue=hpsd / hpsd[year==1999] * 100) %>%
  ggplot(aes(year, ivalue, colour=taxtype)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 100) +
  # scale_colour_manual(values=c("blue", "red")) +
  labs(x=NULL, y="Index (1999=100)") +
  scale_y_continuous(limits=c(0, NA)) +
  ggtitle("Deviation-from-trend volatility over time, U.S. as a whole, other taxes",
          subtitle="Indexed to 1999=100") +
  theme_report +
  legend_notitle
# p2

p1 + p2


```

## Volatility over time by state, total taxes

Observations:

-   The volatility increase was widespread.
-   But far bigger in some places than others.
-   Some significant exeptions.
-   Not declining everywhere
-   I need a more-informative visualization.

```{r eval=TRUE, include=TRUE, fig.width=10}

p <- vol_roll %>%
  filter(year >= 1985) %>%
  filter(stabbr != "US", taxtype=="tottax") %>%
  select(stabbr, taxtype, year, hpsd) %>%
  left_join(stcodes %>% select(stabbr, beargn.name, cenrgn.name)) %>%
  arrange(beargn.name, stabbr) %>%
  group_by(stabbr, taxtype) %>%
  mutate(ivalue=hpsd / hpsd[year==1999] * 100) %>%
  ggplot(aes(year, ivalue, colour=stabbr)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 100) +
  labs(x=NULL, y="Index (1999=100)") +
  scale_y_continuous(limits=c(0, NA)) +
  ggtitle("Deviation-from-trend volatility over time by state",
          subtitle="Indexed to 1999=100") +
  theme_report +
  legend_notitle +
  facet_wrap(~beargn.name, scales="fixed", ncol=3)

p

```

# Impact of policy changes

It is better, conceptually, to adjust for policy changes. But how much difference does it make?

```{r}
glimpse(pewtax)
count(pewtax, year) # 1994-2014
count(pewtax, taxtype) # 14, including cit, gst, iit, mft, tottax
count(pewtax, vartype) # adjusted, policy, raw
# value is percent change, nominal
summary(pewtax)

```

## Exploratory look at the impact of adjusting for policy changes

### Percent change in adjusted and unadjusted revenue, selected states

```{r eval=TRUE, include=TRUE, fig.width=8, fig.height=8}
# data(package="bdata")
#    beargn  beargn.name          n
#    <chr>   <chr>            <int>
#  1 ""      ""                   2
#  2 "fwr"   "Far West"           6
#  3 "glr"   "Great Lakes"        5
#  4 "mdatl" "Mid Atlantic"       6
#  5 "neng"  "New England"        6
#  6 "plr"   "Plains"             7
#  7 "rmr"   "Rocky Mountain"     5
#  8 "ser"   "Southeast"         12
#  9 "swr"   "Southwest"          4
# 10 "usr"   "United States"      1
# find the state with the largest tax revenue in each region, by type of tax

states_ranked <- censustax %>%
  filter(year==2019, stabbr %in% state.abb) %>%
  left_join(stcodes %>% select(stabbr, beargn, beargn.name)) %>%
  group_by(name, beargn) %>%
  arrange(desc(value)) %>%
  mutate(rank=row_number()) %>%
  ungroup

f <- function(tax, taxlabel){
  p <- pewtax %>%
        filter(taxtype==tax) %>%
        select(stabbr, vartype, year, pch) %>%
        left_join(states_ranked %>%
        filter(name==tax) %>%
        select(stabbr, beargn.name, rank), by="stabbr") %>%
        group_by(beargn.name) %>%
        filter(rank==min(rank)) %>%
        ungroup %>%
        mutate(vartype=factor(vartype, 
                              levels=c("raw", "adjusted", "policy"),
                              labels=c("actual", "adjusted", "policy change")),
               stname=stname(stabbr),
               stregion=paste0(beargn.name, ": ", stname)) %>%
        arrange(beargn.name, vartype) %>%
        ggplot(aes(year, pch, colour=vartype)) +
        geom_point() +
        geom_line() +
        scale_colour_manual(values=colors) +
        scale_y_continuous(name="% change", breaks=seq(-1, 1, .05),
                           labels=scales::percent_format(accuracy = 1), limits=c(NA, NA)) +
        scale_x_continuous(name=NULL, breaks=seq(1950, 2030, 5)) +
        geom_hline(aes(yintercept=0)) +
        facet_wrap(~stregion, ncol=2, scales="free_y") +
        ggtitle(paste0(taxlabel, " revenue: Percent change in actual versus policy-adjusted revenue"),
                subtitle="Largest state in each region with data for this revenue source") +
        labs(caption="Source: Data provided by Pew Charitable Trusts") +
        theme_report +
        legend_notitle
    
        ggsave(filename = here::here("results", paste0("policy_", tax, "_selstates.png")), 
               plot = p, 
               width=8, height=8)
        p
}

colors <- c("blue", "red", "grey")


f("tottax", "Total tax")
f("iit", "Individual income tax")
f("gst", "General sales tax")
f("cit", "Corporate income tax")
# f("mft", "Motor fuel tax")
  

```

## FIX THIS: Specific volatility measures, with and without policy changes

```{r test}
# construct levels from the pewtax data
# pewtax

pewtax %>% filter(stabbr=="NY", taxtype=="tottax", vartype=="raw")


vbase <- pewtax %>%
  filter(vartype != "policy") %>%
  left_join(sgdpfy %>%
              filter(name=="gdp") %>%
              rename(sgdp=value) %>%
              select(year, stabbr, sgdp),
            by=c("stabbr", "year")) %>%
  arrange(stabbr, taxtype, vartype, year) %>% 
  group_by(stabbr, taxtype, vartype) %>%
  mutate(pch=pch * 100,
         dlvalue=dl(level),
         dlsgdp=dl(sgdp),
         trend=hptrend(level, smooth=6.25)) %>%
  ungroup


# construct volatility measures for each state, raw and adjusted
vol <- pewtax %>%
  filter(vartype != "policy") %>%
  left_join(sgdpfy %>%
              filter(name=="gdp") %>%
              rename(gdp=value) %>%
              select(year, stabbr, gdp),
            by=c("stabbr", "year")) %>%
  arrange(stabbr, taxtype, vartype, year) %>% 
  group_by(stabbr, taxtype, vartype) %>%
  mutate(pch=pch * 100,
         dlvalue=dl(level),
         dlgdp=dl(gdp)) %>%
  ungroup %>%
  group_nest(stabbr, taxtype, vartype) %>%
  mutate(hpf=map2(data, "level", safely(hpdf)),
         sre=map(data, safely(sre)))
glimpse(vol)


# analyze several measures
vol2 <- vol %>%
  hoist(data,
        year="year",
        level="level",
        pch="pch") %>%
  select(-data) %>%
  hoist(hpf, trend=c("result", "trend")) %>%
  select(-hpf) %>%
  unnest(cols=c(year, level, pch, trend))


txs <- c("tottax", "iit", "gst", "mft", "cit", "proptax")
txs <- c("tottax", "iit", "gst", "cit")

 vol %>%
   select(stabbr, taxtype, vartype, data) %>%
   unnest(data) %>%
   group_by(stabbr, taxtype, vartype) %>%
   summarise(pchsd=sd(pch, na.rm=TRUE), .groups="drop") %>%
   pivot_wider(names_from = vartype, values_from = pchsd) %>%
   filter(taxtype %in% txs) %>%
   ggplot(aes(x=raw, y=adjusted, label=stabbr)) +
   geom_point(colour="blue", size=0.5) +
   geom_text(colour="blue", size=2) +
   geom_abline(slope=1) +
        legend_notitle +
   facet_wrap(~taxtype, ncol=2, scales = "free") +
   theme_report
 

glimpse(vol2) 
vmeasures <-  vol2 %>%
  mutate(pdtrend=level / trend * 100 - 100) %>%
  group_by(stabbr, taxtype, vartype) %>%
  summarise(pchsd=sd(pch, na.rm=TRUE),
            pdtrendsd=sd(pdtrend, na.rm=TRUE),
            sre=first(sre),
            .groups="drop") %>%
  hoist(sre, sre_coeff=c("result", "coeff")) %>%
  select(-sre)%>%
  pivot_longer(-c(stabbr, taxtype, vartype), names_to = "measure") %>%
  group_by(taxtype, vartype, measure) %>%
  arrange(value) %>%
  mutate(rank=row_number())


txs <- c("tottax", "iit", "gst", "cit")
vranks <- vmeasures %>%
  filter(taxtype %in% txs) %>%
  select(-value) %>%
  pivot_wider(names_from = vartype,
              values_from = rank)
count(vranks, measure)
  

vranks %>%
  group_by(taxtype, measure) %>%
  summarise(cval=cor(raw, adjusted,
                     use="pairwise.complete.obs", method="spearman")) %>%
  pivot_wider(names_from = measure, values_from = cval)



vranks %>%
  filter(taxtype=="tottax", measure=="pdtrendsd") %>%
  ungroup %>%
  select(raw, adjusted) %>%
  correlate()


# Orange %>% 
#   group_by(Tree) %>%
#   summarize(correlation = cor(age, circumference))
#  
# library(corrr)
# d %>% 
#   correlate() %>% 
#   focus(mpg:drat, mirror = TRUE) %>% 
#   network_plot()
 
#  
# df3 <-  df2 %>%
#   # select(-sre) %>%
#   unnest_wider(hpf) %>%
#   unnest_wider(result) %>%
#   unnest(c(data, trend)) %>%
#   select(stabbr, taxtype, vartype, year, value, pch, trend, sre) %>%
#   mutate(pdtrend=value / trend * 100 - 100) 

df4 <- df3 %>%
  pivot_wider(names_from = vartype, values_from = pdtrend) %>%
  mutate(change = adjusted - raw)

changes <- df3 %>%
  group_by(stabbr, taxtype, vartype) %>%
  summarise(pchsd=sd(pch, na.rm=TRUE),
            pdtrendsd=sd(pdtrend, na.rm=TRUE), .groups="drop")

changes %>%
  select(-pchsd) %>%
  pivot_wider(names_from = vartype, values_from = pdtrendsd) %>%
  mutate(change=adjusted - raw) %>%
  arrange(-abs(change))

 
df3 %>%
   group_by(taxtype) %>%
   summarise(n=n(),
             raw=median(raw, na.rm=TRUE),
             adjusted=median(adjusted, na.rm=TRUE),
             change=median(change, na.rm=TRUE)) %>%
   filter(n > 9) %>%
   arrange(desc(change))

df2 %>%
  filter(stabbr=="NY", taxtype=="tottax") %>%
  arrange(year)

df3 %>%
  filter(stabbr=="NY", taxtype=="tottax") %>%
  arrange(year)
  
 
 
 changes <- df2 %>%
   select(stabbr, taxtype, vartype, data) %>%
   unnest(data) %>%
   group_by(stabbr, taxtype, vartype) %>%
   summarise(pchsd=sd(pch, na.rm=TRUE), .groups="drop") %>%
   pivot_wider(names_from = vartype, values_from = pchsd) %>%
   mutate(change=adjusted - raw)
 
 changes %>%
   group_by(taxtype) %>%
   summarise(n=n(),
             raw=median(raw, na.rm=TRUE),
             adjusted=median(adjusted, na.rm=TRUE),
             change=median(change, na.rm=TRUE)) %>%
   filter(n > 9) %>%
   arrange(desc(change))
 
 
 #  hoist(data, coeff=c("result", "coeff"))

sredf <- df2 %>%
  hoist(sre, coeff=c("result", "coeff")) %>% # reach within result to get coeff
  select(stabbr, taxtype, vartype, coeff) %>%
  group_by(taxtype, vartype) %>%
  arrange(desc(coeff)) %>%
  mutate(srank=row_number()) %>%
  ungroup

count(sredf, taxtype)

txs <- c("tottax", "iit", "gst", "mft", "cit", "proptax")
txs <- c("tottax", "iit", "gst", "cit")

sredf %>%
  select(-srank) %>%
  pivot_wider(names_from = vartype, values_from = coeff) %>%
  # filter(taxtype=="tottax") %>%
  filter(taxtype %in% txs) %>%
  ggplot(aes(x=raw, y=adjusted, label=stabbr)) +
  geom_point(colour="blue", size=0.5) +
  geom_text(colour="blue", size=2) +
  geom_abline(slope=1) +
        legend_notitle +
  facet_wrap(~taxtype, ncol=2, scales = "free") +
  theme_report

txs <- c("tottax", "iit", "gst", "mft", "cit", "proptax")
txs <- c("tottax", "iit", "gst", "cit")
sredf %>%
  select(-coeff) %>%
  pivot_wider(names_from = vartype, values_from = srank) %>%
  filter(taxtype %in% txs) %>%
  ggplot(aes(x=raw, y=adjusted, label=stabbr)) +
  geom_point(colour="blue", size=0.5) +
  geom_text(colour="blue", size=2) +
  geom_abline(slope=1) +
        legend_notitle +
  scale_y_reverse() +
  scale_x_reverse() +
  facet_wrap(~taxtype, ncol=2, scales = "free") +
  theme_report
    
# ggsave(filename = here::here("results", paste0("policy_", tax, "_scatter.png")), 
#        plot = p, 
#        width=8, height=8)

sredf %>%
  pivot_wider(names_from = vartype, values_from = coeff)

# which state had the largest change in volatility??
txs <- c("tottax", "iit", "gst", "mft", "cit", "proptax", "sevtax")
txs <- c("tottax", "iit", "gst", "cit", "sevtax")
changes %>%
  filter(taxtype %in% txs) %>%
  arrange(desc(abs(change))) %>%
  filter(row_number() < 15)


```

# Volatility measures and economic structure

```{r}
opts_chunk$get()
# check
measures  # 326 measures -- 7 tax types, 51 states, includes US, not DC
count(measures, taxtype) 
count(measures, stabbr)

```

```{r regs, eval=TRUE, include=TRUE}
# measures
#  tax shares
#  wage shares
#  pop size

# get average wage shares
qshares <- qcew_shares %>%
  filter(year %in% 1990:2020) %>%
  group_by(stabbr, name) %>%
  summarise(across(construct:total, ~ median(.x, na.rm=TRUE)), .groups="drop")

tshares_avg <- taxshares %>%
  filter(year %in% 1990:2020) %>%
  group_by(stabbr) %>%
  summarise(across(-c(year), ~ median(.x, na.rm=TRUE)))

pop_avg <- spop.a %>%
  filter(year %in% 1990:2020) %>%
  group_by(stabbr) %>%
  summarise(pop=median(value), .groups="drop") %>%
  mutate(lpop=log(pop))

df <- measures %>%
  left_join(qshares %>% filter(name=="emp"), by="stabbr") %>%
  left_join(tshares_avg %>% select(-tottax), by="stabbr") %>%
  left_join(pop_avg, by="stabbr")
  

df2 <- df %>% filter(taxtype=="tottax")

frm <- hpsd ~ finance + manuf + iit + cit + sevtax
rhs <- "finance + manuf + iit + cit + sevtax"
rhs <- "finance + manuf + iit + cit + sevtax + lpop"
rhs <- "log(finance) + log(manuf) + log(iit + 1) + log(cit + 1) + log(sevtax + 1) + log(pop)"
rhs <- "log(finance) + log(manuf) + iit + cit + sevtax + lpop"

frm <- paste0("hpsd", " ~ ", rhs)
frm <- as.formula(paste0("hpsd", "~", rhs))
terms(frm)

f <- function(lhs, rhs) paste0(lhs, " ~ ", rhs)  # formula as string

mod <- lm(hpsd ~ finance + manuf + iit + cit + sevtax, data=df2)  # finance no
mod <- lm(frm, data=df2)  # finance no
mod <- lm(f("hpsd", rhs), data=df2)  # finance no
mod <- lm(f("pchsd", rhs), data=df2)  # finance no
summary(mod)

mod <- lm(pchsd ~ finance + manuf, data=df2)  # finance yes
summary(mod)

mod <- lm(sre ~ finance + manuf, data=df2)  # finance no
mod <- lm(f("sre", rhs), data=df2)  # finance no
summary(mod)

mod <- lm(srese ~ finance + manuf, data=df2)  # no
summary(mod)


df2 <- df %>% filter(taxtype=="sevtax")
mod <- lm(pchsd ~ finance + manuf + mining, data=df2)  # manuf mining neg sign
summary(mod)




```

# Capital gains

```{r eval=TRUE}
# % change in gdp, tax revenue, cg
cgtax <- gdpfy %>%
  full_join(capgains,
            by="year") %>%
  full_join(censustax %>%
              filter(name %in% c("tottax", "iit", "gst")) %>%
              rename(tax=value),
            by="year") %>%
  drop_na() %>%
  group_by(stabbr, name) %>%
  arrange(year) %>%
  mutate(gdp_pch=pchya(gdp, year),
         cg_pch=pchya(capgains, year),
         tax_pch=pchya(tax, year),
         cggdp=capgains / gdp * 100) %>%
  ungroup
  
```

```{r gdpcgtax-plot}
cgtax %>%
  filter(stabbr=="US") %>%
  select(year, gdp_pch, cg_pch, tax_pch) %>%
  pivot_longer(-year) %>%
  ggplot(aes(year, value, colour=name)) +
    geom_point() +
    geom_line() +
    geom_hline(yintercept = 0) +
  ggtitle("Percent change in GDP, capital gains, and tax revenue for the United States",
          subtitle = "Not adjusted for inflation")
  
```

```{r cggdp-plot, eval=TRUE, include=TRUE, fig.height=6, fig.width=6}
# fix to use calendar year??
rec <- recessions %>%
  filter(rec_year > 1953)

anno <- cgtax %>%
  filter(stabbr=="US") %>%
  mutate(cggdp=cggdp / 100)

p <- cgtax %>%
  filter(stabbr=="US") %>%
  select(year, cggdp) %>%
  ggplot(aes(year, cggdp / 100)) +
  geom_point(colour="blue") +
  geom_line(colour="blue") +
  scale_y_continuous(name="% of GDP", labels=scales::percent_format(accuracy = .1), limits=c(0, NA)) +
  scale_x_continuous(name=NULL, breaks=seq(1950, 2030, 5)) +
  gband(rec$peak_decimal, rec$trough_decimal) +
  geom_hline(aes(yintercept=median(cggdp, na.rm=TRUE) / 100)) +
  annotate("text", x=1955, y=median(anno$cggdp, na.rm=TRUE) +.001, label = "median", size=2.5) +
  annotate("text", x=1987, y=.0765, label = "Federal\ntax reform", size=2.5) +
  annotate("text", x=2000.25, y=.0675, label = "Dot-com\nbubble", size=2.5) +
  annotate("text", x=2007.25, y=.0685, label = "Real estate\nbubble", size=2.5) +
  ggtitle("Capital gains as a percent of GDP",
          subtitle = "Recession periods are shaded") +
  labs(caption="Horizontal line marks median for the period shown") +
  theme_report
p
ggsave(filename = here::here("results", "cggdp_us.png"), plot = p, width=8, height=6)


# theme(
#   plot.title = element_text(),
#   plot.subtitle.title = element_text(),
#   plot.caption = element_text()
# )

# p + theme(
#   plot.caption = element_text(hjust = 0)
#   )

  
```

# Construct Seegert volatility measure

```{r}
df <- censustax %>%
  filter(stabbr=="US", name=="tottax", year >= 1959) %>%
  select(year, tax=value) %>%
  left_join(gdpfy, by="year") %>% # nominal gdp
  arrange(year)

df2 <- df %>%
  pivot_longer(-year) %>%
  group_by(name) %>%
  arrange(year) %>%
  mutate(trend=hptrend(value, smooth=100),
         vol1=log(abs(value - trend)),
         vol2=(value - trend) / trend * 100)

df2 %>%
  filter(year >= 1970) %>%
  ggplot(aes(year, vol2, colour=name)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0) +
  scale_x_continuous(name=NULL) +
  scale_y_continuous(name="volatility (% difference from smoothed trend)",
                     breaks=seq(-20, 20, 2)) +
  ggtitle("Volatility of State Government Tax Revenue and Gross Domestic Product, United States",
          subtitle="Volatility measured as % difference from smoothed trend (HP filter)") +
  theme_bw()

# Colorado-type approach
df2 <- df %>%
  mutate(return=tax / lag(tax),
         dreturn=log(return)*100)

df2 %>%
  filter(year >= 1970) %>%
  ggplot(aes(year, vol)) +
  geom_line() +
  geom_point()

```

## 3-panel graph - explore

```{r}

caiit <- censustax %>%
  filter(stabbr=="CA", name=="iit", year %in% 1975:2005) %>%
  prep_3panel()

p1f(caiit, title="abc", colors,
                xinterval=5, xlims=c(NA, NA), 
                ylims=c(NA, NA), 
                scale=1, scaled_units=NULL)

p3f(caiit, title="abc")

p <- panel3(caiit, "California income tax revenue", scale=1e6, scaled_units="$ billions")
p
ggsave(filename = here::here("results", "caiit_3panel.png"), plot = p, width=8, height=8, scale=1)


censustax %>%
  filter(stabbr=="CA", name=="iit", year %in% 1970:2020) %>%
  prep_3panel() %>%
  panel3("California income tax revenue", scale=1e6, scaled_units="$ billions")

p1 <- censustax %>%
  filter(stabbr=="NY", name=="gst", year %in% 1975:2020) %>%
  prep_3panel() %>%
  panel3("New York sales tax revenue", scale=1e6, scaled_units="$ billions", xlims=c(1975, 2020))

p2 <- stack %>%
  filter(stabbr=="NY", name=="rsgdp", year %in% 1975:2020) %>%
  prep_3panel() %>%
  panel3("New York state real GDP", scale=1e3, scaled_units="$ billions", xlims=c(1975, 2020))

p1 | p2


revdf <- censustax %>%
  filter(stabbr=="NY", name=="gst", year %in% 1975:2020) %>%
  prep_3panel()
econdf <- stack %>%
  filter(stabbr=="NY", name=="rsgdp", year %in% 1975:2020) %>%
  prep_3panel()

revecon4(revdf, econdf,
         revtitle="New York sales tax revenue",
         econtitle="New York state real GDP")


revdf <- censustax %>%
  filter(stabbr=="OH", name=="selsalestax", year %in% 1975:2020) %>%
  prep_3panel()

econdf <- stack %>%
  filter(stabbr=="OH", name=="rsgdp", year %in% 1975:2020) %>%
  prep_3panel()

revecon4(revdf, econdf,
         revtitle="Ohio selsalestax revenue",
         econtitle="Ohio state real GDP")


```

## Introduction and motivation

### Setting the scene - year over year % change

```{r pchplot, fig.height=4.5, fig.width=8, eval=TRUE, include=TRUE}
# sd10c
#  pchplot, fig.height=4.5, fig.width=8, eval=FALSE}
# summary(sgtax.a)
irecs <- which(recessions$rec_year >= 1960)

p2 <- taxgdpus %>%
  filter(year >= 1960) %>%
  ggplot(aes(year, vpch, colour=name)) +
  geom_line(size=1) +
  geom_point(size=1.15) +
  scale_y_continuous(name="% change", breaks=seq(-20, 20, 2)) +
  scale_x_continuous(name=NULL, breaks=seq(1960, 2030, 5)) +
  scale_colour_manual(values=c("red", "blue")) +
  geom_hline(yintercept = 0) +
  annotate("rect",
           xmin = recessions$peak[irecs] %>% decimal_date(),
           xmax = recessions$trough[irecs] %>% decimal_date(),
           fill = "lightgrey",
           alpha = .5, # larger alpha is darker rectangle
           ymin = -Inf, ymax = Inf) +
  ggtitle("Year over year growth in state government tax revenue and GDP for the United States",
          subtitle="Recession periods are shaded.") +
  labs(caption="Source: U.S. Bureau of the Census and Bureau of Economic Analysis") +
  theme_bw() +
  theme(plot.caption = element_text(hjust=0, size=9)) +
  theme(legend.title = element_blank())
p2
ggsave(filename = here::here("results", "tax_gdp_us.png"), plot = p2)


```

## Ways to measure tax revenue volatility

-   Past research
    -   Rolling variance (Dauchy 2013)
    -   Deviation from trend
        -   Kwak 2013 says White 1983; Dye and McGuire 1991; Braun and Otsuka 1998; Aisen and Jose´ Veiga 2008 all used deviation from trend
        -   Orthogonal Distance Regression (Kwak 2013)

# APPENDICES BELOW HERE

## Project management and organization

## Deadlines and schedule

-   July 3, 2020 Project start
-   October 3, 2020 Provide list of external reviewers
-   June 3, 2021 Draft report due (plan on sooner)
-   August 3, 2021 Project end and final report due

## Links

-   [Zotero folder](https://www.zotero.org/groups/2541602/pew_tax_revenue_volatility/library)
-   [Google drive folder](https://drive.google.com/drive/folders/1fM7HIP7qrKbIdndF_wQ30WxVb1j_Cq06)
-   [Google sheet](https://docs.google.com/spreadsheets/d/1hSWy_7ep4RbM9lfSA-LYnDeGI1c6p9Ud0GqcSlR9xJ8/edit?usp=sharing) (same as link below)

## Goals and issues

-   Ways to measure tax revenue volatility

-   Recent trends in revenue volatility

    -   Has tax revenue become more volatile over time?
    -   If so, why?

-   Drivers of state revenue volatility - to what extent driven by

    -   The nature of national economic changes (e.g., characteristics of recession)?
    -   State economic structure?
    -   State tax structure?
        -   The portfolio of taxes?
        -   The structure of individual taxes?

-   Implications for states in current recession and beyond

-   What does the Covid-19 recession imply for state tax volatility?

-   How much can states reduce volatility with alternative tax structures?

Note: For details on references see [this](https://docs.google.com/spreadsheets/d/1hSWy_7ep4RbM9lfSA-LYnDeGI1c6p9Ud0GqcSlR9xJ8/edit?usp=sharing).

## Details about the project

### Overall goals

Examine tax revenue volatility and its impacts on state governments

Provide deep and broad understanding of:

-   Ways to measure tax revenue volatility
-   Recent trends in revenue volatility
-   Drivers of state revenue volatility
-   Implications for states in current recession and beyond

Ultimate goal: Insights that can help policymakers manage budgets

### Important sub-questions

-   To what extent is volatility driven by:
    -   The nature of national economic changes (e.g., characteristics of recession)?
    -   State economic structure?
    -   State tax structure? The portfolio of taxes? The structure of individual taxes?
-   Has tax revenue become more volatile over time? If so, why?
-   How much can states reduce volatility with alternative tax structures?

All have implications for states in current recession and beyond.

### Overall plan

1.  Review relevant literature
2.  Build database for analyzing volatility
    -   Annual state tax revenue, all states, by tax
    -   Auxiliary data to allow population & inflation adjustment
    -   Selected other data
3.  Multiple methods
    -   Descriptive analytic review of history
    -   Decomposition
    -   Simulations of portfolio effects
    -   Exploratory econometrics for certain topics

### Core data source

-   Census Bureau Annual Survey of State Finances
    -   State tax revenue, annual, all states, many taxes
    -   Comprehensive (not just general fund)
    -   Relatively consistent definitions over time and states
    -   Annual period consistent with policymaking cycle
    -   Includes multiple recessions (some data back to mid-1950s):
        -   1980-81, 1990, 2001, and 2007 (2020 won't be available)
        -   may be able to include 1960, 1969, 1973 recessions

### Selected technical & data issues

-   Tax data implicitly include embedded policy changes. Use Pew tax-change data to help answer: How much does adjusting for policy changes affect conclusions?
-   How much do alternative volatility measures affect conclusions?
-   Data on structure of individual taxes are limited. To extent practical, use these limited data to help answer: How does structure of individual taxes affect volatility?

### Topics and approaches

+----------------------------------------------------------------------------------------------------------+-------------------------------------------+
| Topic                                                                                                    | Methods                                   |
+==========================================================================================================+===========================================+
| Ways to measure revenue volatility                                                                       | -   Literature review                     |
|                                                                                                          | -   Conceptual differences among measures |
|                                                                                                          | -   Trends with different measures        |
+----------------------------------------------------------------------------------------------------------+-------------------------------------------+
| Recent revenue volatility trends                                                                         | -   Descriptive analysis                  |
+----------------------------------------------------------------------------------------------------------+-------------------------------------------+
| Drivers of revenue volatility (Nature of economic change, State economic structure, State tax structure) | -   Descriptive analysis                  |
|                                                                                                          | -   Decomposition                         |
|                                                                                                          | -   Econometric analysis                  |
|                                                                                                          | -   Simulation of alternative structures  |
+----------------------------------------------------------------------------------------------------------+-------------------------------------------+
| Implications in current recession and beyond                                                             | -   Synthesis of above                    |
+----------------------------------------------------------------------------------------------------------+-------------------------------------------+

### Comments from the August 4, 2020 methods review

-   Clarify key terms with thorough definitions and describe the applied metrics, highlighting areas of standardization. Absent those details, some confusion may arise. Specifically,
    -   Convey the practical differences between the proposed metrics of revenue volatility for state and other policymakers. These metrics can vary considerably depending on the underlying data distribution, meaningfully changing what is observed as volatility.
    -   Elaborate on how revenue streams may need to be combined/collapsed for analytical tractability. If doing so, highlight the thresholds for those modification, if these differ by state or source, and if those decisions affect the results.\
    -   Account for the possible interpretations of concepts like "state economic structures" and economies "at risk", and how those choices shift the meaning of the analysis and findings. If these concepts are applied differently by state, explain how these may be standardized.
-   Ensure that the selected time frames (periods and trends) and their decision rules are transparent in the final report. Specifically:
    -   Discuss how partial recessionary periods (i.e. quarters) will be addressed since the study will generally leverage annual data, to speak to policymaker budgeting cycles/priorities.
    -   Clarify that "recent trends" may reflect multi-decade periods of fiscal management, which may not be self-evident. Explain how the bounds of "trend periods" are chosen, whether through an assessment of business cycles, regression analyses, inductive analyses or other techniques.
-   Ensure consistency between this and prior Pew studies, or provide a crosswalk, particularly if the methods have been modified or augmented. Focus on what sources of uncertainty remain embedded in each measure of volatility.
    -   Assess whether correcting for tax policy adjustments changes estimated volatility. Also, since the policy adjusted time series is shorter, estimate whether a shorter series will complicate the development of robust volatility estimates using the more technical/complex approaches.
    -   Highlight --- for more technical methods --- how trends are being estimated, and how differences in those trends might change the interpretation of volatility, as the embedded sources of uncertainty differ (e.g. by extracting cyclicality).

```{r}
# data checking
# what to do about negative tax values?? they will become nan in logs, and we need to be robust to that
# what about zero values in the middle that look like they must be missing
# OH cit 2014 is negative (-$118)
# NM cit has pos 1951-1954, zero 1955-1966, then positive

tax_stack %>%
  filter(value <= 0) %>%
  as.data.frame()

tax_stack %>%
  filter(stabbr=="OH", name=="cit", type=="nominal")

tax_stack %>%
  filter(stabbr=="NM", name=="cit", type=="nominal")

measures %>%
  filter(stabbr=="OH", name=="cit", type=="nominal")
```

## Earlier notes

```{r pchplot_old, fig.height=4.5, fig.width=8, eval=FALSE}
# summary(sgtax.a)
irecs <- which(recessions$rec_year >= 1960)

p1 <- sgt1 %>%
  filter(year >= 1960) %>%
  ggplot(aes(year, vpch)) +
  geom_line(colour="blue", size=1) +
  geom_point(colour="blue", size=1.15) +
  scale_y_continuous(name="% change", breaks=seq(-20, 20, 2)) +
  scale_x_continuous(name=NULL, breaks=seq(1960, 2030, 5)) +
  geom_hline(yintercept = 0) +
  annotate("rect",
           xmin = recessions$peak[irecs] %>% decimal_date(),
           xmax = recessions$trough[irecs] %>% decimal_date(),
           fill = "lightgrey",
           alpha = .5, # larger alpha is darker rectangle
           ymin = -Inf, ymax = Inf) +
  ggtitle("Year over year growth in state government tax revenue for the United States",
          subtitle="Recession periods are shaded.") +
  labs(caption="Source: U.S. Bureau of the Census") +
  theme_bw() +
  theme(plot.caption = element_text(hjust=0, size=9))
p1

```

## Experimental area

```{r}

# how correlated are measures
# summarize(correlation = cor(age, circumference))
# pchsd  hpsd       sre sre_stderr
# measures %>%
#   group_by(taxtype) %>%
#   summarise(pch_hp=cor(pchsd, hpsd, use="pairwise.complete.obs"),
#             pch_sre=cor(pchsd, sre, use="pairwise.complete.obs"),
#             pch_srese=cor(pchsd, sre_stderr, use="pairwise.complete.obs"),
#             hp_sre=cor(hpsd, sre, use="pairwise.complete.obs"),
#             hp_srese=cor(hpsd, sre_stderr, use="pairwise.complete.obs"),
#             sre_srese=cor(sre, sre_stderr, use="pairwise.complete.obs"))


# vol2 <- vol %>%
#   mutate(temp = map(sremod, function(mod) unname(mod$result$coefficients["dlgdp"])))
# 
# 
# vol2 <- vol %>%
#   unnest_auto(sremod)
# 
# vol3 <- vol2 %>%
#   tidy(result)
# 
# names(vol2$result[[1]])
#   
#   
#   mutate(mod2 = map(sreunlist(sremod))
# 
# 
# coef %>% 
#                          as.list %>%
#                          as_tibble)
# 
# vol2 %>%vol2 %>%tidy()
#   as_tibble() %>%
#   tidy %>% dplyr::select(term, estimate) %>% spread(term, estimate)))  %>% dplyr::select(id, model1)
# 
# 
# vol2 %>%
#   mutate(coeff=as.double(temp[[1]]))
# 
# d1 <- vol %>%
#   unnest_wider(sremod)
# names(d1$result[[1]])
# 
# d1 %>%
#   hoist(result, "coefficients")
#   unnest_auto(result)
#   
#   hoist(sremod, "coefficients")
# 
# df2 %>%
#   
#   hoist(sre, coeff=c("result", "coeff"))  # reach within result to get coeff
# 
# tmp <- df2 %>%
#   filter(stabbr=="NY", name=="iit")
# str(tmp$sre)
# str(tmp$sre[[1]])
# names(tmp$sre[[1]]$result)
# 
# coef(tmp$sre[[1]])
# 
# tmp$sre[[1]]
# names(tmp$sre[[1]])
# names(tmp$sre[[1]]$result)
# names(tmp$sre[[1]]$result$coeff)
# 
# tmp %>%
#   select(stabbr, name, sre) %>%
#   unnest(cols = c(sre))
# 
# tmp %>%
#   hoist(sre, coeff=list("result", 1))
# 
# tmp %>%
#   hoist(sre, coeff=c("result", "coeff"))
# 
# 
# tmp %>%
#   hoist(sre, "result", "mod", "coefficients", "dlgdp")
# 
# tmp %>%
#   unnest_wider(sre) %>%
#   unnest_wider(result)
# 
# 
# 
# tmp %>%
#   select(stabbr, name, hpf) %>%
#   unnest(cols = c(hpf))
# 
# 
# 
# tmp %>%
#   ungroup %>%
#   hoist(sre,
#         mod="mod")
# 
# 
# tmp %>%
#   ungroup %>%
#   hoist(sre,
#         # mod="mod",
#         coeff="coeff")
# 
# 
# names(tmp2$sre[[1]])
# tmp2 %>%
#   ungroup %>%
#   hoist(sre,
#         # mod="mod",
#         coeff="coeff")
# 
# 
# tmp <- df2 %>%
#   filter(stabbr=="NY", name=="iit", year >= 1965) %>%
#   nest()
# 
# tmp2 <- tmp %>%
#   mutate(sre=map(data, sredf))
# tmp2$sre
# 
# mod <- lm(dlvalue ~ dlgdp, data=tmp$data[[1]])
# str(mod)
# str(mod$coefficients["dlgdp"])
# mod$coefficients["dlgdp"]
# unname(mod$coefficients["dlgdp"])
# str(coef(mod))
# 
# 
# # names(belchers$info[[1]])
# # belchers %>% hoist(info,
# #   name = "name",
# #   age = "age",
# #   dad = "father",
# #   firstborn = list("children", 1L)
# # )
# 
# names(tmp2$sre[[1]])
# tmp2 %>%
#   ungroup %>%
#   hoist(sre,
#         # mod="mod",
#         coeff="coeff")
# 
# tmp2 %>%
#   unnest(sre)
# tmp2 %>%
#   unnest_wider(sre)
# 
# tmp2 %>%
#   hoist(sre, "coeff")
# 
# tmp2 %>%
#   unnest_wider(sre$coeff)
# 
# 
# df3 <- df2 %>%
#   unnest(cols = c(data, mod)) %>%
#   mutate(vol=value / trend * 100 - 100,
#          absvol=abs(vol)) %>%
#   group_by(stabbr, name) %>%
#   mutate(mdn=median(vol),
#          absmdn=median(absvol)) %>%
#   ungroup
# 
# volsum <- df3 %>%
#   group_by(name, stabbr) %>%
#   summarise(hpmdn=median(abs(vol)),
#             hpmean=mean(abs(vol)),
#             hpsd=sd(vol),
#             pchsd=sd(pch, na.rm=TRUE),
#             .groups="drop") %>%
#   arrange(-hpmdn)
# 
# volsum %>%
#   arrange(-hpmdn)
# 
# volsum %>%
#   filter(stabbr=="US")
# 
# volsum %>%
#   filter(name=="tottax")
# 
# # dev.off()
# 
# # US facet plot, one panel for each tax
# usdata <- df3 %>%
#   filter(stabbr=="US") %>%
#   group_by(stabbr, year) %>%
#   mutate(comp=vol[name=="tottax"]) %>%
#   select(stabbr, name, year, vol, comp) %>%
#   pivot_longer(cols=c(vol, comp), names_to = "type")
# 
# usdata %>%
#   ggplot(aes(year, value, colour=type)) +
#   geom_point() +
#   geom_line() +
#   geom_hline(yintercept = 0) +
#   facet_wrap(~name, ncol=3, scales="free")
# 
# stdata <- df3 %>%
#   group_by(name, year) %>%
#   mutate(us=vol[stabbr=="US"]) %>%
#   select(stabbr, name, year, vol, us) %>%
#   pivot_longer(cols=c(vol, us), names_to = "geo")
# 
# stdata %>%
#   filter(name=="tottax") %>%
#   filter(stabbr %in% c("CA", "NY", "AL", "MS", "MA", "CT", "OH")) %>%
#   ggplot(aes(year, value, colour=geo)) +
#   geom_point() +
#   geom_line() +
#   geom_hline(yintercept = 0) +
#   facet_wrap(~stabbr, ncol=3, scales="free")
# 
# 
# sts <- c("AL", "CA", "CT", "FL", "MA", "MS", "NY", "OH","OR", "PA", "TN", "TX", "UT", "VA", "WA")
# df3 %>%
#   filter(name=="tottax") %>%
#   filter(stabbr %in% sts) %>%
#   select(stabbr, name, year, absvol, absmdn) %>%
#   pivot_longer(cols=c(absvol, absmdn), names_to = "measure") %>%
#   ggplot(aes(year, value, colour=measure)) +
#   geom_point() +
#   geom_line() +
#   geom_hline(yintercept = 0) +
#   facet_wrap(~stabbr, ncol=3, scales="free")
# 
# sts <- c("AL", "CA", "CT", "FL", "MA", "MS", "NY", "OH","OR", "PA", "TN", "TX", "UT", "VA", "WA")
# df3 %>%
#   filter(stabbr=="US") %>%
#   select(stabbr, name, year, absvol, absmdn) %>%
#   pivot_longer(cols=c(absvol, absmdn), names_to = "measure") %>%
#   ggplot(aes(year, value, colour=measure)) +
#   geom_point() +
#   geom_line() +
#   geom_hline(yintercept = 0) +
#   facet_wrap(~name, ncol=3, scales="free")

```


# DEADWOOD

```{r}
fig_pdtrend <- function(pdata, gtitle, capt=NULL, seq=.02, acc=.1){
  # ncolors <- length(unique(pdata$name))
  
  years <- min(pdata$year):max(pdata$year)
  recs <- recdata(years)
  
  p <- pdata %>%
    ggplot(aes(year, value, colour=name)) +
    # put recession bands down before lines and dots
    geom_band(recs) +
    geom_line(size=1.5) +
    geom_point(size=1.5) +
    geom_hline(yintercept = 0) +
    scale_x_continuous(name=NULL, breaks=seq(1950, 2030, 5)) +
    scale_y_continuous(breaks=seq(-1, 1, seq),
                       labels=label_percent(accuracy=acc)) +
    scale_colour_brewer(type="qual", palette="Set1") +
    # scale_color_viridis(discrete=TRUE) +
    # scale_color_manual(values=bluered) +
    ggtitle(gtitle,
            subtitle="Percentage deviation from trend") +
    labs(x=NULL, y="% deviation from trend",
         caption=capt) +
    theme_report +
    legend_notitle +
    caption_left
  p
}

```


```{r}
# data prep

vorder <- c("tottax", "iit", "gst", "selsalestax", "cit", "sevtax", "othertax")
(vnames <- taxnames$namef[match(vorder, taxnames$name)])

# we want a df with group=(total, detail), type=(iit, cit, ...othertax), and pdtrend
taxtype_order <- c("tottax", "iit", "gst", "selsalestax", "cit", "sevtax", "othertax")
level_order <- c("detail", "tottax")
# we want a df with level=(tottax, detail), taxtype=(iit, cit, ...othertax), and pdtrend
pdata  <- vbase %>%
  filter(name %in% vorder,
         realnom=="nominal",
         stabbr=="US",
         year >= 1970) %>%
  select(year, name, pdtrend) %>%
  # not sure why eval is required for self-join, but it is
  left_join(x=eval(.) %>% filter(name != "tottax") %>% select(year, taxtype=name, detail=pdtrend),
            y=eval(.) %>% filter(name=="tottax") %>% select(year, tottax=pdtrend),
            by="year") %>%
  pivot_longer(cols = c(detail, tottax), names_to = "level") %>%
  # add names for figure
  left_join(taxnames %>% rename(taxtype=name, taxtypef=namef), by = "taxtype") %>%
  mutate(levelf=ifelse(level=="detail", "Comparison tax", "Total taxes")) %>%
  # ensure that the data sort as desired
  mutate(taxtype=factor(taxtype, levels=taxtype_order),
         taxtypef=fct_reorder2(taxtypef, taxtypef, taxtype, .desc=FALSE),
         level=factor(level, levels=level_order),
         levelf=fct_reorder2(levelf, levelf, level, .desc=FALSE))

# verify ordering
count(pdata, taxtypef, taxtype) 
count(pdata, levelf, level)


```


```{r }

years <- min(pdata$year):max(pdata$year)
recs <- recdata(years)
lsize <- .75
psize <- .75

p <- pdata %>%
  filter(taxtype %in% c("iit", "gst", "selsalestax")) %>%
  ggplot(aes(year, value, colour=levelf)) +
  # put recession bands down before lines and dots
  geom_band(recs) +
  geom_line(size=lsize) +
  geom_point(size=psize) +
  geom_hline(yintercept = 0) +
  scale_x_continuous(name=NULL, breaks=seq(1950, 2030, 5)) +
  scale_y_continuous(breaks=seq(-1, 1, .02),
                     # limits=c(-.35, .35),
                     labels=label_percent(accuracy=1)) +
  # scale_colour_brewer(type="qual", palette="Set1") +
  # scale_color_viridis(discrete=TRUE) +
  scale_color_manual(values=bluegreen) +
  ggtitle("Major taxes",
          subtitle="Percentage deviation from trend") +
  labs(x=NULL, y="% deviation from trend",
       caption=NULL) +
  theme_report +
  legend_notitle +
  caption_left +
  facet_wrap(~taxtypef, ncol=3, scales="fixed")
p

savefig("fig_big3_pdtrend", .p=p, .pdata=pdata, width=10, height=6, scale=1.5) 


```


```{r }
p <- pdata %>%
  filter(taxtype %in% c("cit", "sevtax", "othertax")) %>%
  ggplot(aes(year, value, colour=levelf)) +
  # put recession bands down before lines and dots
  geom_band(recs) +
  geom_line(size=lsize) +
  geom_point(size=psize) +
  geom_hline(yintercept = 0) +
  scale_x_continuous(name=NULL, breaks=seq(1950, 2030, 5)) +
  scale_y_continuous(breaks=seq(-1, 1, .05),
                     labels=label_percent(accuracy=1)) +
  # scale_colour_brewer(type="qual", palette="Set1") +
  # scale_color_viridis(discrete=TRUE) +
  scale_color_manual(values=bluegreen) +
  ggtitle("Smaller taxes",
          subtitle="Percentage deviation from trend") +
  labs(x=NULL, y="% deviation from trend",
       caption=NULL) +
  theme_report +
  legend_notitle +
  caption_left +
  facet_wrap(~taxtypef, ncol=3, scales="fixed")
pbottom


```


```{r }

# vorder <- c("tottax", "iit", "gst", "selsalestax")
vorder <- c("iit", "gst", "selsalestax")
(vnames <- taxnames$namef[match(vorder, taxnames$name)])
pdata  <- vbase %>%
  filter(name %in% vorder,
         realnom=="nominal",
         stabbr=="US",
         year >= 1970) %>%
  select(year, name, value=pdtrend) %>%
  mutate(name=factor(name, levels=vorder, labels=vnames))

p1 <- fig_pdtrend(pdata, gtitle="Volatility of major taxes", acc=1)
p1
savefig("fig_big3_pdtrend", .p=p1, .pdata=pdata, width=10, height=6) 

```


```{r }

# vorder <- c("tottax", "iit", "gst", "selsalestax")
vorder <- c("cit", "sevtax", "othertax")
(vnames <- taxnames$namef[match(vorder, taxnames$name)])
pdata  <- vbase %>%
  filter(name %in% vorder,
         realnom=="nominal",
         stabbr=="US",
         year >= 1970) %>%
  select(year, name, value=pdtrend) %>%
  mutate(name=factor(name, levels=vorder, labels=vnames))

p2 <- fig_pdtrend(pdata, gtitle="Volatility of other taxes", acc=1, seq=.05)
p2
savefig("fig_other_pdtrend", .p=p2, .pdata=pdata, width=10, height=6) 

```



```{r }

vorder <- c("tottax", "iit", "gst", "selsalestax", "cit", "sevtax", "othertax")
(vnames <- taxnames$namef[match(vorder, taxnames$name)])

# we want a df with group=(total, detail), type=(iit, cit, ...othertax), and pdtrend
taxtype_order <- c("tottax", "iit", "gst", "selsalestax", "cit", "sevtax", "othertax")
level_order <- c("detail", "tottax")
# we want a df with level=(tottax, detail), taxtype=(iit, cit, ...othertax), and pdtrend
pdata  <- vbase %>%
  filter(name %in% vorder,
         realnom=="nominal",
         stabbr=="US",
         year >= 1970) %>%
  select(year, name, pdtrend) %>%
  # not sure why eval is required for self-join, but it is
  left_join(x=eval(.) %>% filter(name != "tottax") %>% select(year, taxtype=name, detail=pdtrend),
            y=eval(.) %>% filter(name=="tottax") %>% select(year, tottax=pdtrend),
            by="year") %>%
  pivot_longer(cols = c(detail, tottax), names_to = "level") %>%
  # add names for figure
  left_join(taxnames %>% rename(taxtype=name, taxtypef=namef), by = "taxtype") %>%
  mutate(levelf=ifelse(level=="detail", "Comparison tax", "Total taxes")) %>%
  # ensure that the data sort as desired
  mutate(taxtype=factor(taxtype, levels=taxtype_order),
         taxtypef=fct_reorder2(taxtypef, taxtypef, taxtype, .desc=FALSE),
         level=factor(level, levels=level_order),
         levelf=fct_reorder2(levelf, levelf, level, .desc=FALSE))

# verify ordering
count(pdata, taxtypef, taxtype) 
count(pdata, levelf, level)

years <- min(pdata$year):max(pdata$year)
recs <- recdata(years)
lsize <- .75
psize <- .75

p <- pdata %>%
  ggplot(aes(year, value, colour=levelf)) +
  # put recession bands down before lines and dots
  geom_band(recs) +
  geom_line(size=lsize) +
  geom_point(size=psize) +
  geom_hline(yintercept = 0) +
  scale_x_continuous(name=NULL, breaks=seq(1950, 2030, 5)) +
  scale_y_continuous(breaks=seq(-1, 1, .05),
                     limits=c(-.36, .36),
                     labels=label_percent(accuracy=1)) +
  # scale_colour_brewer(type="qual", palette="Set1") +
  # scale_color_viridis(discrete=TRUE) +
  scale_color_manual(values=bluegreen) +
  ggtitle("Tax revenue volatility: Major taxes in top row, smaller taxes in second row",
          subtitle="Percentage deviation from trend") +
  labs(x=NULL, y="% deviation from trend",
       caption=NULL) +
  theme_report +
  legend_notitle +
  caption_left +
  facet_wrap(~taxtypef, ncol=3, scales="fixed")
p

savefig("fig_all_pdtrend", .p=p, .pdata=pdata, width=10, height=6, scale=1.5) 

```



