---
title: "Data prep for state tax revenue volatility"
author: "Don Boyd"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_notebook:
    df_print: paged
    fig_height: 6
    fig_width: 8
    toc: yes
    toc_depth: 3
    toc_float: true
    number_sections: yes
editor_options:
  chunk_output_type: console
---


<!-- <style> -->
<!-- body .main-container { -->
<!--   max-width: 1500px; -->
<!-- } -->
<!-- </style> -->



```{r runall, eval=FALSE, echo=FALSE} 
# When we want a final report, run the following code selectively "by hand" (interactively)
# -- NEVER using Knit with eval=TRUE
# note <br> breaks line for html output but \n should break for pdf; also could try |

rmdfn <- "./prep_data.rmd" # this file
outfn <- paste0("prepdata_", format(Sys.time(), "%Y-%m-%d"), ".html")
rmarkdown::render(rmdfn, output_format="html_document", output_file=outfn)

# library("RCurl")
# ftpUpload(outfn, "ftp://kffrig:boyd4812@files.000webhost.com/public_html/KFF_RIG_MedicaidCuts_new.html")

# Note may b safest to fully exit RStudio and restart it before running whole thing. Otherwise knitr may get confused
# and include repetitive information in the output html file.

```


```{r setup, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(echo = FALSE, include=FALSE, eval=FALSE, message = FALSE, warning = FALSE)
```


```{r renv_notes, eval=FALSE}
# periodically do:
# renv::snapshot()

# https://rstudio.github.io/renv/articles/renv.html

# Custom R Package Repositories
# Custom and local R package repositories are supported as well. The only requirement is that these repositories
# are set as part of the repos R option, and that these repositories are named. For example, you might use:
#
# repos <- c(CRAN = "https://cloud.r-project.org", WORK = "https://work.example.org")
# options(repos = repos)
# to tell renv to work with both the official CRAN package repository, as well as a package repository you have hosted and set up in your work environment.

# install local packages
# local_package_sources <- r"(C:\RPrograms PC\ProjectsCurrent\Pew_Revenue_Volatility\local\)"
# repos <- c(CRAN = "https://cloud.r-project.org", LOCAL = local_package_sources)
# options(repos = repos)

# note that tarballs are at: C:\RPrograms PC\Packages\bdata_0.6.tar.gz
# create tarballs of the desired packages and save them in the renv\local directory here, so that
# they will be portable. users will then have to install them from the renv\local directory
# steps below

# 1) Define directories
# pkgsource_dir is relevant for Boyd only - it is the location of source files for any packages I created
# pkgsource_dir <- r"(C:\RPrograms PC\Packages\)"  # raw string
# pkgdir is where "tarballs" of local packages will be placed -- this will be...
# pkgdir <- r"(C:\RPrograms PC\ProjectsCurrent\Pew_Revenue_Volatility\renv\local\)"
# pkgdir <- r"(/home/donboyd/Documents/R_projects/projects/Pew_Revenue_Volatility/renv/local/)"
# pkgsource_dir <- r"(C:\RPrograms PC\Packages\)" 

# 2) This step is for Boyd machine only:
# uncomment code below and build and save my local packages tarballs in the renv local directory
devtools::build(pkg = paste0(pkgsource_dir, "btools"),  # bdata BEAData btools
                path = pkgdir,
                binary = FALSE,
                vignettes = FALSE,
                manual = FALSE,
                args = NULL,
                quiet = FALSE)
# do the above for any other needed local packages

# 3) On Boyd machine and on user machine, install these packages from tarballs
# you may have to define the pkgdir location on your machine
# if the package is in use and you are updating it, then first: detach(package:bdata)
# pkgs <- c("bdata_0.7.tar.gz", "BEAData_0.6.0.tar.gz", "btools_0.9.tar.gz")

renv::install(paste0(pkgdir, "bdata_0.7.tar.gz"))
renv::install(paste0(pkgdir, "BEAData_0.6.0.tar.gz"))
renv::install(paste0(pkgdir, "btools_0.9.tar.gz"))
renv::snapshot()

# further notes
# to get it all started:
# renv::init()

# to remove it all:
# renv::deactivate()
# then delete the renv folder

# periodically execute:
# renv::snapshot()

# to see what directory is being used by renv for local package sources, execute:
# renv:::renv_paths_local()

# ${RENV_PATHS_LOCAL}/<package>_<version>.tar.gz

```


```{r get_libs_and_functions, include=FALSE, eval=TRUE}
source(here::here("r", "libs_base.r"))
source(here::here("r", "libs_ts.r"))
source(here::here("r", "functions.r"))
devtools::session_info()
# devtools::package_info()

```


# What this program does

*   General files
    +   State fiscal year for each state (sfy_startmonths)
    +   save all in (general.rdata)
    
*   NIPA national data saved in (econ_national.rdata)
    +   Data frames on state fiscal year basis, 1947-2019
    +   national nominal GDP, state fiscal year basis, 1947-2019 (gdpfy)
    +   national GDP price index, state fiscal year basis, 1947-2019 (gdppi)
    +   Recession dates data frame (recessions)    +   
    
*   NIPA national data for classifying recessions saved in (not yet saved)    +   
    +   CIGXM quarterly 1929- mid 2020 ()
    +   Industrial data ()   
    
*   State economic data saved in (econ_state.rdata)
    +   state gdp, state fiscal year basis, real and nominal (sgdpfy)
    
*   Tax data saved in (taxdata.rdata)
    +   Census by state and tax type, summarized by category,  annual (taxkeep)
    
    
# Test load data

```{r get_data}
# source(here::here("r", "functions.r"))
load(file = here::here("data", "general.RData"), verbose=TRUE)
load(file = here::here("data", "econ_national.RData"), verbose=TRUE)
load(file = here::here("data", "econ_state.RData"), verbose=TRUE)
load(file = here::here("data", "taxdata.RData"), verbose=TRUE)

```


# Data prep

## Data tasks and status
### Done

* State tax revenue from Census by tax, state, and year, 1951-2019, collapsed to the following categories (currently):
  + total taxes (tottax)
  + individual income tax (iit)
  + general sales tax (gst)
  + selective sales taxes such as tobacco, motor fuel, motor vehicle fees, alcohol… (selsalestax)
  + corporate income tax (cit)
  + severance taxes (sevtax)
  + all other taxes (othertax)
* Pew data on policy changes  
* Economic data
  + gross domestic product, by state fiscal year, 1951-2019 (still must adjust for states with fiscal years different from July 1)
  + gross domestic price index (chained), by state fiscal year, 1951-2019 (same adjustment required)
  + official NBER dates for recessions during relevant time periods
  + obtained NIPA details for description of recessions
  + obtained state-level consumption data from BEA

### TODO

* organize NIPA data to measure recessions
* organize NIPA data for consumption details
* organize state consumption data
* Possibly get data from Dauchy
* Graph pch tax vs econ
* IRS data components income
* Measures of income tax progressivity
* Similar measures sales tax?
* Consider base = revenue / rate as robustness check

Certain chunk names in this section start with "ONETIME". In those chucks, I create and save data that will be later read from files. The data only need to be created one time (or updated and re-saved if I decide to update the data).


# General data prep

## Create data frame indicating the start of each state's fiscal year

So that we can adjusting certain data to correspond to state fiscal years.

*   46 states have a July 1 fiscal year
*   NY has April 1
*   TX has Sept 1
*   AL, MI have Oct 1
*   For example, see [this](https://www.ncsl.org/research/fiscal-policy/fy-2021-state-budget-status.aspx)


```{r state_fiscal_years}
# state fiscal year starts  https://www.ncsl.org/research/fiscal-policy/fy-2021-state-budget-status.aspx
# NY April, TX Sept, AL & MI Oct, 46 states July
sfy_startmonths <- tibble(stabbr=state.abb) %>%
  mutate(startmonth=case_when(stabbr=="NY" ~ 4,
                              stabbr=="TX" ~ 9,
                              stabbr %in% c("AL", "MI") ~ 10,
                              TRUE ~ 7))

# saveRDS(sfy_startmonths, here::here("data", "sfy_startmonths.rds"))

```


## Save any general files in a single Rdata file

```{r save_general}
save(sfy_startmonths, file = here::here("data", "general.RData"))

```


## National data

*   I have a package called `BEAData` that includes virtually all major data from the Bureau of Economic Analysis (BEA) in relation to the National Income and Products Accounts (NIPA) and in relation to regional data.
*   For details on data creation see the [github site](https://github.com/donboyd5/BEAData).
*   The package is stored locally in the renv/local folder and can be installed from there.
*   `data(package="BEAData")` shows the files available
*   NIPA national data are in dataframe `nipa`
*   `comment(nipa)` shows creation date

In this section I create data frames with national GDP (nominal) and the GDP price index, on a state fiscal year basis.


```{r nipa_national}
glimpse(nipa)
nipa %>%
  filter(freq=="Q", str_detect(vdesc, "Gross domestic product, Current")) %>%
  group_by(vname, vdesc) %>%
  filter(date==max(date))
# A191RC US nominal gdp
# "A191RG" US GDP price index

gdpfy <- nipa %>%
  filter(vname=="A191RC", freq=="Q") %>%
  # fix NY AL MI TX
  mutate(fyear=ifelse(month(date) >= 7, year + 1, year) %>% as.integer) %>%
  group_by(vname, fyear) %>%
  summarise(gdp=mean(value, na.rm=TRUE), .groups="drop") %>%
  filter(fyear < 2020) %>%
  select(year=fyear, gdp)
gdpfy %>% ht  # 1947, 2019

 
baseyear <- 2019
gdppi <- nipa %>%
  filter(vname=="A191RG", freq=="Q") %>%
  # fix NY AL MI TX
  mutate(fyear=ifelse(month(date) >= 7, year + 1, year) %>% as.integer) %>%
  group_by(vname, fyear) %>%
  summarise(gdppi=mean(value, na.rm=TRUE), .groups="drop") %>%
  filter(fyear < 2020) %>%
  select(year=fyear, gdppi) %>%
  mutate(igdppi=gdppi[year==baseyear] / gdppi) 
summary(gdppi)

```


```{r nber_recessions}
# I have a data frame in the package bdata that has NBER recession dates
# they are pulled from the NBER recessions page, updated in May 2020
# https://www.nber.org/research/data/us-business-cycle-expansions-and-contractions
# see my github page https://github.com/donboyd5/bdata
# particularly https://github.com/donboyd5/bdata/blob/master/DataConversionPrograms/bdata_recessions.R
# data(package="bdata")
# comment(recessions)
recessions # 1857, ..., 2020


```


```{r nipa_national_data_for_recessions}
# https://apps.bea.gov/iTable/iTable.cfm
# we want two kinds of breakdowns:
#    C + I + G + (X - M)
#    Industrial breakdown

#.. CIGXM ----
# Table 1.1.6. Real Gross Domestic Product, Chained Dollars
tab <- NIPAvars %>%
  filter(tabnum == "1.1.6") %>%
  arrange(line)

cigxm_vars <- tab %>%
  filter(line %in% c(1:6, 7, 8, 14, 15, 15, 22))
cigxm_vars

cigxm <- nipa %>%
  right_join(cigxm_vars %>% 
               select(vname), 
             by = "vname")
summary(cigxm)


#.. Industrial breakdown ----
# A150RX
NIPAvars %>%
  filter(vname == "A150RX") 

tmp <- NIPAvars %>%
  filter(str_detect(vdesc, coll("manuf", ignore_case = TRUE))) 
 
mtables <- count(tmp, tabname)
# https://apps.bea.gov/industry/Release/XLS/GDPxInd/ValueAdded.xlsx
# https://apps.bea.gov//industry/iTables%20Static%20Files/AllTablesHist.zip



# Table 5.8.6B. Real Private Inventories and Real Domestic Final Sales by Industry, Chained Dollars
# Table 6.16D. Corporate Profits by Industry
# Table 6.5D. Full-Time Equivalent Employees by Industry

#.. Employment breakdown ----
# https://www.bls.gov/web/empsit/cesseriespub.htm
# https://download.bls.gov/pub/time.series/ce/ce.data.01a.CurrentSeasAE  
# https://download.bls.gov/pub/time.series/ce/ce.txt says:
# ce.data.01a.CurrentSeasAE - contains every seasonally adjusted all employee series and its complete history.    

#.. Causes and characteristics ----
# https://www.investopedia.com/articles/economics/08/past-recessions.asp
# inflation: 1969
# oil and energy crises: 1973, 1980, 1981
# war: 1990 (Gulf)
# 9/11:  2001
# financial crisis: 2007 Great Recession
# real estate: 2007 GR
# 
#
# variables:
# oil prices
# real estate prices
# stock market
# manufacturing, tourism, other sectors
# unemployment rate

```


```{r national_capital_gains}
# historical capital gains for the U.S., annual, calendar (tax) year
# primary sourdces are U.S. Treasury for data before 1995, and CBO for 1995+
# note that ~ 2020+ is CBO forecasts
# see spreadsheet for further details on sources
fn <- "NationalCapitalGains.xlsx"
sheet <- "CapGains"
capgains <- read_excel(here::here("data", fn), sheet=sheet, range="A5:B83", 
                       col_types = c())
capgains <- capgains %>%
  mutate(year=as.integer(year),
         capgains=capgains * 1000)  # put it in millinos of dollars
glimpse(capgains)

```


    
    
```{r save_econ_national}
save(gdpfy, gdppi, recessions, capgains, file = here::here("data", "econ_national.RData"))

```


## BEA state GDP data
```{r gdp_sfy}
data(package="BEAData")
# glimpse(sgdp.q)
summary(sgdp.q)  # starts in 2005-01-01 so this will not be good
summary(sgdp.a) # 1997-2019, only a bit better
summary(sgdp_spliced.a)  # 1963-2019 -- good
count(sgdp_spliced.a, name) # gdp and rgdp

df <- sgdp_spliced.a %>%
  left_join(sfy_startmonths, by = "stabbr") %>%
  group_by(stabbr, name) %>%
  mutate(lagvalue=value[match(year - 1, year)],
         current_share=(startmonth - 1) / 12,
         value_sfy=value * current_share + lagvalue * (1 - current_share))
df

sgdpfy <- df %>%
  ungroup %>%
  filter(!is.na(value_sfy)) %>%
  select(year, stabbr, name, value=value_sfy)
summary(sgdpfy)


```

```{r save_econ_state}
save(sgdpfy, file = here::here("data", "econ_state.RData"))

```

    
## Census Bureau tax data

### Clean tax data

Steps:

1.    Get Census state govt tax data from bdata (previously created)
2.    Keep desired taxes and create an other taxes category
3.    Identify and explore anomalies

```{r get_tax}
# tax1 has untouched data from Census, only the items we want, long format
# tax2 is same, but with the other tax variable


glimpse(sgtax.a)  # 2019 end year; most values from ~1951+
# count(sgtax.a, year) %>% as.data.frame()
# count(sgtax.a, item, desc)

# determine the tax items we want
taxnames <- sgtax.a %>% select(item, desc) %>% distinct()
items <- c("C105", "T09", "C109", "T40", "T41", "T53")
varnames <- c("tottax", "gst", "selsalestax", "iit", "cit", "sevtax")
sumvars <- setdiff(varnames, "tottax")


tax1 <- sgtax.a %>%
  filter(year >= 1951) %>%
  filter(item %in% items) %>%
  mutate(name=factor(item, levels=items, labels=varnames)) %>%
  select(stabbr, name, year, value) %>%
  arrange(stabbr, name, year)
summary(tax1)  # no missing values

# tax2 calculates other tax
tax2 <- tax1 %>%
  # now calculate other tax
  pivot_wider(names_from = name) %>%
  mutate(nonother = rowSums(across(all_of(sumvars)), na.rm=TRUE),
         othertax=tottax - nonother) %>%
  select(-nonother) %>%
  pivot_longer(-c(year, stabbr)) %>%
  filter(!is.na(value))  # we can safely drop NA values because we didn't have any in tax1
summary(tax2)

```

Identify and explore anomalies:

+   missing
+   zero
+   missing or zero at start (always assumed bad)
+   missing or zero at end (always assumed bad)
+   missing in middle
+   zero in middle
+   negative in middle (logs won't work -- only 1 found)
+   short runs (fewer than X positive values in a row)


```{r anomalies}
# identify different anomalies
# CO sevtax  2017 0
# OH cit 2014 -118 
tax2 %>% filter(stabbr=="AK", name=="iit")
tax2 %>% filter(stabbr=="AK", name=="gst")
tax2 %>% filter(stabbr=="AZ", name=="sevtax")
tax2 %>% filter(stabbr=="CO", name=="sevtax")
tax2 %>% filter(stabbr=="OH", name=="cit")
tax2 %>% filter(stabbr=="NH", name=="sevtax")
tax2 %>% filter(stabbr=="SD", name=="sevtax")

# vectors used to test the functions below
a1 <- c(0, 0, NA, 1, 2, 3, 0, 4, 5, 6, 0, 0)
a2 <- c(0, 0, 0, 0)
a3 <- c(0, NA, 0, 0)
a4 <- c(7, 9, 0, 0, NA, 1, 2, 3, 0, 4, 5, 6)

firstnz <- function(x) {
  # find index of first nonzero value, to help us identify zeroes at the start
  ifelse(all(x==0, na.rm=TRUE), 
         Inf, 
         which(x != 0)[[1]])
}

# test firstnz
firstnz(a1)
firstnz(a2)
firstnz(a3)
firstnz(a4)

lastnz <- function(x) {
  # find index of last nonzero value, to help us identify zeroes at the end
  ifelse(all(x==0, na.rm=TRUE),
         -Inf, 
         tail(which(x != 0), 1))
}

# test lastnz
lastnz(a1)
lastnz(a2)
lastnz(a3)
lastnz(a4)


firstgap <- function(gap) {
  # gap is a a vector of 0 or 1 indicators for a series
  # 0 indicates the observation is not part of a gap and 1 indicates it is part of a gap
  # firstgap finds the index at which a gap in a series starts (if a gap exists)
  ifelse(all(gap !=TRUE, na.rm=TRUE),
         Inf, # return Inf if there is no gap in the series
         which(gap==TRUE)[[1]]) # index of the start of the first gap
}


lastgap <- function(gap) {
  # index of the final observation in last gap
  ifelse(all(gap !=TRUE, na.rm=TRUE),
         -Inf, # return -Inf if there is no gap in the series
         tail(which(gap==TRUE), 1)) # index of the final observation in the last gap of a series
}


# identify anomalies within each state-tax combination
# also, for series that have gaps in them, identify observations that occur before the first gap or
# after the last gap, as we will only want to keep one or the other
anomalies <- tax2 %>%
  group_by(stabbr, name) %>%
  mutate(missval = is.na(value),  # we should not have any missing values but check anyway
         zeroval = value == 0,
         negval = value < 0,
         leadzero = row_number() < firstnz(value),   # identify all leading-zero rows
         trailzero = row_number() > lastnz(value),  # identify all trailing-zero rows
         gap = zeroval & !(leadzero | trailzero),  # a gap occurs if we have a zero that is not leading or trailing
         pregap = row_number() < firstgap(gap),   # identify data that occurs before the first gap
         postgap = row_number() > lastgap(gap))   # identify data that occurs after the last gap
summary(anomalies)
# what should we do if a series has multiple gaps?? look to make sure it doesn't happen; I don't think so

anomalies %>% filter(stabbr=="AK", name=="iit")
anomalies %>% filter(stabbr=="AK", name=="gst")
anomalies %>% filter(stabbr=="AZ", name=="sevtax")
anomalies %>% filter(stabbr=="AZ", name=="cit") # gap in 1957
anomalies %>% filter(stabbr=="CO", name=="sevtax")
anomalies %>% filter(stabbr=="OH", name=="cit")
anomalies %>% filter(stabbr=="NH", name=="sevtax")
anomalies %>% filter(stabbr=="NV", name=="sevtax")
anomalies %>% filter(stabbr=="SD", name=="sevtax")
anomalies %>% filter(gap)


anomalies %>% 
  filter(stabbr=="NV", name=="sevtax") %>%
  lastgap()
# note that NV sevtax jumped way up after 1987 -- investigate

# rules:
# - drop all lead zero and trail zero
# - take the post gap periods
# individually and decide - take the post gap

# get rid of leading and trailing zeros
anomalies_v2 <- anomalies %>%
  filter(!(leadzero | trailzero))

```


```{r anomalies_gaps}
# explore the gaps -- the only bad items -- and decide what to keep
summary(anomalies_v2)

# what gaps do we have?
gaps <- anomalies_v2 %>%
  group_by(stabbr, name) %>%
  filter(any(gap != 0)) %>%
  ungroup

count(gaps, stabbr, name) # only 7 combinations

gaps %>%
  group_by(stabbr, name) %>%
  summarise(gap_start=min(year[gap==TRUE]))

anomalies_v2 %>% filter(stabbr=="AZ", name=="cit")
anomalies_v2 %>% filter(stabbr=="CO", name=="sevtax")  # 2017 is zero; some big changes in early years
anomalies_v2 %>% filter(stabbr=="MO", name=="sevtax")  # a string of zeroes before 1978
anomalies_v2 %>% filter(stabbr=="NH", name=="sevtax") # we'll want to drop this for other reasons
anomalies_v2 %>% filter(stabbr=="NM", name=="cit")  # 1967+ is good
anomalies_v2 %>% filter(stabbr=="NV", name=="sevtax") # 1988+ is good - low values or zero prior
anomalies_v2 %>% filter(stabbr=="SD", name=="sevtax") # 1976+ is good

# there are only 7 so identify them by hand
gapkeep <- tribble(~stabbr, ~name, ~year1, ~year2,
                   "AZ", "cit", 1958, 2019,
                   "CO", "sevtax", 1951, 2016,
                   "MO", "sevtax", 1978, 2019,
                   "NH", "sevtax", 1951, 1987,
                   "NM", "cit", 1967, 2019,
                   "NV", "sevtax", 1988, 2019,
                   "SD", "sevtax", 1976, 2019,
                   ) 
gapkeep

anomalies_v3 <- anomalies_v2 %>%
  left_join(gapkeep, by=c("stabbr", "name")) %>%
  mutate(drop=ifelse(!is.na(year1) & {(year < year1) | (year > year2)},
                     TRUE, FALSE)) %>%
  filter(!drop)
anomalies_v3 %>% filter(stabbr=="SD", name=="sevtax") 

taxkeep <- anomalies_v3 %>%
  select(stabbr, name, year, value) %>%
  ungroup
summary(taxkeep)
glimpse(taxkeep)
# saveRDS(taxkeep, here::here("data", "taxkeep.rds"))

```


```{r save_taxdata}
save(taxkeep, file=here::here("data", "taxdata.RData"))

```

## Pew data on enacted revenue changes

*   Pew constructed estimates of the percent change in major taxes, by year, state, and major tax type for work it did in ___
*   In a 10/8/2015 email, on my request Brenna Erford <BErford@pewtrusts.org> provided the file "Boyd Data Request 100815.xlsx" to me (donald.boyd@rockinst.suny.edu), which had data underlying the project
*   Pew's short description of the data, in the "Cover" tab of the workbook, was:
    
<blockquote style="color:grey;font-size:100%;margin-left: 4em">
  <p>Volatility scores for each state’s total tax revenue and specific tax sources were calculated using the U.S. Census Bureau’s State Government Tax Collections historical data series for 1994 to 2014, accessed April 21, 2015. Data were adjusted to control for the effects of tax policy change using the National Conference of State Legislatures (NCSL) State Tax Action reports for 1994 to 2013, accessed in January 2015.</p>
</blockquote>


```{r ONETIME_pew_revenue_data, eval=FALSE}
# data file Boyd Data Request 100815.xlsx
xlfn <- here::here("data", "Boyd Data Request 100815.xlsx")
sheets <- excel_sheets(xlfn)
sheets

get_pewsheet <- function(sheet){
  print(sheet)
  range <- ifelse(str_detect(sheet, "Major"), "A5:V190", "A5:V56")
  read_excel(xlfn, sheet=sheet, range=range) %>%
    mutate(sheet=sheet)
}

pewdf <- map_dfr(sheets[-1], get_pewsheet)
glimpse(pewdf)
count(pewdf, sheet)
count(pewdf, `Major Revenue Source`)

#    `Major Revenue Source`               n
#    <chr>                            <int>
#  1 Amusements                           9
#  2 Corporation                          9
#  3 Corporation net income              90
#  4 Documentary and stock transfer       6
#  5 General sales and gross receipts   135
#  6 Individual income                  123
#  7 Motor fuels                         99
#  8 Motor vehicle                       12
#  9 Occupation and business, NEC         6
# 10 Property taxes                      27
# 11 Public utilities                     9
# 12 Severance                           27
# 13 Tobacco products                     3
# 14 Total                              153

type_map <- read_delim("revsource; taxtype
Amusements; amusetax
Corporation; corptax
Corporation net income; cit
Documentary and stock transfer; doctax
General sales and gross receipts; gst
Individual income; iit
Motor fuels; mft
Motor vehicle; mvl
Occupation and business, NEC; occtax
Property taxes; proptax
Public utilities; utiltax
Severance; sevtax
Tobacco products; cigtax
Total; tottax", 
delim=";", trim_ws = TRUE)

pewlong <- pewdf %>%
  select(stname=1, revsource=2, everything()) %>%
  left_join(bdata::stcodes %>% select(stabbr, stname), by="stname") %>%
  left_join(type_map, by="revsource") %>%
  mutate(vartype=word(sheet, 3)) %>%
  pivot_longer(-c(stabbr, stname, revsource, sheet, vartype, taxtype), names_to = "year") %>%
  mutate(year=as.integer(year)) %>%
  select(stabbr, stname, sheet, vartype, revsource, taxtype, year, value) %>%
  arrange(stabbr, vartype, taxtype, year)
  
glimpse(pewlong)

# caution: US data do not have major taxes, just the total
count(pewlong, stabbr, stname)

pewlong %>% filter(stabbr=="US")
saveRDS(pewlong, here::here("data", "pewlong.rds"))

```


```{r vol_measures}
# hp_df <- function(df, vname){
#   vts <- ts(df[vname])
#   vts_hp <- hpfilter(vts, freq=NULL, type=c("lambda"), drift=FALSE)
#   df %>% mutate(trend=as.numeric(vts_hp$trend),
#                 cycle=as.numeric(vts_hp$cycle))
# }


# takes a LONG LONG time
# vol1 <- tax_real %>%
#   filter(name=="tottax") %>%
#   group_by(stabbr, name, type) %>%
#   arrange(year) %>%
#   hp_df(., "value")
#   



# df <- df %>%
#   f(., "iit") %>%
#   mutate(dtrend=cycle / trend * 100,
#          pch=iit / iit[match(year - 1, year)] * 100 - 100,
#          dtrend_sd10=rollsd_p(dtrend, 10),
#          pch_sd10=rollsd_p(pch, 10))


```



```{r explore}
# remove trend
df <- sgtax.a %>%
  filter(year >= 1951, stabbr=="NY", item=="T40") %>%
  select(year, stabbr, iit=value)
df %>% ht  

# iit <- ts(df$iit, start = 1951, end=2019)
# iit <- ts(df$iit)
# plot(iit)
# 
# stl(iit, s.window = "periodic")
# # hpfilter(iit, freq=NULL, type=c("lambda","frequency"),drift=FALSE)
# iit_hp <- hpfilter(iit, freq=NULL, type=c("lambda"), drift=FALSE)
# iit_hp$trend
# iit_hp$cycle

f <- function(df, vname){
  vts <- ts(df[vname])
  vts_hp <- hpfilter(vts, freq=NULL, type=c("lambda"), drift=FALSE)
  df %>% mutate(trend=as.numeric(vts_hp$trend),
                cycle=as.numeric(vts_hp$cycle))
}

df <- df %>%
  f(., "iit") %>%
  mutate(dtrend=cycle / trend * 100,
         pch=iit / iit[match(year - 1, year)] * 100 - 100,
         dtrend_sd10=rollsd_p(dtrend, 10),
         pch_sd10=rollsd_p(pch, 10))

glimpse(df)


df %>%
  select(year, iit, trend) %>%
  pivot_longer(cols=c(iit, trend)) %>%
  ggplot(aes(year, value, colour=name)) +
  geom_line() +
  geom_point()


df %>%
  ggplot(aes(year, dtrend)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0)
  
df %>%
  ggplot(aes(year, pch)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0)

# df %>%
#   select(year, dtrend, pch) %>%
#   pivot_longer(cols=-year) %>%
#   ggplot(aes(year, value, colour=name)) +
#   geom_line() +
#   geom_point() +
#   geom_hline(yintercept = 0)
# 
# df %>%
#   select(year, dtrend, pch) %>%
#   mutate(pch=pch - mean(pch, na.rm=TRUE)) %>%
#   pivot_longer(cols=-year) %>%
#   ggplot(aes(year, value, colour=name)) +
#   geom_line() +
#   geom_point() +
#   geom_hline(yintercept = 0)

# another change

df %>%
  select(year, dtrend_sd10, pch_sd10) %>%
  pivot_longer(cols=-year) %>%
  ggplot(aes(year, value, colour=name)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0)
 
```


## Calculate volatility measures

```{r vfunctions}
hpdf <- function(data) {
  mod <- hpfilter(data$value, freq=6.25, type=c("lambda"), drift=FALSE)
  tibble(trend=as.numeric(mod$trend),
         cycle=as.numeric(mod$cycle))
}


dl <- function(vec) {
  # change in log
  # ASSUME data are soreted by year and there are no gaps and no zeros
  lvec = log(vec)
  laglvec = c(NA, head(lvec, -1))
  lvec - laglvec
}



```


```{r vcalc}
df <- taxkeep %>%
  arrange(name, year) %>%
  nest_by(name) %>%
  filter(year >= 1960, stabbr=="US")

# %>% mutate(new=hp_df(., "value"))

taxkeep %>%
  filter(year >= 1960, stabbr=="US") %>%
  arrange(name, year) %>%
  group_by(name) %>%
  mutate(hp=hpfilter(value, freq=1200, type=c("lambda"), drift=FALSE))

vec <- df %>% filter(name=="tottax") %>% arrange(year) %>% .$value
res <- hpfilter(vec, freq=1200, type=c("lambda"), drift=FALSE)
str(res)
typeof(res)
res$trend[, 1]




df2 <- taxkeep %>%
  left_join(sgdpfy %>%
              filter(name=="gdp") %>%
              rename(gdp=value) %>%
              select(year, stabbr, gdp),
            by=c("stabbr", "year")) %>%
  arrange(stabbr, name, year) %>% 
  group_by(stabbr, name) %>%
  mutate(pch=value / value[match(year - 1, year)] * 100 - 100,
         pch_rollsd = rollsd(pch, 10),
         dlvalue=dl(value),
         dlgdp=dl(gdp)) %>%
  filter(year >= 1964) %>%
  ungroup %>%
  group_nest(stabbr, name) %>%
  mutate(hpf=map(data, hpdf),
         sre=map(data, srddf))

# Problem with `mutate()` column `dlvalue`.
# ℹ `dlvalue = dl(value)`.
# ℹ NaNs produced
# ℹ The warning occurred in group 224: stabbr = "OH", name = "cit". 


sredf <- function(data) {
  # short-run elasticity
  mod = lm(dlvalue ~ dlgdp, data=data)
  list(mod=mod, coeff=unname(mod$coefficients["dlgdp"]))
}

tmp <- df2 %>%
  filter(stabbr=="NY", name=="iit", year >= 1965) %>%
  nest()

tmp2 <- tmp %>%
  mutate(sre=map(data, sredf))
tmp2$sre

mod <- lm(dlvalue ~ dlgdp, data=tmp$data[[1]])
str(mod)
str(mod$coefficients["dlgdp"])
mod$coefficients["dlgdp"]
unname(mod$coefficients["dlgdp"])
str(coef(mod))


names(belchers$info[[1]])
belchers %>% hoist(info,
  name = "name",
  age = "age",
  dad = "father",
  firstborn = list("children", 1L)
)

names(tmp2$sre[[1]])
tmp2 %>%
  ungroup %>%
  hoist(sre,
        # mod="mod",
        coeff="coeff")

tmp2 %>%
  unnest(sre)
tmp2 %>%
  unnest_wider(sre)

tmp2 %>%
  hoist(sre, "coeff")

tmp2 %>%
  unnest_wider(sre$coeff)


df3 <- df2 %>%
  unnest(cols = c(data, mod)) %>%
  mutate(vol=value / trend * 100 - 100,
         absvol=abs(vol)) %>%
  group_by(stabbr, name) %>%
  mutate(mdn=median(vol),
         absmdn=median(absvol)) %>%
  ungroup

volsum <- df3 %>%
  group_by(name, stabbr) %>%
  summarise(hpmdn=median(abs(vol)),
            hpmean=mean(abs(vol)),
            hpsd=sd(vol),
            pchsd=sd(pch, na.rm=TRUE),
            .groups="drop") %>%
  arrange(-hpmdn)

volsum %>%
  arrange(-hpmdn)

volsum %>%
  filter(stabbr=="US")

volsum %>%
  filter(name=="tottax")

# dev.off()

# US facet plot, one panel for each tax
usdata <- df3 %>%
  filter(stabbr=="US") %>%
  group_by(stabbr, year) %>%
  mutate(comp=vol[name=="tottax"]) %>%
  select(stabbr, name, year, vol, comp) %>%
  pivot_longer(cols=c(vol, comp), names_to = "type")

usdata %>%
  ggplot(aes(year, value, colour=type)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 0) +
  facet_wrap(~name, ncol=3, scales="free")

stdata <- df3 %>%
  group_by(name, year) %>%
  mutate(us=vol[stabbr=="US"]) %>%
  select(stabbr, name, year, vol, us) %>%
  pivot_longer(cols=c(vol, us), names_to = "geo")

stdata %>%
  filter(name=="tottax") %>%
  filter(stabbr %in% c("CA", "NY", "AL", "MS", "MA", "CT", "OH")) %>%
  ggplot(aes(year, value, colour=geo)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 0) +
  facet_wrap(~stabbr, ncol=3, scales="free")


sts <- c("AL", "CA", "CT", "FL", "MA", "MS", "NY", "OH","OR", "PA", "TN", "TX", "UT", "VA", "WA")
df3 %>%
  filter(name=="tottax") %>%
  filter(stabbr %in% sts) %>%
  select(stabbr, name, year, absvol, absmdn) %>%
  pivot_longer(cols=c(absvol, absmdn), names_to = "measure") %>%
  ggplot(aes(year, value, colour=measure)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 0) +
  facet_wrap(~stabbr, ncol=3, scales="free")

sts <- c("AL", "CA", "CT", "FL", "MA", "MS", "NY", "OH","OR", "PA", "TN", "TX", "UT", "VA", "WA")
df3 %>%
  filter(stabbr=="US") %>%
  select(stabbr, name, year, absvol, absmdn) %>%
  pivot_longer(cols=c(absvol, absmdn), names_to = "measure") %>%
  ggplot(aes(year, value, colour=measure)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 0) +
  facet_wrap(~name, ncol=3, scales="free")



# df3 %>%
#   filter(name %in% c("iit", "gst", "tottax")) %>%
#   ggplot(aes(year, vol, colour=name)) +
#   geom_point() +
#   geom_line() +
#   geom_hline(yintercept = 0)
# 
# df3 %>%
#   filter(name %in% c("cit", "tottax")) %>%
#   ggplot(aes(year, vol, colour=name)) +
#   geom_point() +
#   geom_line() +
#   geom_hline(yintercept = 0)
# 
# 
# df3 %>%
#   filter(name %in% c("selsalestax", "tottax")) %>%
#   ggplot(aes(year, vol, colour=name)) +
#   geom_point() +
#   geom_line() +
#   geom_hline(yintercept = 0)
#   
# 
# df3 %>%
#   filter(name %in% c("cit", "tottax")) %>%
#   ggplot(aes(year, vol, colour=name)) +
#   geom_point() +
#   geom_line() +
#   geom_hline(yintercept = 0)
  
  


```


