---
title: "State tax revenue volatility"
subtitle: "Methods review discussion"
author: "Don Boyd"
date: "August 4, 2020"
output:
  beamer_presentation:
    slide_level: 2
    includes:
      in_header: header_pagenos.tex
classoption: "aspectratio=169"      
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## States have highly volatile tax revenue

<!-- I put this section after first slide header or else beamer adds a blank slide

"`r format(Sys.time(), '%B %d, %Y')`"

ioslides
output:
  ioslides_presentation:
    widescreen: true
    smaller: true

beamer
output:
  beamer_presentation:
    slide_level: 2
classoption: "aspectratio=169"         
    
Themes: https://hartwork.org/beamer-theme-matrix/    
-->

<!-- For more details on authoring R presentations please visit <https://support.rstudio.com/hc/en-us/articles/200486468>. -->


```{r include=FALSE}
source(here::here("r", "libs_base.r"))
# source(here::here("r", "libs_ts.r"))
source(here::here("r", "functions.r"))
devtools::session_info()

glimpse(sgtax.a)
# df <- sgtax.a
# count(df, year) %>% as.data.frame()
# count(df, vname)

recessions

sgt1 <- sgtax.a %>%
  mutate(year=as.integer(year)) %>%
  filter(stabbr=="US", item=="C105", year>=1950) %>%
  mutate(vpch=pchya(value, year),
         sd10=rollsd_p(vpch, 10),
         sd10c=rollsd_p2(vpch, 10)) # population sd

```


```{r pchplot, fig.height=4.5, fig.width=8}
# summary(sgtax.a)
irecs <- which(recessions$rec_year >= 1960)

p1 <- sgt1 %>%
  filter(year >= 1960) %>%
  ggplot(aes(year, vpch)) +
  geom_line(colour="blue", size=1) +
  geom_point(colour="blue", size=1.15) +
  scale_y_continuous(name="% change", breaks=seq(-20, 20, 2)) +
  scale_x_continuous(name=NULL, breaks=seq(1960, 2030, 5)) +
  geom_hline(yintercept = 0) +
  annotate("rect",
           xmin = recessions$peak[irecs] %>% decimal_date(),
           xmax = recessions$trough[irecs] %>% decimal_date(),
           fill = "lightgrey",
           alpha = .5, # larger alpha is darker rectangle
           ymin = -Inf, ymax = Inf) +
  ggtitle("Year over year growth in state government tax revenue for the United States",
          subtitle="Recession periods are shaded.") +
  labs(caption="Source: U.S. Bureau of the Census") +
  theme_bw() +
  theme(plot.caption = element_text(hjust=0, size=9))
p1

```



```{r vplot, eval=FALSE}
# summary(sgtax.a)

irecs <- which(recessions$rec_year >= 1960)

p <- sgt1 %>%
  filter(year >= 1960) %>%
  ggplot(aes(year, sd10)) +
  geom_line(colour="blue") +
  geom_point(colour="blue", size=0.75) +
  scale_y_continuous(name="Standard deviation (%)", breaks=seq(0, 10, .5), limits=c(0, NA)) +
  scale_x_continuous(name=NULL, breaks=seq(1960, 2030, 5)) +
  annotate("rect",
           xmin = recessions$peak[irecs] %>% decimal_date(),
           xmax = recessions$trough[irecs] %>% decimal_date(),
           fill = "lightgrey",
           alpha = .5, # larger alpha is darker rectangle
           ymin = -Inf, ymax = Inf) +
  ggtitle("Volatility in aggregate state government tax revenue for the United States",
          subtitle="Trailing 10-year standard deviation of annual % change. Recession periods are shaded.") +
  theme_bw()
p    


```



## Tax revenue volatility and its impacts on state governments

Provide deep and broad understanding of:

* Ways to measure revenue volatility
* Recent trends in revenue volatility
* Drivers of state revenue volatility
*	Implications for states in current recession and beyond

Ultimate goal: Insights that can help policymakers manage budgets


## Important sub-questions

* To what extent is volatility driven by:
    + The nature of national economic changes (e.g., characteristics of recession)?
    + State economic structure?
    + State tax structure? The portfolio of taxes? The structure of individual taxes?
* Has tax revenue become more volatile over time? If so, why?
* How much can states reduce volatility with alternative tax structures?

All have implications for states in current recession and beyond.


## Overall plan

1. Review relevant literature
2. Build database for analyzing volatility
    + Annual state tax revenue, all states, by tax
    + Auxiliary data to allow population & inflation adjustment
    + Selected other data
3. Multiple methods
    + Descriptive analytic review of history
    + Decomposition
    + Simulations of portfolio effects
    + Exploratory econometrics for certain topics


## Core data source

* Census Bureau Annual Survey of State Finances
  + State tax revenue, annual, all states, many taxes
  + Comprehensive (not just general fund)
  + Relatively consistent definitions over time and states
  + Annual period consistent with policymaking cycle
  + Includes multiple recessions (some data back to mid-1950s):
      + 1980-81, 1990, 2001, and 2007 (2020 won't be available)
      + may be able to include 1960, 1969, 1973 recessions

## Selected technical & data issues

* Tax data implicitly include embedded policy changes. Use Pew tax-change data to help answer: How much does adjusting for policy changes affect conclusions?
* How much do alternative volatility measures affect conclusions?
* Data on structure of individual taxes are limited. To extent practical, use these limited data to help answer: How does structure of individual taxes affect volatility?
  

## Topics and approaches
+----------------------------------+----------------------------------------------+
| Topic                            | Methods                                      |
+==================================+==============================================+
| Ways to measure                  | - Literature review                          |
| revenue volatility               | - Conceptual differences among measures      |
|                                  | - Trends with different measures             |
+----------------------------------+----------------------------------------------+
| Recent revenue volatility trends | - Descriptive analysis                       |
+----------------------------------+----------------------------------------------+
| Drivers of revenue volatility    | - Descriptive analysis                       |
|   (Nature of economic change,    | - Decomposition                              |
|    State economic structure,     | - Econometric analysis                       |
|    State tax structure)          | - Simulation of alternative structures       |
+----------------------------------+----------------------------------------------+
| Implications in current          | - Synthesis of above                         |
| recession and beyond             |                                              |
+----------------------------------+----------------------------------------------+


## Concluding comments

Ultimate goal: Provide insights that will help policymakers understand tax revenue volatility and manage its consequences for budgets.  


## Appendix


## Research review questions (1)

* How much data?
  + ~60 years x 50 states x ~25 tax types (collapsed down to ~5)
  + ~15k year-state-tax observations after collapsing
  
* Why annual?
  + It is the major policy making cycle for states
  + Intra-year volatility does matter, but not as much
  + Good data are available
  
## Research review questions (2)

* Why not general fund?
  + Often much smaller than total tax revenue
  + Only captures some of the relevant volatility
  + Not consistent across states or over time
  + Not good data
  
* Why not fees?
  + Generally not as volatile as tax revenue
  + Much smaller than tax revenue (less budgetary importance)
  + Hard to measure, not good data
  
## Auxiliary data sources I definitely will use

* Census Bureau annual state population data for per-capita adjustment
* Bureau of Economic Analysis (BEA): GDP price index
  +  National, annual
  +  For price adjustment
* Pew state tax policy adjustments
  +  Only available for a subset of relevant time periods and taxes
  +  Noisy
  +  Main purpose is to learn/identify how much our view of volatility changes if we adjust for policy changes.
* BEA state gross domestic product, industrial detail, annual
  +  For analyzing importance of differences in industrial composition
  +  Especially during recessions
  +  Not available for all years (SIC --> NAICS)
  
## Potential additional data sources I may use

* BEA state regional price parities, annual
  +  Might be useful for alternative view of price adjustment
* IRS Statistics of Income, state, annual
  +  Breakdown of components of income
  +  Certain components such as capital gains can play important role in states with income taxes
  +  Only available for recent years (~ 2000's+)
* State sales tax base characteristics, if practical
  +  Potentially valuable for disentangling role of economic structure
  +  Potential sources include FTA or Mikesell

## Ways to measure revenue volatility (1)

* Strengths and weaknesses of potential measures, such as
    + Standard deviation of percent change
    + Interquartile range of percent change
    + Deviation from trend
    + Others based on literature review

## Ways to measure revenue volatility (2)

* Selected measurement issues
    + Adjusting for inflation
    + Per-capita adjustment
    + Adjusting for policy changes
      + Will use the Pew policy-change data (only available for 1995-2014)
      + What questions are best addressed with these adjustments?
      + How much does it matter?
      + (And how good are these data for this purpose?)

## Recent trends in revenue volatility

* Variation over different time periods
* Variation across states
* Variation across tax revenue sources

